\documentclass{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{hyperref}
\begin{document}
\section*{Plan: Gömb tuner --- parameter count visibility and SNAP joint AUC search}

\textbf{Date:} 2026-05-12 \quad \textbf{Slug:} gomb-tune-n-params-auc

\subsection*{Scope and goal}
\begin{itemize}
  \item Surface \texttt{n\_params} from each \texttt{run\_gomb\_smoke} JSON row in
    \texttt{run\_gomb\_tune} trial logs and per-dataset phase summaries.
  \item Mildly expand compact search on Slashdot/Epinions (extra \texttt{d\_embed}
    choice; joint compact \texttt{topk} cap $24 \rightarrow 28$) to seek higher
    test AUROC under existing joint-slot subsample contract.
  \item Bump default Slashdot trial/epoch counts in
    \texttt{run\_gomb\_external\_auc\_tuning.sh} for deeper random search.
\end{itemize}

\subsection*{Affected files}
\texttt{signedkan\_wip/src/run\_gomb\_tune.py},
\texttt{signedkan\_wip/experiments/run\_gomb\_external\_auc\_tuning.sh},
\texttt{signedkan\_wip/tests/test\_hymeko\_gomb.py},
\texttt{docs/plans/2026-05-12-gomb-tune-n-params-auc/*},
\texttt{reports/2026-05-12-gomb-tune-n-params-auc.md}.

\subsection*{CORE.YAML}
None touched.

\subsection*{Interface changes}
None to public Rust; Python-only: internal helper
\texttt{\_n\_params\_from\_row}; behaviour change: SNAP joint compact
\texttt{topk} clamp constant; optional new summary keys \texttt{best\_n\_params}.

\subsection*{Test strategy}
Unit tests: \texttt{\_n\_params\_from\_row} parsing; update joint compact
clamp assertion ($\le 28$); regression: existing tuner/smoke tests pass.

\subsection*{Performance budget}
Subprocess RSS unchanged in intent; optional \texttt{topk} up to 28 may increase
VRAM vs 24 --- user requested AUC tuning; cap remains below full-menu 40.

\subsection*{Rollback path}
Revert clamp constant and \texttt{d\_embed} menu; remove log/summary fields.

\subsection*{Risks}
Joint compact \texttt{topk}=28 may OOM on smallest GPUs; mitigated by
\texttt{joint\_slot\_cap} and walk caps unchanged.

\end{document}
