\documentclass[11pt]{article}
\usepackage[a4paper,margin=2cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}

\title{Plan: 3-Orthogonal-Dimension Architectural Factorial\\
  HSiKAN $\times$ CPML $\times$ Clifford-FIR}
\author{}
\date{2026-05-11}

\begin{document}
\maketitle

\section{Scope}

We have spent the day disentangling architectural contributions by
ablating one at a time. The pattern caught two false attributions
already: ``CPML breaks Epinions'' was an embedding-lookup artefact,
``+ricci-mod is best HyMeYOLO config'' was a wrong-metric
artefact. The pattern that keeps catching them is the
\emph{orthogonal-dimension factorial}: each architectural feature
is toggled independently, every combination cell is measured, and
attribution becomes a question of effect decomposition over the
factorial design.

This plan formalises the **three architectural axes** that have
emerged as the candidate ICLR / MLSys contributions of this
codebase. Each is independently toggleable, and any subset can be
active in a given model:

\begin{enumerate}
  \item \textbf{HSiKAN} (Catmull-Rom signed-spline aggregator)
  \item \textbf{CPML} (tier-stratified multi-layer topology)
  \item \textbf{Clifford-FIR} (closed-form Clifford-derivative
    linear aggregator)
\end{enumerate}

The factorial is $2^3 = 8$ cells, plus a baseline embedding-only
floor for each cell (cycles on/off), giving $16$ measurements per
seed. With 5 seeds and 2 datasets (Bitcoin OTC + Epinions) this is
$160$ runs total. At observed warm-cache training cost ($\sim 30$ s
per Bitcoin OTC run, $\sim 200$ s per Epinions run) the full
factorial costs $\sim 1.5$ hours of CPU time per dataset.

\section{Mathematical decomposition}

Let $A, C, F \in \{0, 1\}$ denote whether axes HSiKAN, CPML, FIR
are active. Let $L : \text{dataset} \times \{A, C, F\}^3 \to [0, 1]$
be the test AUC. The factorial decomposition is

\[
  L(D, A, C, F)
  \;=\;
  \mu_D \;+\; \alpha \cdot A \;+\; \gamma \cdot C \;+\; \varphi \cdot F
  \;+\; \alpha\gamma \cdot AC \;+\; \alpha\varphi \cdot AF \;+\; \gamma\varphi \cdot CF
  \;+\; \alpha\gamma\varphi \cdot ACF
  \;+\; \varepsilon,
\]
where $\mu_D$ is the dataset's embedding-only floor and the six
greek-letter coefficients are the (single-axis, pairwise, and
three-way) effects. We solve for them by standard
$2$-level factorial regression on the $8$-cell design.

Practically interesting null hypotheses:
\begin{itemize}
  \item $\alpha = 0$: HSiKAN adds nothing (Catmull-Rom splines have
    no marginal effect over MLP/linear).
  \item $\varphi = 0$: Clifford-FIR adds nothing (closed-form
    Clifford derivative is no better than autograd MLP).
  \item $\alpha\varphi = 0$: HSiKAN and FIR are not synergistic
    (combining them gives no more than their sum).
  \item $\gamma \cdot C$ is dataset-dependent: $\gamma_{\text{Bitcoin}} = 0$
    but $\gamma_{\text{Epinions}} > 0$ would say CPML helps only on
    hub-heavy graphs.
\end{itemize}

Statistical significance: paired $t$-test on the 5-seed deltas at
each cell against its embedding-only floor. Effects with paired
$|t| < 2$ are reported as ``not distinguishable from noise.''

\section{The 8 cells (operational definition)}

Each cell is a model class. Let $X^{(0)}$ be the learnable per-vertex
embedding. Let $\mathrm{Agg}_{\text{MLP}}$ denote the default MLP
aggregator (always present as a fallback). Let
$\mathrm{Agg}_{\text{HSiKAN}}$ and $\mathrm{Agg}_{\text{FIR}}$ be
the optional aggregators of the corresponding axes. Let $\mathrm{T}$
denote a single forward through the topology (Flat = one layer of
the active aggregators; CPML = $L$ tier-stratified layers).

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{|c|c|c|c|l|}
\hline
$A$ & $C$ & $F$ & cell label & active components \\
\hline
0 & 0 & 0 & \texttt{baseline}      & MLP only, $L=1$ (the embedding-floor reference) \\
0 & 0 & 1 & \texttt{fir}           & Clifford-FIR layer, $L=1$ \\
0 & 1 & 0 & \texttt{cpml}          & MLP layer per tier, $L=3$ \\
0 & 1 & 1 & \texttt{cpml-fir}      & Clifford-FIR per tier, $L=3$ \\
1 & 0 & 0 & \texttt{hsikan}        & Catmull-Rom signed-spline layer, $L=1$ \\
1 & 0 & 1 & \texttt{hsikan-fir}    & HSiKAN and Clifford-FIR in parallel, $\alpha$-routed, $L=1$ \\
1 & 1 & 0 & \texttt{cpml-hsikan}   & HSiKAN per tier, $L=3$ \\
1 & 1 & 1 & \texttt{cpml-hsikan-fir} & HSiKAN + FIR in parallel per tier, $\alpha$-routed, $L=3$ \\
\hline
\end{tabular}
\end{center}

\textbf{Routing when both A=1 and F=1}: in the same layer, run
HSiKAN and Clifford-FIR in parallel branches, then combine via a
learnable softmax-normalised $\alpha = \mathrm{softmax}(\alpha_{\mathrm{logits}})$
of size 2:
\[
  \mathrm{Agg}_{A,F=1}(X) \;=\; \alpha_0 \cdot \mathrm{HSiKAN}(X) \;+\; \alpha_1 \cdot \mathrm{Clifford\text{-}FIR}(X).
\]
This is the same $\alpha$-routing primitive that gave the
Slashdot SOTA result via multi-arity routing — we re-use it here
as inter-aggregator routing.

\section{Affected files}

\textbf{New / modified (when reopened):}
\begin{lstlisting}
signedkan_wip/src/
├── cpml.py
│   ├── ClifFIRTierAggregator           — already shipped (today)
│   ├── HybridTierAggregator            — NEW: hosts the α-routed
│   │                                    HSiKAN + Clifford-FIR pair
│   └── CPMLConfig (extended)           — three boolean axes
└── run_cpml_factorial_v2.py            — NEW: 8-cell factorial
                                          runner with multi-seed +
                                          paired stats output
docs/plans/2026-05-11-hsikan-cpml-fir-orthogonal/
├── plan.tex                            — this doc
├── plan.pdf                            — built
├── plan.tikz                           — 8-cell hyper-cube diagram
└── plan.mmd                            — Mermaid flow of the 3 axes
\end{lstlisting}

\textbf{CORE.YAML items touched:} none. All edits are within the
existing CPML scaffolding and the factorial runner.

\section{Test strategy}

\textbf{Pre-factorial unit tests:}
\begin{itemize}
  \item Each cell's forward + backward shape test (12 LOC each;
    parametric over $(A, C, F)$).
  \item No-cycles ablation for each cell (validates the
    embedding-floor measurement; if any cell's no-cycles AUC
    diverges from \texttt{baseline}'s no-cycles AUC, the floor
    estimation is biased and the cell's lift attribution is
    invalid).
  \item HybridTierAggregator $\alpha$-routing: starts uniform
    (verify $\alpha_0 \approx \alpha_1 \approx 0.5$ at init); per-seed
    convergence trace tests confirm $\alpha$ drifts to non-uniform
    only when the data demands it.
\end{itemize}

\textbf{Factorial measurement (per seed, per dataset, 5 seeds, 2 datasets):}
\begin{itemize}
  \item For each of the 8 cells, measure paired (cell, baseline)
    validation AUC at $n_{\text{epochs}} = 50$.
  \item Embedding-only floor measured once per (seed, dataset), used
    as the paired baseline for all 8 cells.
  \item Output: 5-seed paired Δ + paired SD + paired $t$ per cell
    per dataset. Plus the 7 factorial-decomposition coefficients
    ($\alpha, \gamma, \varphi, \alpha\gamma, \alpha\varphi, \gamma\varphi, \alpha\gamma\varphi$)
    fitted on the 8-cell averages, with their estimated standard
    errors.
\end{itemize}

\textbf{Validation gate}: a single-axis effect ($\alpha$, $\gamma$, or
$\varphi$) is reported as ``architecturally non-trivial'' only if
its 5-seed paired $|t| \ge 2$ on at least one dataset. An interaction
($\alpha\varphi$, etc.) is reported as ``synergistic'' only if its
$|t| \ge 2$ AND both single-axis effects are non-trivial.

\section{Performance budget}

\begin{itemize}
  \item Per-cell training time (Epinions, 50 epochs, warm cache):
    $\sim 200$ s.
  \item 8 cells $\times$ 5 seeds $\times$ 2 datasets = 80 runs.
  \item Wall: $\sim 4-6$ hours total (sequential, single CPU).
    GPU would compress to $\sim 1$ hour.
  \item Peak RSS per run: $\le 4$ GB (matches the existing
    \texttt{test\_memory\_below\_4gb\_at\_scale} test in
    \texttt{test\_cpml.py}).
\end{itemize}

\section{Risk anticipation}

\begin{itemize}
  \item \textbf{Embedding-floor dominance}: today's prior factorial
    showed all aggregators converge to the embedding floor on
    random 80/20 Epinions. Mitigation: run on a structurally-harder
    split (e.g.\ time-stratified for Bitcoin OTC, or
    inductive-subset for Epinions) where the embedding-lookup
    advantage is weaker. Specify split protocol per dataset.
  \item \textbf{$\alpha$-route degeneracy}: if HSiKAN and FIR
    branches collapse to one with $\alpha_0 = 1$ or $\alpha_1 = 1$,
    the hybrid cells are not meaningfully different from
    single-aggregator cells. Mitigation: report observed $\alpha$
    distribution at end-of-training; if degenerate, the hybrid
    cells become hygienic checks rather than novel claims.
  \item \textbf{Statistical power}: 5 seeds with paired $t$-test
    can resolve effects $\ge \sim 1.5$ pp at $\alpha=0.05$ when the
    per-seed SD is $\sim 1$ pp. Effects below this floor are not
    distinguishable from noise. Mitigation: report 95\% confidence
    intervals on each coefficient; if intervals span zero, declare
    null result, not ``no effect.''
  \item \textbf{Cycle-cache fingerprint drift}: env-var-triggered
    cache invalidation already caught one silent bug today. Any
    new env var in the factorial (e.g.\ \texttt{HSIKAN\_AGGREGATOR\_KIND})
    must be added to \texttt{cycle\_cache.\_topk\_fingerprint}.
    Already enforced by \texttt{feedback\_cycle\_cache\_fingerprint.md}.
\end{itemize}

\section{Rollback path}

The 8-cell factorial is additive; no existing pipeline is modified.
If results are uninformative, the factorial runner and the
HybridTierAggregator can be ignored. The other architectural plans
(Nagare, CPML, HymeKo-Nagare-to-HSMM) are independent.

\section{Sequencing}

\begin{enumerate}
  \item Implement HybridTierAggregator + extend CPMLConfig (today's
    1-hour task; trivial because Clifford-FIR is already a
    TierAggregator).
  \item Unit tests for forward/backward shapes + $\alpha$-routing
    init.
  \item Single-seed dry run (8 cells × 1 seed × 2 datasets = 16
    runs, $\sim 1$ hour) to validate runner.
  \item Full 5-seed sweep ($\sim 4-6$ hours) on the overnight queue.
  \item Factorial decomposition + report.
\end{enumerate}

\section{Hypothesis register}

Pre-registered hypotheses to test (with priors):
\begin{enumerate}
  \item \textbf{H1 (FIR helps where HSiKAN doesn't)}: $\varphi > 0$
    on Bitcoin OTC, $\alpha \approx 0$ on Bitcoin OTC. \emph{Prior:}
    HSiKAN's spline expressiveness is wasted on small dense
    graphs; FIR's tiny parameter count regularises better.
  \item \textbf{H2 (HSiKAN helps where FIR doesn't)}: $\alpha > 0$
    on Slashdot/Epinions (large, heterogeneous), $\varphi \approx 0$.
    \emph{Prior:} Catmull-Rom spline captures non-linear cycle
    importance that linear FIR misses.
  \item \textbf{H3 (CPML helps multi-scale graphs only)}:
    $\gamma > 0$ on Epinions/WikiSigned (degree-skewed),
    $\gamma \approx 0$ on Bitcoin OTC (uniform). \emph{Prior:}
    tier-stratified depth needs a multi-scale to exploit.
  \item \textbf{H4 (HSiKAN + FIR synergistic on real-world graphs)}:
    $\alpha\varphi > 0$ on Slashdot, Epinions, WikiSigned. \emph{Prior:}
    spline captures non-linearity, FIR captures sign-structure,
    they decompose orthogonally and compose constructively.
\end{enumerate}

Pre-registration is the discipline that keeps the result honest;
the bug-find pattern of the day suggests this is exactly the kind
of work that needs it.

\section{Venue}

\textbf{Primary}: ICLR 2027 (clean architectural-ablation paper
with formal factorial decomposition and pre-registered hypotheses).

\textbf{Secondary}: NeurIPS 2026 GLF workshop (small-paper version
of the negative-result decompositions if the headline factorial
fails to find a single-axis effect).

\textbf{Tertiary}: KDD applied (companion to the Slashdot-SOTA
paper; uses the 3-axis factorial as the ablation section that
justifies the chosen model architecture).

\end{document}
