% SignedKAN — Regulariser, Pruning, and Multi-Layer Technical Note
% Companion to the WiP paper. Catalogues every accuracy lever explored
% during development with quantified deltas, plus the combined
% "kitchen sink" scenario that stacks the working ones.

\documentclass[11pt, a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsfonts}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}

\title{SignedKAN: Regulariser, Pruning, and Multi-Layer Technical Note}
\author{Csaba Hajdu}
\date{}

\begin{document}
\maketitle

\begin{abstract}
A companion technical note to the SignedKAN WiP submission,
catalogues every accuracy lever explored during development
together with the combined ``kitchen sink'' configuration that
stacks the working ones. Each lever is reported with its
measured contribution to AUC and macro-$F_1$ on Bitcoin Alpha
and Bitcoin OTC. All numbers are trained from scratch, three
seeds per cell, $h\!=\!32$, $G\!=\!5$, $200$ epochs with early
stopping on a single mid-range GPU.
\end{abstract}

\section{The Four-Part Claim}

The headline contribution of the WiP paper is a four-part claim
about the trained-from-scratch SignedKAN architecture:
(i) competitive AUC on signed-graph link-sign prediction,
(ii) 77--78\% pruning resilience with AUC \emph{improving} by
$+0.016$--$+0.021$, (iii) sinusoidal symbolic distillation in
80--91\% of surviving activations, basis-invariant across three
spline parameterisations, (iv) sub-minute training time on a
single mid-range GPU.

This note adds quantified detail on the regulariser landscape that
shapes the trained model and the multi-layer architecture that the
final scenario uses.

\section{Regulariser Landscape}

We tested a family of regularisers, none of which alone closes the
residual gap to engineered signed-GNN baselines but which compose
to deliver the macro-$F_1$ wins reported in the WiP paper. Each
regulariser is independently switchable via a flag in
\texttt{run\_compare.run\_one}.

\subsection{Spectral entropy with Lyapunov-safe KL feedback}
\label{sec:reg-entropy}

Acts on the singular-value distribution of the node-embedding
matrix $\mathbf{A} = \mathrm{node\_embed.weight}$:
\begin{align*}
\mathbf{p}_t &= \frac{\sigma_i^2(\mathbf{A}_t)}{\sum_j \sigma_j^2(\mathbf{A}_t)}, \\
H_{\mathrm{norm}} &= \tfrac{1}{\log_2 \mathrm{rank}(\mathbf{A})} H(\mathbf{p}), \\
\Delta_{KL}(t) &= D_{KL}(\mathbf{p}_{t-1} \,\|\, \mathbf{p}_t), \\
\lambda_{\mathrm{eff}}(t) &= \mathrm{clamp}(\lambda_0 e^{-\eta \Delta_{KL}(t)},\ 0.1\lambda_0,\ 10\lambda_0), \\
\mathcal{L}_H &= \lambda_{\mathrm{eff}} \bigl[\lambda_a (H_{\mathrm{norm}} - H^*)^2 + \lambda_b H_{\mathrm{norm}}\bigr].
\end{align*}
The KL-feedback schedule reduces regularisation pressure when the
spectrum is moving fast (transient training phase) and lets it
recover when settled (steady-state). At $\lambda_0\!=\!10^{-2}$,
$H^*\!=\!0.5$, $\eta\!=\!5$: macro-$F_1$ $+0.005$ on Alpha,
$+0.007$ on OTC; AUC roughly tied.

\subsection{R2 — vertex-degree participation}
\label{sec:reg-r2}

\[
\mathcal{L}_{R_2} = \frac{\lambda}{|V|} \sum_v \deg_H(v)^2 \cdot \|\mathbf{h}_v\|_2^2,
\]
where $\deg_H(v) = |\{t : v \in t\}|$ is the number of triads
incident to vertex $v$. Penalises hub vertices that dominate the
aggregation. At $\lambda\!=\!0.05$, R2 is the only single regulariser
that delivers a \emph{simultaneous} AUC and macro-$F_1$ win on Alpha
($+0.005$ AUC, $+0.008$ F1m), plus $+0.020$ macro-$F_1$ on OTC at
neutral AUC.

\subsection{Hypergraph triad loss (Cartwright--Harary)}
\label{sec:reg-triad}

Per-triad margin loss with structural balance indicator:
\[
s(t) = \sum_{(i, j) \in t} \sigma_{ij}\, \hat{\mathbf{h}}_i \cdot \hat{\mathbf{h}}_j, \quad
\beta_t = \prod_{(i, j)} \sigma_{ij}, \quad
\mathcal{L}_T = \mathrm{relu}(m - \beta_t s(t)).
\]
At $\alpha\!=\!1.0$, $m\!=\!0.5$ in $\mathcal{L} = \mathcal{L}_{\mathrm{BCE}} + \alpha \mathcal{L}_T$:
the best macro-$F_1$ on OTC measured anywhere in the saga, $0.775$
($+0.029$ over EC). Encodes Cartwright--Harary balance theory in the
loss rather than only in the architecture.

\subsection{Signed-triad attention with attention-entropy}
\label{sec:reg-attention}

Replace the mean-pool aggregation $\mathbf{h}_e = \tfrac{1}{|T_e|}\sum_t \mathbf{h}_t$
with $\mathbf{h}_e = \sum_t \tanh(\mathbf{a}^\top [\mathbf{h}_u; \mathbf{h}_v; \mathbf{h}_t]) \mathbf{h}_t$
(\emph{tanh}, not softmax — signed weights), regularised by
per-edge attention entropy
$\mathcal{L}_{ae} = -\lambda_{ae}\,\mathrm{mean}_e\, H(|\boldsymbol{\alpha}_e| / \|\boldsymbol{\alpha}_e\|_1)$
to prevent collapse onto a single triad. At $\lambda_{ae}\!=\!0.01$:
the best Alpha macro-$F_1$ measured anywhere, $0.742$ ($+0.044$ over EC).

\subsection{Cross-branch information regulariser}
\label{sec:reg-crossbranch}

Pairwise cosine penalty between sign-branch coefficient tensors,
forces the per-sign branches to learn distinct functions:
\[
\mathcal{L}_{\mathrm{cross}} = \frac{\lambda}{|\mathrm{pairs}|} \sum_{i < j} \mathrm{mean}_c\, |\cos(\mathbf{c}^{(i)}_{c, :}, \mathbf{c}^{(j)}_{c, :})|.
\]
Modest macro-$F_1$ wins ($+0.005$--$+0.020$); does not compose with
the spectral entropy reg above.

\subsection{Hyperedge-density (HD) regulariser — null result}
\label{sec:reg-hd}

\[
\mathcal{L}_{HD} = \frac{\lambda}{|T|} \sum_t d(t) \|\mathbf{h}_t\|^2,
\]
$d(t) = $ number of other triads sharing at least one vertex with
$t$. Across $\lambda \in \{10^{-3}, 10^{-2}, 5\!\times\!10^{-2}\}$:
roughly neutral on AUC, slightly negative on macro-$F_1$.
Mechanistic reading: HD is duplicative of spectral entropy
(both penalise embedding magnitude weighted by a structural
quantity). Reported here as honest null.

\section{Pruning and Symbolic Distillation}

After training, every $(\sigma\text{-branch}, c)$ inner and outer
spline has a coefficient vector $\mathbf{c}^{(s, c)}$. Per-spline
activity is its $L_2$ norm; pruning sets the vector to zero when
its norm is below a threshold $\tau$.

\textbf{Headline}: at $\tau\!=\!1.8$, $77$--$78\%$ of the $128$
inner+outer splines are zeroed, AUC \emph{improves} by $+0.021$ on
Alpha and $+0.016$ on OTC (median over three seeds), macro-$F_1$
declines modestly. AUC only collapses past $\tau\!=\!2.5$
($\sim\!85\%$ pruning).

Surviving splines are sampled on a $200$-point grid in $[-1, 1]$
and fitted to a small symbolic library. \textbf{Headline}:
$80$--$91\%$ of (branch, channel) splines fit best as
$a \sin(\omega x + \phi) + c$, with cubics as the runner-up
($6$--$11\%$). The pattern is invariant across B-spline,
Catmull--Rom, and Kochanek--Bartels parameterisations.

\section{Multi-Layer with LayerNorm and Weight-Sharing}

The naive multi-layer extension (independent SignedKAN layers,
mean-pool inter-layer aggregation, classifier sees only last layer)
is empirically worse than $L\!=\!1$ on this benchmark. Three fixes
together (Jumping Knowledge across layers, sum-pool inter-layer
aggregation, LayerNorm on $\mathbf{h}_v$ between layers) deliver
the multi-layer win, and \emph{weight sharing} across layers --- one
shared SignedKANLayer applied $L$ times --- adds the
parameter-efficiency the link-sign benchmark needed.

\textbf{Headline}: $L\!=\!3$ jkCAT-sum + LN + share gives OTC AUC
$0.8729$ ($+0.0029$ over $L\!=\!1$ EC) and macro-$F_1$ $0.7613$
($+0.015$), at $\sim\!1\%$ \emph{fewer} parameters than plain
$L\!=\!2$ due to weight sharing. The recurrent multi-layer view
(same transform applied $L$ times to refine vertex embeddings)
beats the depth-stacked view (independent transforms) on this
benchmark.

\section{Combined Scenario: ``Kitchen Sink''}

We compose the load-bearing levers into a single scenario:
\begin{itemize}
\item Class-weighted BCE + early stopping (the EC recipe;
  closes half the AUC gap to engineered baselines on its own).
\item Spectral entropy regulariser with Lyapunov-safe KL feedback,
  $\lambda_0\!=\!10^{-2}$, $H^*\!=\!0.5$, $\eta\!=\!5$.
\item R2 vertex-degree participation, $\lambda\!=\!0.05$.
\item Multi-layer $L\!=\!3$ with weight sharing, LayerNorm between
  layers, and Jumping-Knowledge concatenation.
\item Threshold pruning at $\tau\!=\!0.5$ post-training (a
  Pareto-balanced threshold; $\tau\!=\!1.8$ trades macro-$F_1$
  for additional AUC).
\end{itemize}

The combined scenario is the recommended deployment configuration
for SignedKAN on signed-graph link-sign prediction. Its measured
performance on Bitcoin Alpha and Bitcoin OTC is reported in
Table~\ref{tab:kitchen-sink} (numbers added on completion of the
combined-scenario run).

\begin{table}[h]
\centering
\caption{Overnight cross-dataset run: $L\!=\!1$ EC baseline,
$L\!=\!1$ EC$+R_2$ (the cleanest single-regulariser win), and the
full kitchen-sink scenario. Median over three seeds. Slashdot
results at $h\!=\!8$ + Catmull--Rom; the kitchen-sink configuration
on Slashdot OOMed at $h\!=\!8$ even with weight-sharing
(multi-layer plus the 580k triads exceed the 7.6 GB GPU budget).}
\label{tab:kitchen-sink}
\begin{tabular}{@{}lrrrrrr@{}}
\toprule
recipe & \multicolumn{2}{c}{Bitcoin Alpha} & \multicolumn{2}{c}{Bitcoin OTC} & \multicolumn{2}{c}{Slashdot} \\
       & AUC & F1m & AUC & F1m & AUC & F1m \\
\midrule
$L\!=\!1$ EC (baseline) & $0.8547$ & $0.6975$ & $0.8700$ & $0.7459$ & $0.7601$ & $0.5278$ \\
$L\!=\!1$ EC$+R_2$      & $\mathbf{0.8593}$ & $\mathbf{0.7051}$ & $0.8687$ & $\mathbf{0.7655}$ & $0.7600$ & $0.5277$ \\
Kitchen-sink            & $0.8258$ & $0.6537$ & $0.8706$ & $0.7526$ & ---  & ---  \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Empirical verdict on the kitchen sink.} Stacking all
positive-effect levers does not compose into a strict win. The
multi-layer machinery in the kitchen-sink config over-parameterises
the small Bitcoin Alpha fixture, costing $-0.029$ AUC and
$-0.044$ macro-$F_1$ relative to the $L\!=\!1$ baseline; on
Bitcoin OTC the kitchen-sink ties EC$+R_2$ on AUC and is slightly
behind on macro-$F_1$. \textbf{The recommended deployment
configuration is therefore $L\!=\!1$ EC$+R_2$}, not the full
stack: a single-layer SignedKAN with the EC training recipe and
the $R_2$ vertex-degree participation regulariser at
$\lambda\!=\!0.05$. Each additional lever (entropy with KL
feedback, multi-layer with LayerNorm $+$ weight-sharing,
heterogeneous skip placement) is worth keeping in the codebase
as an optional flag for fixtures or metrics where it specifically
helps, but not all-on by default.

\section{Levers Not in the Headline Scenario}

Several levers worked on individual cells but did not
unambiguously stack with the rest. They remain in the codebase as
optional flags; we report them here as honest catalogue entries.

\begin{itemize}
\item \textbf{Hypergraph triad loss} (\S\ref{sec:reg-triad}). Best
  macro-$F_1$ on OTC at $0.775$, but the AUC tradeoff on Alpha
  ($-0.015$) is stronger than R2's. Consider for OTC-only
  deployments where macro-$F_1$ is the load-bearing metric.
\item \textbf{Signed-triad attention with attention-entropy}
  (\S\ref{sec:reg-attention}). Best Alpha macro-$F_1$ at $0.742$,
  AUC roughly tied on OTC; replaces the mean-pool aggregation with
  a tanh-attention head.
\item \textbf{Bilinear endpoint head}. Best simultaneous OTC AUC
  ($0.8754$) and OTC macro-$F_1$ ($0.8096$) of the entire saga,
  but high variance on Alpha. Useful when OTC-class fixtures are
  the deployment target.
\item \textbf{Heterogeneous skip placement}. (inner: residual,
  outer: none) gives a free OTC AUC bump ($+0.0033$) with zero
  added parameters. Validates the CV head-neck-spine analogy
  (the inner per-vertex spline is the ``backbone'').
\item \textbf{Catmull--Rom activation}. $11\times$ per-iter
  speedup at essentially the same accuracy; Section~V of the WiP
  paper recommends it for cost-sensitive applications.
\end{itemize}

\section{Negative Results (for Completeness)}

\begin{itemize}
\item \textbf{Hyperedge-density regulariser} (\S\ref{sec:reg-hd}):
  duplicative of spectral entropy.
\item \textbf{$L_1$-sparsity during training}: at $\lambda \geq 10^{-4}$
  the model collapses to predicting a single class globally.
  $L_1$ forces zero too hard; cannot reproduce post-hoc
  threshold-pruning's gentle selective sparsity.
\item \textbf{Iterative pruning + retraining}: retraining
  surviving splines after pruning UNDOES the post-hoc pruning
  regularisation benefit on most seeds. Post-hoc threshold-pruning
  at $\tau\!=\!1.8$ is genuinely the optimal one-shot intervention.
\item \textbf{Spectral-Laplacian eigenvector init}: looked
  promising on a single seed; multi-seed median is roughly neutral.
  Pre-training is a stronger structural-prior intervention than
  initialisation alone.
\end{itemize}

\section{Future Directions}

See \texttt{signedkan\_wip/FUTURE\_DIRECTIONS.md} for the full
catalogue. Highlights:
\begin{itemize}
\item Variable-arity hyperedges (clique clusters, $k$-uniform
  generalisation; unlocks the structural prior in $R_3$).
\item LSTM-style gated hyperedge blocks (independent forget /
  input / output gates, generalising the residual / highway / JK
  family).
\item Total-model parameter compression beyond splines (product
  quantisation, top-$k$ sparse embeddings).
\item Matched-protocol re-implementation of engineered baselines
  for a strict apples-to-apples AUC comparison.
\end{itemize}

\end{document}
