\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{enumitem}
\setlist{nosep,leftmargin=*}
\lstset{basicstyle=\ttfamily\small,frame=single,breaklines=true}
\title{Plan: HyMeYOLO --- $k$-cycle structural attention\\
       on convolutional feature maps for object detection}
\author{HyMeKo / HSiKAN}
\date{2026-05-11}
\begin{document}
\maketitle

\section{Scope and goal}
HyMeYOLO inserts a \textbf{$k$-cycle structural attention layer} between
a YOLO-style convolutional backbone and the detection head.  The
$k$-cycle layer operates on a signed graph induced by spatial feature
similarities: vertices are spatial locations on a feature map, edges
are local-neighbourhood connections, and signs are
$\text{sign}(\langle f_i, f_j \rangle)$ for feature vectors $f_i,
f_j$.  Top-$K$ balanced cycles enumerated on this signed graph
contribute extra channels (cycle-density heatmaps) to the detection
head's input.

\textbf{Why now}: pure HSiKAN-on-vision was a documented null result
(\texttt{project\_hsikan\_vision\_redo\_2026\_05\_06.md}: HGNN-style
on MNIST/Fashion lost to CNN at h=16/n=1).  Vision regime needs
translation equivariance, which CNN provides and HGNN-on-pixels lacks.
\textbf{HyMeYOLO keeps the CNN backbone} and adds signed-cycle
reasoning \emph{on top of conv features}.  Tonight's infrastructure
(SoA cycle enum, vertex prefilter, FPN-tiered caps, dedup
optimisations) brings the per-feature-map cycle cost from $\sim 100$~ms
down to $\sim 20$~ms --- fast enough to be a single GPU step.

\textbf{Concrete goal}: a feasibility experiment on Cluttered MNIST
(synthetic, $5{,}000$ images, 1--3 overlapping digits each) measures
whether HyMeYOLO improves mAP$_{0.5}$ vs a matched vanilla
YOLO-style baseline at the \textbf{same backbone, same train budget}.
If feasibility passes ($\geq +0.01$ mAP$_{0.5}$, paired across 5
seeds), promote to PASCAL VOC 2007.  Only if that passes, promote
to COCO.

\section{Affected files}
\begin{itemize}
\item \texttt{hymeko\_yolo/} --- NEW.  Rust+Python crate/module
  containing the cycle-attention layer, the synthetic data generator,
  and the YOLO-style detection head.  Layout:
  \begin{lstlisting}
hymeko_yolo/
├── Cargo.toml                   (Rust crate, depends hymeko_graph + hymeko_py)
├── pyproject.toml
├── README.md
├── src/
│   ├── lib.rs                   Rust ndarray helpers (sign, neighbourhood)
│   ├── cycle_features.rs        Signed-feature-graph builder (Rust)
│   └── python/
│       ├── __init__.py
│       ├── synthetic.py         Cluttered MNIST generator
│       ├── backbone.py          ResNet-18 / DarkNet-Tiny wrappers
│       ├── cycle_layer.py       PyTorch module wrapping the Rust enum
│       ├── detection_head.py    YOLO-style anchor + class head
│       ├── train.py             Training loop, mAP eval
│       └── viz.py               Cycle-heatmap overlays for sanity check
└── tests/
    ├── test_synthetic.py        synth data invariants
    ├── test_signed_graph_from_feats.py   sign/neighbour correctness
    ├── test_cycle_layer.py      forward/backward parity
    └── test_train_smoke.py      end-to-end 5-image smoke
  \end{lstlisting}
\item No changes to existing \texttt{hymeko\_graph}, \texttt{hymeko\_py},
  or \texttt{signedkan\_wip} --- HyMeYOLO is a \textbf{consumer} of
  the cycle infrastructure, not a modifier.
\end{itemize}

\section{CORE.YAML items touched}
\textbf{None.}  HyMeYOLO is a new module in its own directory.

\section{Interface changes}

\subsection*{Rust: signed feature graph builder}
\begin{lstlisting}[language=rust]
// hymeko_yolo/src/cycle_features.rs
//
// Build a SignedGraph from a (B, C, H, W) feature map.  Per-spatial-
// position feature vector f_{h,w} ∈ R^C.  Edges are 8-connected;
// signs are sign(<f_i, f_j>) of normalised feature vectors.
pub fn build_signed_feature_graph(
    feats: &[f32],           // shape (H*W, C) row-major after batch unbinding
    h: usize, w: usize, c: usize,
    neighbourhood: u8,       // 4 | 8 | 16 (k-NN ring)
    sign_threshold: f32,     // tau; |dot| < tau → drop edge
) -> SignedGraph;
\end{lstlisting}

\subsection*{Python: PyTorch cycle-attention layer}
\begin{lstlisting}[language=python]
# hymeko_yolo/src/python/cycle_layer.py
class CycleAttention(nn.Module):
    """k-cycle structural attention on conv feature maps.

    Forward:
      x : (B, C, H, W)   conv feature map
      → for each batch item:
          1. f_normalised = F.normalize(x.view(B,C,-1), dim=1)
          2. signed_graph = hymeko_yolo.build_signed_feature_graph(...)
          3. cycles = hymeko.enumerate_top_k_per_vertex_cycles_signed_filtered_rs(
                 ..., k_len=self.k, m_per_vertex=self.m,
                 'fraction_negative', 'balance',
                 'compose:degree,triangle', 2,
             )
          4. cycle_heatmap : (B, k+1, H, W)
             - 1 channel = cycle-membership count per spatial position
             - k channels = score of cycles whose centroid lands here
                            stratified by canonical-rotation-position
      out : torch.cat([x, cycle_heatmap], dim=1)   (B, C+k+1, H, W)
    """
    def __init__(self, k=3, m=16, neighbourhood=8, sign_threshold=0.1,
                 detach_graph=True):
        ...
    def forward(self, x): ...
\end{lstlisting}

\textbf{Differentiability caveat}: the cycle enumeration itself is
\emph{not} differentiable w.r.t.\ feature values.  The
\texttt{detach\_graph} flag (default \texttt{True}) means gradients
do not flow through the graph-building step; the cycle-heatmap is
treated as an extra feature.  This is the v1 approach (\textbf{cycle
features as auxiliary signal}, not learned via end-to-end
attention).  A future v2 could implement differentiable subgraph
selection via straight-through-estimator on the edge-sign threshold.

\subsection*{Python: detection head}
\begin{lstlisting}[language=python]
# hymeko_yolo/src/python/detection_head.py
class YOLOHead(nn.Module):
    """Standard YOLO-style head: anchors × (4 box + 1 conf + n_cls).
    Accepts (B, C', H, W) input where C' = backbone_C + cycle_channels.
    """
\end{lstlisting}

\section{Test strategy (CLAUDE.md \S3)}

\subsection*{Unit (Rust)}
\begin{itemize}
\item \texttt{build\_signed\_feature\_graph}: 3×3 hand-coded feature
  map → expected edge set + signs.
\item Sign threshold $\tau = 0$ produces all edges with $\pm 1$
  signs; $\tau \to \infty$ produces no edges.
\item 4-, 8-, 16-neighbourhood produce expected edge counts.
\end{itemize}

\subsection*{Unit (Python)}
\begin{itemize}
\item \texttt{CycleAttention} forward on a fixed seed produces
  bit-deterministic output.
\item Cycle heatmap shape \texttt{(B, k+1, H, W)} and dtype
  \texttt{torch.float32}.
\item Cluttered MNIST generator: 2 digits / image, 1 digit / image,
  and 3 digits / image cases all produce valid bbox annotations
  (no negative widths/heights, all in-image).
\item Backward pass on the detection head only (graph detached)
  produces non-zero gradients for backbone + head.
\end{itemize}

\subsection*{Integration}
\begin{itemize}
\item End-to-end 5-image smoke: 1 epoch on 5 Cluttered MNIST images,
  check loss decreases.
\item mAP eval on a held-out 10-image fixture: model output structure
  matches detection-eval expectations.
\end{itemize}

\subsection*{Performance test}
\begin{itemize}
\item \texttt{CycleAttention.forward(x)} on a \texttt{(2, 64, 32, 32)}
  input: median wall $\le 30$~ms (5-iter post-warmup, RTX 2070 SUPER).
\item Memory: forward-pass extra RSS $\le 200$~MB per image (cycle
  storage + heatmap).
\item End-to-end Cluttered MNIST train epoch: $\le 60$~s (200 batches
  × 32 images).
\end{itemize}

\section{Performance budget}
\begin{itemize}
\item Feasibility wall: $\le 60$~min total for 50-epoch training on
  Cluttered MNIST (RTX 2070 SUPER, batch=32).
\item Memory: $\le 6$~GB GPU, $\le 8$~GB CPU.
\item mAP$_{0.5}$ vs vanilla baseline: $\ge +0.01$ paired 5-seed.
\item If feasibility passes: PASCAL VOC training budget $\le 6$~h
  for full 50-epoch run.
\end{itemize}

\section{Risk anticipation (CLAUDE.md \S2)}

\subsection*{Performance-contract preservation}
The cycle layer consumes \texttt{hymeko}'s PyO3 bindings as-is.
The post-SoA-refactor wall budget (\texttt{enumerate\_top\_k\_per\_vertex\_cycles\_signed\_filtered\_rs}
at $\sim 100$~ms on a 1k-vertex graph is the contract we depend on).
If this regresses, HyMeYOLO's per-batch cost blows up
proportionally.  Mitigation: pin a \texttt{bench\_cycle\_layer.py}
that asserts the 30~ms forward-pass budget; add to CI.

\subsection*{Aspirational comments to honour}
The HSiKAN-vision null
(\texttt{project\_hsikan\_vision\_redo\_2026\_05\_06}) says ``vision
regime favours translation equivariance over signed-cycle
structure''.  HyMeYOLO sidesteps this by keeping the CNN backbone;
the cycle layer is \emph{additive}, not replacing the backbone.  If
the feasibility experiment fails (no mAP improvement), the conclusion
is consistent with the prior null: signed-cycle structure alone is
not enough for vision tasks.  We then \emph{do not} promote to
PASCAL.

\subsection*{Worst-case input}
Cluttered MNIST: $32 \times 32$ feature map (so 1024 vertices),
batch size 32, $k=3$, $m=16$.  Total cycles enumerated per batch
$\approx 32 \times 16{,}000 = 512{,}000$ cycle ids.  Subsampled to
$\le 1{,}000$ cycles per image before feeding the heatmap; this
keeps the post-enum memory bounded.

\subsection*{What could go wrong}
\begin{itemize}
\item \textbf{Cycle features are constant-ish across positions}: if
  every position participates in many cycles, the cycle-density
  heatmap looks like uniform noise and the detection head can't use
  it.  Mitigation: visualise the heatmap on a couple of images
  early; if it's flat, the sign threshold or neighbourhood is
  wrong, not the architecture.
\item \textbf{Sign threshold $\tau$ is the wrong knob}: cosine-sim
  sign depends on normalisation; might need learnable scale.  Add as
  a v2 path if v1 fails.
\item \textbf{Synthetic→real gap}: Cluttered MNIST may not transfer
  to PASCAL.  Mitigation: the plan has a halt gate after each
  dataset stage; we don't promote to the next dataset unless the
  previous one shows signal.
\item \textbf{Backbone choice biases the result}: ResNet-18 is a
  reasonable default but may interact poorly with cycle features.
  Mitigation: also try DarkNet-Tiny.
\end{itemize}

\section{Halt conditions}
\begin{itemize}
\item Cluttered MNIST feasibility: HyMeYOLO mAP$_{0.5}$ $<$ vanilla
  baseline + 0.005 (paired 5-seed) → halt, document as null
  result.
\item Cycle layer wall $> 50$~ms on a $32{\times}32$ feature map →
  performance contract violated, halt and investigate.
\item PASCAL VOC mAP$_{0.5}$ regresses vs the matched baseline →
  halt, do not promote to COCO.
\end{itemize}

\section{Implementation order}

\subsection*{Phase 0 (this plan, $\sim$3-5 days)}
\begin{enumerate}
\item Rust: \texttt{build\_signed\_feature\_graph} + PyO3 binding
  exposing it as \texttt{hymeko.build\_signed\_feature\_graph\_rs}.
\item Python: \texttt{synthetic.py} Cluttered MNIST generator + 8
  unit tests.
\item Python: \texttt{cycle\_layer.py} \texttt{CycleAttention}
  forward (graph detached); unit tests for shape/determinism.
\item Python: \texttt{detection\_head.py} YOLO-style head;
  forward + backward unit tests.
\item Python: \texttt{train.py} 5-image smoke test; loss decreases
  in 1 epoch.
\item Python: \texttt{viz.py} qualitative visualisation; manual
  inspection of cycle-density heatmaps.
\end{enumerate}

\subsection*{Phase 1 (gated on phase 0 passing, $\sim$2 days)}
\begin{enumerate}
\item Full Cluttered MNIST training: HyMeYOLO vs vanilla baseline,
  5-seed paired, 50 epochs each.  Wall budget $\le 60$~min total.
\item Sweep cycle layer hyperparameters ($k \in \{3, 4\}$,
  $m \in \{8, 16, 32\}$, $\tau \in \{0, 0.1, 0.2\}$).
\item Report: mAP, wall time, peak memory, cycle-heatmap stats.
\end{enumerate}

\subsection*{Phase 2 (gated on Phase 1, $\sim$1 week)}
\begin{enumerate}
\item PASCAL VOC 2007 training, 50 epochs, ResNet-18 backbone.
\item Compare to YOLOv5n baseline trained on the same data with
  the same backbone.  Report mAP$_{0.5}$, mAP$_{0.5:0.95}$,
  per-class AP.
\end{enumerate}

\subsection*{Phase 3 (gated on Phase 2)}
COCO subset, then full COCO.

\section{Rollback path}
HyMeYOLO is a new directory; nothing depends on it.  Reverting
removes \texttt{hymeko\_yolo/} cleanly.

\section{Empty-plan-dir hygiene}
This plan dir is fully populated (4 artefacts).  If abandoned, the
dir gets a \texttt{HALTED.md} with the abort reason.

\end{document}
