\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{enumitem}
\setlist{nosep,leftmargin=*}
\lstset{basicstyle=\ttfamily\small,frame=single,breaklines=true,
        keywordstyle=\color{blue!60!black},stringstyle=\color{red!50!black}}
\title{Plan: HSIKAN Parameter Derivation\\
       (collapsing $\sim$25 env vars into a small set of fundamental knobs)}
\author{HyMeKo / HSiKAN}
\date{2026-05-12}
\begin{document}
\maketitle

\section{Scope and goal}
HSIKAN has accumulated approximately 25 environment variables plus 5
CLI arguments controlling cycle enumeration, vertex selection,
attention, gates, kernels, training, and protocol --- many of them
coupled (e.g.\ \texttt{HSIKAN\_TOPK\_M\_V\_MAX} defaults to
\texttt{HSIKAN\_TOPK\_K}; the various per-arity \texttt{MAX\_K} caps
are typically set together; \texttt{HSIKAN\_CYCLE\_BATCH} is a function
of available memory).  This makes experiments hard to reproduce, hard
to compare across runs, and easy to misconfigure (today's morning OOM
is a direct consequence of a single env-var addition that bypassed an
implicit memory contract --- see
\texttt{project\_rust\_topk\_memory\_followup\_2026\_05\_10}).

\textbf{Concrete goal.}  Introduce an \texttt{HSIKANConfig} dataclass
that exposes a small set of fundamental knobs (5-10) and \emph{derives}
the rest via documented rules.  Each derived parameter has a
\texttt{@cached\_property} with the rule encoded in code.  Env-vars
remain as escape-hatch overrides so existing scripts keep working.

\section{Affected files}
\begin{itemize}
\item \texttt{signedkan\_wip/src/config.py} --- NEW.  The
  \texttt{HSIKANConfig} dataclass + derivation rules +
  dataset-profile defaults.
\item \texttt{signedkan\_wip/src/run\_final\_cell.py} --- replace the
  $\sim$15 \texttt{os.environ.get(...)} calls at the top of
  \texttt{cell\_signed\_graph} with one \texttt{HSIKANConfig.from\_env(g, args)}.
  Propagate the config object downstream; each callee that currently
  reads env-vars switches to \texttt{config.X}.
\item \texttt{signedkan\_wip/src/n\_tuples.py} --- the env-var
  dispatch in \texttt{\_enumerate\_cycles\_fast} reads from a
  \texttt{HSIKANConfig} if provided, else falls back to env-vars.
  Backwards-compatible.
\item \texttt{signedkan\_wip/tests/test\_config.py} --- NEW.  Tests
  for each derivation rule, dataset-profile defaults, env-var
  override priority, and round-trip serialisation.
\item \texttt{signedkan\_wip/src/mixed\_arity\_signedkan.py} ---
  if any env-var reads exist here, switch to config dotted access.
\end{itemize}

\section{CORE.YAML items touched}
\textbf{None.}  All affected files are in \texttt{signedkan\_wip} which
is not under \texttt{CORE.YAML}'s \texttt{crates}/\texttt{files}/\texttt{globs}.

\section{The fundamental knobs (user-facing)}
\begin{lstlisting}[language=python]
@dataclass
class HSIKANConfig:
    # ---- ALGORITHMIC (genuinely independent) ----
    dataset: str                          # epinions | slashdot | ...
    seed: int = 0
    n_epochs: int = 80
    hidden: int = 4
    mixed_tuples: tuple = ("c2","c3","c4","c5","w2","w3")
    pruner: str = "balance"
    scorer: str = "fraction_negative"
    attention_kind: str = "quaternion"
    highway_kind: str = "edge_cr"

    # ---- BUDGET (one knob drives ALL caps) ----
    memory_budget_gb: float = 8.0
    target_total_cycles: int = 600_000

    # ---- STRATEGY (one knob drives selection params) ----
    selection_aggressiveness: float = 0.5   # 0=legacy, 1=tiered+MSG

    # ---- QUALITY (independent) ----
    entropy_lambda: float = 0.0
    sparse_attn_k: int | None = None
    strict_protocol: bool = False

    # ---- BACKEND (independent toggles) ----
    triton_kernel: bool = True
    triton_backward: bool = True
    direct_messaging: bool = False
    per_edge_gate: bool = False
\end{lstlisting}

\section{Derivation rules}

\subsection*{From \texttt{target\_total\_cycles} $\rightarrow$ per-arity caps}
FPN-style decay across cycle lengths --- shorter cycles are abundant
but local; longer cycles are rare but globally informative.  Weights
sum to 1.0:
\[ \text{cap}_k = \text{target\_total} \times w_k, \quad
   w \in \{3{:}0.35,\, 4{:}0.25,\, 5{:}0.18,\, 6{:}0.12,\, 7{:}0.10\} \]

\subsection*{From \texttt{selection\_aggressiveness} $\rightarrow$ topk\_mode + m\_v\_c + vertex\_filter}
\begin{center}
\begin{tabular}{lll}
\toprule
agg.range & \texttt{topk\_mode} & \texttt{vertex\_filter} \\
\midrule
$[0.0, 0.2)$ & \texttt{per\_vertex}          & \texttt{none} \\
$[0.2, 0.7)$ & \texttt{per\_vertex\_adaptive} & \texttt{degree>=2} \\
$[0.7, 1.0]$ & \texttt{per\_vertex\_tiered}   & \texttt{compose:degree>=2,triangle} \\
\bottomrule
\end{tabular}
\end{center}

\[ m_v\_c = 1.0 + (0.5 - \text{aggressiveness}) \]

\subsection*{From \texttt{memory\_budget\_gb} + graph stats $\rightarrow$ cycle\_batch}
At load time we know $|V|$, $|E|$, and \texttt{hidden}.  We compute
the expected per-chunk activation memory $b_{chunk}$ and set:
\[ \text{cycle\_batch} = \max\!\left(500,\ \lfloor 0.3 \cdot \text{memory\_budget\_gb} \cdot 10^9 / b_{chunk} \rfloor\right) \]

\subsection*{From \texttt{dataset} $\rightarrow$ default profile}
\begin{lstlisting}[language=python]
DATASET_PROFILES = {
    "bitcoin_alpha": HSIKANConfig(
        memory_budget_gb=1.0,
        target_total_cycles=50_000,
        selection_aggressiveness=0.3),  # small graph, cheap to be liberal
    "bitcoin_otc": HSIKANConfig(
        memory_budget_gb=1.0,
        target_total_cycles=50_000,
        selection_aggressiveness=0.3),
    "slashdot": HSIKANConfig(
        memory_budget_gb=4.0,
        target_total_cycles=400_000,
        selection_aggressiveness=0.5),
    "epinions": HSIKANConfig(
        memory_budget_gb=8.0,
        target_total_cycles=400_000,
        selection_aggressiveness=0.6),  # large graph, prefer concentration
}
\end{lstlisting}

\section{Env-var override priority}
\begin{enumerate}
\item If \texttt{HSIKAN\_<FIELD>} is set in the environment, it wins.
\item Else if the field is in the dataset profile, use that.
\item Else use the dataclass default.
\end{enumerate}
\texttt{HSIKANConfig.from\_env(g, args)} reads the environment, layers
the dataset profile, and produces a fully resolved config.  The
resolution is logged so the experiment is reproducible.

\section{Interface changes}
\begin{itemize}
\item \texttt{HSIKANConfig} is the only new public API.
\item Existing scripts (overnight runners, 5-seed orchestrators) work
  unchanged: env-vars still parsed, just funneled through the dataclass.
\item Reports include \texttt{config.to\_json()} in provenance instead
  of the env-var dump.
\end{itemize}

\section{Test strategy (CLAUDE.md \S3)}

\subsection*{Unit (\texttt{tests/test\_config.py})}
\begin{itemize}
\item Each derivation rule has a dedicated test asserting the formula
  on hand-coded inputs.
\item Env-var override priority: setting \texttt{HSIKAN\_HIDDEN=8}
  beats the profile's \texttt{hidden=4}.
\item Dataset-profile defaults: \texttt{HSIKANConfig.from\_dataset("epinions")}
  has the documented fields.
\item Round-trip: \texttt{HSIKANConfig.from\_json(c.to\_json()) == c}.
\item \texttt{selection\_aggressiveness=0.0} bit-equals the
  pre-derivation legacy env-var soup.
\end{itemize}

\subsection*{Integration}
\begin{itemize}
\item Smoke run on a 200-vertex SBM fixture using
  \texttt{HSIKANConfig.from\_dataset("sbm\_n200\_k4\_s0")} produces
  the same AUC as the equivalent env-var config (within numerical
  noise).
\item Run with \texttt{selection\_aggressiveness=0.0} matches today's
  fixed-m=128 baseline numbers bit-for-bit on a deterministic seed.
\end{itemize}

\subsection*{Regression}
\begin{itemize}
\item All existing 5-seed orchestrators in
  \texttt{signedkan\_wip/experiments/*.sh} pass under the dataclass
  path (CI: each script runs at minimum its 1-seed=0 smoke).
\end{itemize}

\section{Performance budget}
This refactor is plumbing --- no perf change.  Test gate:
\texttt{config.from\_env(g, args)} resolution must complete in
$<100$~ms on Epinions ($g$ already loaded).

\section{Risk anticipation (CLAUDE.md \S2)}

\subsection*{Performance-contract preservation}
The recently-landed \texttt{n\_tuples.\_subsample\_arr\_to\_tuples}
(2026-05-10) pushes the \texttt{max\_cycles} cap into the numpy-array
subsample step before Python-tuple materialisation.  The dataclass
must preserve this: \texttt{config.cap\_k[k]} is passed in unchanged
to the existing call sites.  Test: the post-refactor
\texttt{construct\_k} call site has the same \texttt{max\_cycles}
argument as the pre-refactor version.

The Rust-side memory followup
(\texttt{project\_rust\_topk\_memory\_followup\_2026\_05\_10}) is
independent of this refactor.

\subsection*{Aspirational comments to honour}
\texttt{n\_tuples.py:367-369}: ``Push the cap into the enumerator so
we never materialise more than \texttt{max\_cycles} Python tuples in
memory.''  This refactor does \emph{not} address that aspirational
comment --- it's a separate Rust-side change.  The config refactor
sits cleanly above that work.

\subsection*{Worst-case input}
Epinions production-config (\texttt{c2,c3,c4,c5,w2,w3}, $h{=}4$,
80 epochs, balance pruner, edge\_cr highway, quaternion attention,
triton kernel on, seed 0--4).  This config's env-var dump is the
canonical mapping target: \texttt{HSIKANConfig.from\_dataset("epinions")}
must yield exactly the same set of resolved parameters as today's
\texttt{run\_adaptive\_mv\_5seed\_2026\_05\_10.sh}.

\subsection*{What could go wrong}
\begin{itemize}
\item \textbf{Hidden env-var reads}: there are 25+ env vars; missing
  one in the dataclass leaves a silent default.  Mitigation: a
  \texttt{config.warn\_on\_unrecognised\_env()} pass that lists every
  \texttt{HSIKAN\_*} env var on the system and warns if any aren't
  consumed by the dataclass.
\item \textbf{Bit-equivalence break}: a derivation rule produces a
  slightly different value than the env-var default
  (e.g., \texttt{cap\_k[4]} is 100\,000 today by CLI default but
  derivation puts it at 150\,000).  Mitigation: dedicated bit-equiv
  test at \texttt{selection\_aggressiveness=0.0} pinning every
  resolved field to today's defaults.
\item \textbf{Dataset-profile coupling}: \texttt{epinions} profile sets
  \texttt{memory\_budget\_gb=8.0} which on a smaller machine would
  OOM.  Mitigation: profiles can be overridden by env-vars; document
  this in \texttt{config.py} module docstring.
\end{itemize}

\section{Empty-plan-dir hygiene}
This plan dir is fully populated.  If abandoned, replace with a
\texttt{HALTED.md} containing the abort reason.

\section{Rollback path}
The refactor is additive: \texttt{HSIKANConfig} is a new module.
Existing env-var parsing in \texttt{run\_final\_cell.py} is preserved as
the fallback path.  Reverting the patch removes \texttt{config.py} and
restores the original env-var-soup top of \texttt{cell\_signed\_graph}.
Existing scripts continue to work in the rollback state.

\section{Halt conditions}
\begin{itemize}
\item Bit-equiv test fails at \texttt{selection\_aggressiveness=0.0}
  --- a derivation rule has the wrong default; halt and reconcile
  before merging.
\item Any of today's queued scripts produces a different AUC under the
  dataclass path vs the env-var path --- silent semantic regression;
  halt.
\item Documented env-var-override priority isn't observed in a test
  --- the layered resolution has a bug.
\end{itemize}

\section{Implementation order (v1 / v2 / v3)}

\subsection*{v1 (this plan)}
\begin{enumerate}
\item \texttt{HSIKANConfig} dataclass + derivation rules.
\item \texttt{from\_env\_or\_dataset} resolution + override priority.
\item All unit tests; bit-equiv test at \texttt{aggressiveness=0.0}.
\item \texttt{cell\_signed\_graph} top swapped to dataclass path.
\item Single-seed smoke on Epinions production config matching today's
  baseline AUC.
\end{enumerate}

\subsection*{v2 (separate plan)}
\begin{enumerate}
\item Expose \texttt{config} as a CLI flag (\texttt{--config path/to/json})
  for full-experiment reproducibility.
\item Wire \texttt{HSIKANConfig} into the orchestrator scripts ---
  generate scripts from config instead of duplicating env-var blocks.
\item Add a registry of named configs that produced published numbers
  (e.g., \texttt{slashdot\_edge\_cr\_5seed\_2026\_05\_09\_config.json})
  so each SOTA-producing experiment has a single-file reproduction
  artifact.
\end{enumerate}

\subsection*{v3 (much later)}
\begin{enumerate}
\item Auto-tuning: given \texttt{memory\_budget\_gb} + dataset, run a
  short calibration to find the \texttt{selection\_aggressiveness} that
  maximises AUC under the budget.  Output: a profile-derived config.
\end{enumerate}

\end{document}
