\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=2.3cm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{array}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}
\newcommand{\sig}{\ensuremath{\sigma}}
\newcommand{\Gomb}{G\"omb}
\newcommand{\Soma}{G\"ombSoma}
\newcommand{\Stim}{\Soma-Stim}
\title{Plan: \Stim{} --- stimulus-driven vision architecture\\
       \large{(content-driven hypergraph; an alternative to YOLO-style fixed-grid detection)}}
\date{2026-05-14}
\author{}
\begin{document}
\maketitle

\section{Framing: the rolling-shutter vs stimulus dichotomy}

YOLO, SSD, RetinaNet, and our own Phase 3-V \texttt{PatchGraphBuilder}
share a \emph{rolling-shutter} topology: a fixed grid of anchors at
every position $\times$ scale, evaluated exhaustively. The grid is
decided at design time, before the image arrives. Most compute is
spent on background.

\Stim{} is the content-driven alternative: initial activations
\emph{vote} for where to look, top-K salient regions become anchors,
hypergraph topology is built from those anchors per-image, and the
existing \Soma{} Walk / Polygon / Triangle layers run on the dynamic
graph. Each step is biologically motivated (Itti--Koch salience,
saccadic vision, capsule routing-by-agreement).

The architectural lever \Stim{} adds over YOLO: \textbf{Cartwright--
Harary $\sig$-balance as an anchor-selection criterion}. Salience
alone is "where is the activation strong?" $\sig$-balance is "which
$K$ regions form a structurally coherent signed substructure?" The
combination filters out high-salience-but-incoherent regions
(clutter / decoys) and surfaces high-salience-and-coherent regions
(actual objects). YOLO has no analogue --- it lacks signed primitives.

\section{Hypothesis}

\textbf{Phase 3-V} (rolling-shutter \Soma) is a baseline. \Stim{}
beats it, beats the prior HyMeYOLO \texttt{+ricci+kcycle} aggregator
($0.204 \pm 0.04$ mAP50 on Cluttered MNIST), and approaches or
matches the strong HyMeYOLO baseline \texttt{+ricci-mod}
($0.723 \pm 0.18$ mAP50). If \Stim{} reaches $\geq 0.5$ mAP50 on
the same task at the same parameter count, the stimulus-driven
sensorimotor-stack hypothesis is empirically supported.

\section{Architecture}

\subsection{Pipeline}

\begin{center}
\begin{tabular}{ll}
\toprule
Stage & Module \\
\midrule
1. Image patchify        & \texttt{PatchPatchifier} (reuses Phase 3-V \texttt{patchify}) \\
2. Initial embedding     & \texttt{Linear(patch\_dim $\to$ d\_hidden)} \\
3. Salience scoring      & \texttt{SalienceHead}: MLP $\to$ scalar per patch \\
4. Top-K anchor selection & \texttt{TopKSelector}: differentiable in training, hard at inference \\
5. Hyperedge construction & \texttt{StimulusGraphBuilder}: similarity edges + $\sig$ signs \\
6. \Soma{} layers         & Walk / Polygon / Triangle (Phase 1-3-G, unchanged) \\
7. Detection head        & per-anchor class + offset (or per-image classification) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Salience scoring}

\[
s_v \;=\; \sigma_\text{sig}\!\bigl( w_s^\top \, \text{GELU}(W_s x_v + b_s) \bigr) \;\in\; [0, 1]
\]
where $W_s, b_s, w_s$ are learnable. Trained jointly with the
detection loss; no auxiliary salience supervision needed --- the
gradient through top-K selection (Gumbel-softmax in training) pulls
$s_v$ up for anchors that contribute to correct detections.

\subsection{Top-K selection}

\emph{Training:} Gumbel-softmax over $\{s_v\}$ with temperature
$\tau$, producing a soft selection mask $m \in [0, 1]^{|V|}$ that
sums to $K$. Differentiable; gradients flow to the salience head.

\emph{Inference:} hard top-$K$ — pick the $K$ largest $s_v$. Same
graph downstream; the soft-vs-hard difference vanishes as $\tau \to 0$.

\subsection{Hyperedge construction among anchors}

Given the $K$ selected anchors $A \subset V$, build a signed
hypergraph on $A$ with the following primitives:

\begin{itemize}
\item \emph{Walks ($w_2, w_3$):} all length-2 and length-3 walks in
  the anchor-feature-similarity $k$-NN graph (each anchor connects
  to its $\nu = 4$ most-similar other anchors).
\item \emph{Polygons ($c_4, c_5, c_6$):} all small cycles in the
  same $k$-NN graph.
\item \emph{Triangles ($c_3$):} all triads of mutually-$k$NN anchors.
\item \emph{Edge signs:} for each edge $(u, v)$,
  $\sig(u, v) = \text{sign}\bigl( \langle x_u, x_v \rangle - \theta \bigr)$
  where $\theta$ is a learned threshold (effectively a feature-
  inner-product polarity).
\item \emph{Walk / polygon / triangle signs:} $\sig$-product over
  constituent edges, as before.
\end{itemize}

The hypergraph is constructed per-image at training time and
is differentiable end-to-end (top-K via Gumbel; $k$-NN via
top-$\nu$ on the soft similarity matrix).

\subsection{Detection head}

\emph{Per-anchor classification (YOLO-style):}
\[
\text{cls}(a) = W_c h_a + b_c, \qquad
\text{offset}(a) = W_o h_a + b_o
\]
where $h_a$ is the anchor's feature after \Soma{} layers. An
anchor with $s_a < \tau_\text{conf}$ is treated as background.

\emph{Per-image classification} (MNIST-only): global pool over
anchor features, single softmax. (Simpler; used as warm-up.)

\section{Affected files}

All under \texttt{signedkan\_wip/}, no CORE.YAML touches:

\begin{itemize}
\item \texttt{src/hymeko\_gomb/soma/vision/stim\_anchor.py} ---
  \texttt{SalienceHead}, \texttt{TopKSelector},
  \texttt{StimulusAnchorEncoder}.
\item \texttt{src/hymeko\_gomb/soma/vision/stim\_graph.py} ---
  \texttt{StimulusGraphBuilder} (signed-similarity $k$-NN hypergraph).
\item \texttt{src/hymeko\_gomb/soma/vision/stim\_classifier.py} ---
  \texttt{StimDrivenImageClassifier} (per-image classification head).
\item \texttt{src/hymeko\_gomb/soma/vision/stim\_detector.py} ---
  \texttt{StimDrivenDetector} (per-anchor cls + offset).
\item \texttt{tests/test\_gomb\_soma\_vision\_stim\_anchor.py}
\item \texttt{tests/test\_gomb\_soma\_vision\_stim\_graph.py}
\item \texttt{tests/test\_gomb\_soma\_vision\_stim\_classifier.py}
\item \texttt{tests/test\_gomb\_soma\_vision\_stim\_detector.py}
\item \texttt{experiments/run\_gomb\_soma\_stim\_cluttered\_mnist\_2026\_05\_*.sh}
\end{itemize}

\section{CORE.YAML items touched}

None.

\section{Phased rollout (one per session, per CLAUDE.md \S6.5 \#11)}

\begin{enumerate}
\item \texttt{SalienceHead} + \texttt{TopKSelector} (Gumbel-softmax + hard top-K) + unit tests.
\item \texttt{StimulusGraphBuilder}: anchors $\to$ $k$-NN $\to$ signed hyperedges; unit tests.
\item \texttt{StimDrivenImageClassifier}: end-to-end on MNIST; unit tests; smoke train.
\item \texttt{StimDrivenDetector}: per-anchor head on Cluttered MNIST; smoke train.
\item Falsification battery against HyMeYOLO baselines + Phase 3-V baseline.
\end{enumerate}

Each phase produces a working module under unit-test coverage.
No phase combines.

\section{Test strategy}

\subsection{Unit tests per module}

For each new module:
\begin{itemize}
\item permutation-equivariance over anchor set;
\item Gumbel temperature $\to$ 0 reproduces hard top-K (in training mode);
\item gradient flow through salience head AND through top-K (the
  Gumbel trick lets gradient pass);
\item parameter count within budget;
\item end-to-end overfit-2-samples test.
\end{itemize}

\subsection{Falsification battery}

\begin{center}
\begin{tabular}{lll}
\toprule
Reference                                & Metric        & \Stim{} prediction \\
\midrule
Phase 3-V (rolling-shutter \Soma)        & MNIST acc.    & match or beat \\
HyMeYOLO \texttt{+ricci+kcycle}          & 0.204 mAP50   & \textbf{beat} \\
HyMeYOLO \texttt{+kcycle (broken)}       & 0.204 mAP50   & beat \\
HyMeYOLO \texttt{+ricci-mod} (positive control) & 0.723 mAP50 & approach \\
HyMeYOLO \texttt{boxes+circles}          & 0.715 mAP50   & approach \\
\bottomrule
\end{tabular}
\end{center}

If \Stim{} cracks $\geq 0.5$ mAP50 on Cluttered MNIST at comparable
parameter count, the stimulus-driven hypothesis is empirically
supported and we have a real architectural contribution to write up.

\section{Performance budget}

\begin{itemize}
\item Peak RSS: $\leq 14$\,GB (cgroups MemoryMax 16\,GB).
\item Per-seed wall on Cluttered MNIST (5000 imgs, 50 epochs): $\leq 25$\,min on the RTX 2070 SUPER (HyMeYOLO baseline was $\sim$13\,min).
\item GPU memory: $\leq 7$\,GB.
\item Falsification battery (5 seeds, 1 dataset): $\leq 2$\,h total.
\end{itemize}

\section{Risk anticipation}

\begin{itemize}
\item \emph{Top-K differentiability.} Gumbel-softmax over $K$
  selections may exhibit training instability. Fallback: straight-
  through estimator with hard top-K + soft gradient via the salience
  head; or sigmoid + threshold (loses cardinality control).

\item \emph{Anchor count too small.} K too small under-represents
  the scene. Budget $K \in \{8, 16, 32\}$, default 16; choose
  per-dataset empirically (no Optuna, just a few hand-picked values).

\item \emph{$\sig$-balance criterion may not bite.} If anchor-pair
  signs are random-ish on natural images, $\sig$-balance adds noise
  rather than signal. Mitigation: the learned threshold $\theta$
  in $\sig(u, v)$ should drive $\sig = +1$ for within-object pairs
  and $\sig = -1$ for cross-object pairs. Test by inspecting
  $\sig$-distribution on planted Cluttered MNIST samples.

\item \emph{HyMeYOLO baseline reproducibility.} The
  \texttt{+ricci-mod} 0.723 mAP50 came from a specific training
  protocol; \Stim{} should use the same protocol (5000 images,
  50 epochs, same data splits) for a fair comparison.

\item \emph{Performance-contract preservation.} The Phase 1-3-G
  layers already enforce \texttt{max\_cycles} / top-$k$ caps via the
  \texttt{HypergraphConvConfig}; \Stim{} must inherit. Verified by
  the \texttt{StimulusGraphBuilder} unit tests that cap polygon /
  walk counts before handing off.

\item \emph{Empty-plan-dir hygiene.} If \Stim{} is shelved, the
  plan dir is deleted before commit; no half-finished implementations
  left behind.
\end{itemize}

\section{Decision tree}

\begin{itemize}
\item \Stim{} $\geq 0.5$ mAP50 on Cluttered MNIST $\Rightarrow$
  stimulus-driven hypothesis confirmed; write up as a paper
  contribution: ``content-driven hypergraph topology for detection.''
\item \Stim{} $\geq$ HyMeYOLO \texttt{+kcycle} (0.204) but $< 0.5$
  $\Rightarrow$ partial confirmation; the architecture is doing
  something but not enough alone. Combine with Phase 4 (TriangleConv
  + Cartwright--Harary gate) and re-test.
\item \Stim{} $<$ HyMeYOLO \texttt{+kcycle} $\Rightarrow$
  stimulus-driven approach was wrong; document as negative result;
  return to Phase 3-V rolling-shutter as the canonical \Soma{}
  vision encoder.
\item \Stim{} matches HyMeYOLO \texttt{+ricci-mod} (0.723) $\Rightarrow$
  paper-grade contribution; YOLO-replacement plausible at this
  parameter scale.
\end{itemize}

\section{Why this could actually beat YOLO at its own game}

YOLO's anchor grid is the bottleneck of small-object / cluttered
detection. Every anchor predicts independently; the network has no
mechanism to say ``these three anchors together form a coherent
object.'' \Stim{} natively expresses that: a triangle of anchors
with $\pi(c) = +1$ is a balanced clique, i.e., a coherent
3-region object hypothesis. The architecture \emph{is} the
relational reasoner; the detection head reads off the consensus.

Cluttered MNIST is the canonical falsification target because the
clutter is literally circles that look like digits --- to beat the
clutter you need to reason \emph{about which combinations of
regions actually form a digit}, which is precisely what
$\sig$-balanced triangles encode.

\section{Bottom line}

\Stim{} replaces the rolling-shutter encoder of Phase 3-V with a
content-driven anchor selector + dynamic signed hypergraph builder.
The Phase 1-3-G layers (Walk, Polygon, Triangle) are unchanged;
\Stim{} is purely a graph-construction-stage swap. The
falsification battery is concrete (Cluttered MNIST against four
HyMeYOLO baselines plus the rolling-shutter \Soma{} reference).
A confirmed result is a paper-grade contribution: a YOLO-class
vision architecture built from signed hypergraph primitives, with
the architectural mechanism for relational object reasoning that
YOLO lacks.

\end{document}
