\section{Introduction}
\label{sec:introduction}

Kolmogorov--Arnold Networks (KAN) replace each fixed neural-network
activation with a learnable univariate spline, achieving
parameter-efficient function approximation on small-to-mid-scale
benchmarks~\cite{liu2024kan}. Recent extensions reach ordinary graphs
and unsigned hypergraphs, but no published variant ingests the
\emph{signed-incidence} hypergraph model that signed-graph
theory~\cite{cartwright1956structural} and applied work on signed
networks~\cite{leskovec2010signed,derr2018sgcn,huang2019sigat} have
established as the right primitive when edges carry direction-of-
information-flow semantics. The literature gap is structural: the KAN
formalism lifts cleanly to incidence-typed hypergraphs but no one has
done it.

We close that gap. \textbf{SignedKAN} is, to our knowledge, the first
KAN variant on signed hypergraphs. Its forward pass decomposes each
hyperedge's vertex aggregation into three sign-conditioned
sub-aggregations (one each for $\sigma \in \{+1, -1, \sim 0\}$), with
a separate inner+outer Cox--de Boor B-spline pair learned per
(vertex, edge, sign) triple. The inner spline shapes per-vertex
contributions before aggregation; the outer spline reshapes the
aggregated edge embedding before passing it on to the classifier.
This is \emph{Option C} of the design space we map in
\S\ref{sec:signedkan}; the choice is motivated by isolating signs at
the inner-spline level, where they have the most representational
leverage with the lowest parameter cost.

\subsection*{Contributions}

\textbf{(C1)} A KAN architecture for signed-incidence hypergraphs
with three-branch sign-conditioned aggregation
(\S\ref{sec:signedkan}). The architecture is fully differentiable,
end-to-end trainable, and has no hyperparameter beyond standard KAN
ones (spline order $k$, grid $G$).

\textbf{(C2)} A reproducible benchmark on Bitcoin Alpha and Bitcoin
OTC link-sign prediction~\cite{kumar2016bitcoin} with
sign-balance-triad hyperedge construction. SignedKAN reaches
AUC $\approx 0.80$ on Alpha and AUC $\approx 0.85$ on OTC at
$\sim 122$\,k parameters with no per-dataset tuning
(\S\ref{sec:experiments}).

\textbf{(C3)} An ablation that isolates the contribution of signed
semantics: the same architecture with all $\sigma$ collapsed to $+1$
underperforms the full SignedKAN on both metrics, quantifying the
value of carrying sign information through the spline activations
(\S\ref{sec:experiments}).

\subsection*{Scope}

This is a work-in-progress submission. The architecture and the
two-dataset characterization are settled; matching SGCN's published
AUC of $\sim 0.93$~\cite{derr2018sgcn} on Bitcoin Alpha is left to
the journal extension. The WiP-track contribution is the
\emph{architecture}, not a state-of-the-art table.
