\section{Background}
\label{sec:background}

\subsection{Kolmogorov--Arnold Networks}

The Kolmogorov--Arnold representation theorem~\cite{kolmogorov1957}
states that any continuous function $f : [0,1]^n \to \mathbb{R}$ can
be expressed as
$f(\mathbf{x}) = \sum_{q=1}^{2n+1} \Phi_q\!\left(\sum_{p=1}^{n} \phi_{p,q}(x_p)\right)$
for some univariate continuous $\Phi_q, \phi_{p,q}$.
Liu~et~al.~\cite{liu2024kan} cast this as a neural-network primitive:
each connection learns a univariate spline activation rather than a
fixed nonlinearity on top of a learnable weight. For Cox--de
Boor~\cite{coxdeboor} B-spline parameterisation with grid $G$ and
order $k$, each activation has $G + k - 1$ learnable coefficients.
The architecture is parameter-efficient on small-to-mid-scale
function-regression benchmarks; recent extensions reach ordinary
graphs.

\subsection{Signed graphs and signed networks}

A signed graph carries an edge-label function
$\sigma : E \to \{+1, -1\}$. Cartwright and
Harary~\cite{cartwright1956structural} introduced \emph{structural
balance}: a triangle $(u, v, w)$ is balanced iff
$\sigma(uv)\sigma(vw)\sigma(uw) = +1$.
Leskovec~et~al.~\cite{leskovec2010signed} showed that real signed
networks (Wikipedia voting, Slashdot, Epinions) are predominantly
balanced. Signed-graph neural networks
(SGCN~\cite{derr2018sgcn}, SiGAT~\cite{huang2019sigat},
SDGNN~\cite{huang2021sdgnn}) propagate sign-aware messages, but use
fixed (typically ReLU or LeakyReLU) activations.

\subsection{Hypergraph neural networks}

Hypergraph neural networks~\cite{feng2019hgnn,bai2021hgnn} extend
GNNs to hyperedges through clique-expansion or star-expansion
projections. \emph{Signed} hypergraph extensions are less developed;
Huang and collaborators have proposed sign-aware hypergraph variants
in the SDGNN family but the activations remain fixed. To our
knowledge, no published architecture combines:
(i) signed-incidence hyperedges,
(ii) per-vertex / per-edge / per-sign learnable activations,
(iii) the Kolmogorov--Arnold formalism.
SignedKAN fills this gap.

\subsection{Distinguishing from concurrent work}

Recent unsigned hypergraph KAN variants (e.g.\ Fang et al.\
``HyperKAN'') treat all incidences identically; they cannot ingest
sign information without a separate preprocessing step that projects
signs out at the input. SignedKAN keeps signs first-class throughout
the architecture --- inner spline, outer spline, and sub-aggregation
all branch on $\sigma$ --- which is what the ablation in
\S\ref{sec:experiments} measures the contribution of.
