\documentclass[11pt]{article}
\usepackage[a4paper,margin=2cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\lstset{basicstyle=\ttfamily\small,breaklines=true,frame=single}

\title{Plan: \texttt{cycles.rs} Strategy Refactor\\
  Collapsing 16 Cartesian PyO3 Wrappers into 2 + \texttt{EnumerationConfig}}
\author{}
\date{2026-05-11}

\begin{document}
\maketitle

\section{Scope}

\texttt{hymeko\_py/src/cycles.rs} currently exposes \textbf{16
distinct} \texttt{\#[pyfunction]} enumerator variants whose names
encode a Cartesian product over axes \{enumeration mode, scorer,
pruner, vertex-filter, ABB strategy, batched output, tiered caps\}.
Average wrapper is 134 LOC, mostly copy-paste boilerplate that
re-implements scorer dispatch, pruner dispatch, filter construction,
SoA conversion, and PyArray output.  Total 2251 LOC, 0 structs, 0
traits.

The underlying Rust algorithm crate \texttt{hymeko\_graph} \emph{already}
provides clean Strategy traits and impls
(\texttt{Scorer}, \texttt{Pruner}, \texttt{VertexFilter}, with 4--5
concrete implementations each).  This refactor preserves those traits
and replaces the 16 wrappers with \textbf{two} family entry points
plus one configuration struct.  No algorithm change.

\section{Goal}

\begin{itemize}
  \item One \texttt{\#[pyfunction]} per \emph{family}
    (cycles, walks) — not per Cartesian cell.
  \item All toggles enter through \texttt{EnumerationConfig}
    (\texttt{score\_kind}, \texttt{pruner\_kind}, \texttt{filter\_kind},
    \texttt{abb\_mode}, \texttt{batched}, \texttt{tiered}, \dots).
  \item Internal dispatcher matches \texttt{(mode, abb, batched, tiered)}
    and delegates to the existing parameterized algorithm functions
    in \texttt{hymeko\_graph::topk\_cycles}.
\end{itemize}

\section{Affected files}

\begin{itemize}
  \item \texttt{hymeko\_py/src/cycles.rs} — full rewrite of the
    PyO3 surface.  Internal helpers preserved.
    Expected: 2251 → \textasciitilde 800 LOC.
  \item \texttt{hymeko\_py/src/lib.rs} — \texttt{m.add\_function}
    registrations updated to point at new entry-point names.
  \item \textbf{24 Python call sites} (counted via grep) across
    \texttt{signedkan\_wip/} that currently use the legacy names —
    migrated to the new \texttt{enumerate\_cycles\_rs(\dots, config)}
    signature.  The most touched callers in recent work:
    \texttt{signedkan\_wip/src/run\_gomb\_smoke.py},
    \texttt{signedkan\_wip/src/cycle\_cache.py},
    \texttt{signedkan\_wip/src/n\_tuples.py}.
\end{itemize}

\textbf{CORE.YAML items touched: none.}  \texttt{hymeko\_py} is not
listed in \texttt{Core.yaml} (which protects only the not-yet-existing
\texttt{crates/hymeko\_*} paths).  Refactor permitted without
escalation per CLAUDE.md \S1.

\section{Public API}

\begin{lstlisting}[language=Python]
# New entry points (Python-facing)
enumerate_cycles_rs(
    edges_u, edges_v, edges_s, n_nodes,
    k_len, m_per_vertex,   # m_per_vertex=0 means "all-mode"
    score_kind="fraction_negative",
    pruner_kind="none",
    filter_kind="none",
    filter_min_degree=2,
    abb_mode="none",           # "none" | "start_local" | "global_min"
    batched=True,
    tiered=False,
    fullness_gate=0.25,
    adaptive_m_v=None,
) -> tuple[ndarray, ndarray]   # (cycles (M, k), scores (M,))

enumerate_walks_rs(
    edges_u, edges_v, edges_s, n_nodes,
    walk_len, m_per_vertex,
    score_kind="fraction_negative",
    ...
) -> tuple[ndarray, ndarray]
\end{lstlisting}

The 16 legacy names (\texttt{enumerate\_top\_k\_per\_vertex\_cycles\_signed\_filtered\_batched\_rs},
\dots) are \textbf{removed}.  Callers migrate to the new signature in
the same commit — no shim period (per memory
\texttt{feedback\_decompose\_long\_files}: clean breaks beat deprecation cruft).

\section{Internal structure (Rust)}

\begin{lstlisting}[language=Rust]
#[derive(Debug, Clone)]
pub struct EnumerationConfig<'a> {
    pub k_len: usize,
    pub m_per_vertex: u32,             // 0 = all-mode
    pub score_kind: &'a str,
    pub pruner_kind: &'a str,
    pub filter_kind: &'a str,
    pub filter_min_degree: u32,
    pub abb_mode: AbbMode,             // None | StartLocal | GlobalMin
    pub batched: bool,
    pub tiered: bool,
    pub fullness_gate: f64,
}

fn dispatch_cycle_enum(
    g: &SignedGraph, cfg: &EnumerationConfig,
) -> Result<Batch, EnumError> {
    let scorer = pick_scorer(cfg.score_kind)?;
    let pruner = pick_pruner(cfg.pruner_kind)?;
    let filter = build_vertex_filter(cfg.filter_kind, cfg.filter_min_degree)?;
    let keep = filter.keep_set(g);
    match (cfg.m_per_vertex, cfg.abb_mode, cfg.batched, cfg.tiered) {
        (0, _, _, _)              => /* all-mode */,
        (m, AbbMode::None, true,  false) => /* per-vertex batched */,
        (m, AbbMode::StartLocal, true, false) => /* per-vertex BB batched */,
        (m, AbbMode::GlobalMin,  true, false) => /* per-vertex BB-global batched */,
        (m, _, _, true)            => /* tiered cap variant */,
        ...
    }
}

#[pyfunction]
pub fn enumerate_cycles_rs(...) -> PyResult<(...)> {
    let g = build_signed_graph(...)?;
    let cfg = EnumerationConfig { ... };
    let batch = py.detach(|| dispatch_cycle_enum(&g, &cfg));
    batch_to_pyarray(py, batch?, cfg.k_len)
}
\end{lstlisting}

\section{Test strategy}

\begin{itemize}
  \item \textbf{Rust:} \texttt{cargo test -p hymeko\_py} continues to
    pass.  Add 1 dispatch-table test: every (mode, abb, batched,
    tiered, scorer, pruner) combination produces a non-empty result on
    a small fixture or returns a typed error.
  \item \textbf{Python:} all existing \texttt{signedkan\_wip/tests/}
    pass after migration (currently the suite passes; baseline is the
    13 + 15 + \dots gomb/etc.\ tests we ran today plus the cycle\_cache
    tests at \texttt{signedkan\_wip/tests/test\_cycle\_cache.py}).
  \item \textbf{Regression:} compare cycle output (sorted) between
    the old and new wrappers for at least one (scorer × pruner ×
    filter) cell each on Bitcoin OTC, before deleting the legacy
    pyfunctions.  Output equality (modulo ordering) is required.
\end{itemize}

\section{Performance budget}

No algorithm change.  Expected wall-time delta: \(\pm 5\%\) noise (single
extra match-on-strings per call; negligible vs.\ DFS enumeration).  RSS
unchanged.  If we measure $> 10\%$ regression, CLAUDE.md \S 3 forces
investigation per the symmetric rule.

\section{Risk anticipation}

\begin{itemize}
  \item \textbf{Caller-side breakage}: 24 Python call sites need
    migration.  Mitigation: refactor in one commit; smoke-run
    \texttt{run\_gomb\_smoke.py} on Bitcoin OTC after migration to
    catch typos.
  \item \textbf{Cycle output ordering}: the per-vertex enumerator
    returns cycles in some implementation-defined order.  Sorted-set
    equality required for the regression test, not list equality.
  \item \textbf{Cache fingerprint drift} (memory
    \texttt{feedback\_cycle\_cache\_fingerprint}): \texttt{cycle\_cache.py}'s
    \texttt{\_topk\_fingerprint} dict must include every config field
    that affects the output.  Refactor must preserve fingerprint coverage
    — i.e.\ add \texttt{abb\_mode}, \texttt{tiered}, \texttt{batched} to the
    fingerprint if they weren't already (likely missing — that was
    the canonical failure mode in the prior memory).
  \item \textbf{ABB lossy variants}: the \texttt{\_bb\_} variants
    produce a SUBSET of the non-BB output (UB-pruning by construction).
    Regression test must use the same \texttt{abb\_mode} on both sides,
    not compare across modes.
\end{itemize}

\section{Rollback}

\texttt{git revert <commit>}.  Single commit per CLAUDE.md \S 0
checkpoint discipline.

\section{Sequencing}

\begin{enumerate}
  \item Write the plan (\textbf{this doc}).
  \item Introduce \texttt{EnumerationConfig} + \texttt{AbbMode} +
    \texttt{dispatch\_cycle\_enum} alongside the existing wrappers
    (zero-impact additive change).
  \item Add the 2 new \texttt{\#[pyfunction]}s (\texttt{enumerate\_cycles\_rs},
    \texttt{enumerate\_walks\_rs}) calling the new dispatcher.
  \item Regression test: confirm new wrappers produce identical
    (sorted) output to the legacy ones on Bitcoin OTC for at least
    one cell per axis.
  \item Migrate all 24 Python call sites to the new signature.
  \item Run full test suite (Rust + Python).
  \item Delete the 16 legacy \texttt{\#[pyfunction]}s and their
    \texttt{lib.rs} registrations.
  \item Re-run \texttt{cargo clippy --all-targets -- -D warnings} and
    \texttt{pytest -p no:randomly} as gates per CLAUDE.md \S 6.3.
  \item Write the report (\texttt{reports/2026-05-11-cycles-strategy-refactor.md}).
\end{enumerate}

\section{Acceptance gates}

\begin{itemize}
  \item All Rust tests pass.
  \item All Python tests pass.
  \item \texttt{cargo clippy} clean under \texttt{-D warnings}.
  \item \texttt{wc -l hymeko\_py/src/cycles.rs} $<$ 1000 (target 700--900).
  \item Smoke run: Bitcoin OTC gomb 5-seed (existing
    \texttt{reports/2026-05-11-hymeko-gomb-sphere.md} numbers) reproduces
    within $\pm 0.001$ AUC.
  \item Old pyfunction names are gone (\texttt{grep -c "enumerate\_top\_k"
    hymeko\_py/src/cycles.rs} returns 0 or 1, not 16).
\end{itemize}

\end{document}
