\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.4cm]{geometry}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}

% Convenience macros.
\newcommand{\KA}{\textsc{ka}}
\newcommand{\KAN}{\textsc{kan}}
\newcommand{\HSiKAN}{\textsc{HSiKAN}}
\newcommand{\HyMeKo}{\textsc{HyMeKo}}
\newcommand{\code}[1]{\texttt{\small #1}}
\setlist[itemize]{leftmargin=1.4em,itemsep=2pt,topsep=3pt}

\title{\HSiKAN: Mathematical Foundation\\
\large {\normalfont Eight layers, eight decades}}
\author{}
\date{2026-05-04}

\begin{document}
\maketitle

\noindent
\HSiKAN{} (Hypergraph Signed Kolmogorov--Arnold Network) is the
composition of eight mathematical layers, each pulled from a
different established literature spanning 1946 to 2024.  This
document maps the layers in dependency order, naming the source
results and the role each plays inside the model.

\section*{1.\;Signed-graph theory}
\textit{Heider 1946 \(\to\) Cartwright \& Harary 1956 \(\to\) Davis 1967}

\begin{itemize}
\item Signed graph $G = (V, E, s)$ with $s : E \to \{+1, -1\}$.
\item \textbf{Cycle balance theorem.}\;A cycle $c$ is \emph{balanced}
  iff
  \[
    \prod_{e \in c} s(e) \;=\; +1.
  \]
\item \textbf{Structural balance (Cartwright--Harary 1956).}\;A
  complete signed graph is balanced iff $V$ partitions into two
  mutually-hostile cliques.  Davis 1967 generalises to
  $k$-clusterability (weak balance: no all-negative triangle).
\item \textbf{\HSiKAN's $\sigma$-rule.}\;The per-vertex sign within
  a cycle is the cycle-product specialised to vertex $i$:
  \[
    \sigma_i(c) \;=\; (-1)^{n_i^{-}(c)},
  \]
  where $n_i^{-}(c)$ is the count of negative edges incident to
  vertex $i$ within cycle $c$.
\end{itemize}

\section*{2.\;Hypergraph theory}
\textit{Berge 1973; Thesis~2 (this dissertation)}

\begin{itemize}
\item A $k$-uniform hypergraph generalises edges to $k$-vertex
  relations.
\item Two canonical reductions to ordinary graphs:
  \begin{itemize}
  \item \textbf{Star expansion.}\;Hyperedge $e \mapsto$ centroid
    vertex $c_e$ plus $k$ bipartite edges
    $\{(v_i, c_e) : v_i \in e\}$.  The result is bipartite
    $\Rightarrow$ only even-$k$ cycles exist in the expansion.
    \emph{This is \HSiKAN's input format.}
  \item \textbf{Clique expansion.}\;$k$-edge $\mapsto$
    $\binom{k}{2}$ pairwise edges.  Discards arity information.
  \end{itemize}
\item Sparse signed-incidence buffers enter at training:
  \[
    \mathbf{M}_e^{(k)} \in \mathbb{R}^{|E_{\text{test}}| \times |T_k|}
    \qquad\text{and}\qquad
    \mathbf{M}_{vt} \in \mathbb{R}^{|V| \times |T|}.
  \]
\end{itemize}

\section*{3.\;Kolmogorov--Arnold representation}
\textit{Kolmogorov 1957 \(\to\) Liu et al.\ 2024 (\KAN)}

\begin{itemize}
\item \textbf{KA superposition theorem.}\;Any continuous
  $f : [0, 1]^n \to \mathbb{R}$ admits the decomposition
  \[
    f(x) \;=\; \sum_{q=0}^{2n} \Phi_q\!\left(
      \sum_{p=1}^{n} \psi_{q,p}(x_p)
    \right)
  \]
  with $2n{+}1$ outer and $n(2n{+}1)$ inner univariate functions.
\item \textbf{\KAN s} parameterise $\Phi_q$ and $\psi_{q,p}$ as
  splines.
\item \textbf{KA-rank ladder for signed graphs.}\;Three points on
  the spectrum:
  \begin{itemize}
  \item rank-1: HGNN (linear superposition);
  \item rank-2: Signed-HGNN;
  \item full rank: \HSiKAN{} (per-sign nonlinear inner+outer
    splines).
  \end{itemize}
\end{itemize}

\section*{4.\;Spline approximation}
\textit{Catmull--Rom 1974, de Boor 1978, Kochanek--Bartels 1984}

\begin{itemize}
\item Inner $\varphi_v^\sigma$ and outer $\varphi_e^\sigma$ are
  uniform Catmull--Rom cubic splines on $[-1, 1]$:
  \begin{itemize}
  \item 4-point gather $(P_{i-1}, P_i, P_{i+1}, P_{i+2})$ per
    output;
  \item constant $4 \times 4$ basis matrix $M_{CR}$ folds the
    Hermite blend coefficients;
  \item $C^1$-continuous, interpolatory --- the curve passes
    through the control points.
  \end{itemize}
\item Backbone-invariant: B-splines (Cox--de Boor recursion) and
  Kochanek--Bartels (TCB tension / continuity / bias parameters,
  reducing to Catmull--Rom at $\mathrm{TCB} = (0, 0, 0)$) yield
  the same downstream AUCs to within seed noise.
\end{itemize}

\section*{5.\;Mixed-arity mixing}

The per-arity cycle embeddings are combined by a softmax-normalised
learnable weight vector $\boldsymbol\alpha$:
\[
  \mathbf{H}^{\text{edge}}
  \;=\; \sum_{k \in \mathcal{K}} \operatorname{softmax}(\boldsymbol\alpha)_k
        \cdot \mathbf{M}_e^{(k)}\, \mathbf{H}^{(e, k)}.
\]
$\alpha_k$ acts as a \textbf{quantitative ethogram-like read-out}:
which cycle order does this dataset's signed signal live at?
Empirically: Bitcoin $\to k{=}3$ triads, Slashdot $\to k{=}5$
($46\%$), SBM $\to k{=}4$ matching its 4-community design.

\section*{6.\;Symbolic distillation (Fourier-style)}

After training, each surviving spline is fit against a six-form
symbolic library
\[
  \{
    \text{linear},\;
    \text{quadratic},\;
    \text{cubic},\;
    a\sin(\omega x + \varphi) + c,\;
    \exp,\;
    \tanh
  \}
\]
via Levenberg--Marquardt least squares.  Approximately $91\%$ of
survivors fit best as a sinusoid.  Validated on Bitcoin Alpha + OTC
($3$ seeds, $h{=}16$) against three null baselines:

\begin{center}
\begin{tabular}{lr}
\hline
baseline & sinusoidal fraction \\
\hline
trained \HSiKAN          & $0.901 \pm 0.055$ \\
untrained \HSiKAN        & $0.505 \pm 0.036$ \\
random spline coefs      & $0.583 \pm 0.045$ \\
Gaussian-process draws   & $0.499 \pm 0.021$ \\
\hline
\end{tabular}
\end{center}

\noindent
The $+35$--$40$\,pp gap rules out spline-basis bias and
grid-sampling artefacts.

\section*{7.\;Cycle enumeration (algorithmic)}

\begin{itemize}
\item DFS with $\texttt{Vec<u64>}$ bitset visited (8$\times$ smaller
  than $\texttt{Vec<bool>}$, fits in L1 for graphs up to $\sim
  500\text{k}$ nodes).
\item BFS-distance pruning to the start vertex (skip branches that
  cannot close in the remaining budget).
\item \textbf{Color-coding (Alon--Yuster--Zwick 1995).}\;
  Random $k$-colouring of vertices; only rainbow cycles are
  emitted; per-colouring cost is proportional to
  $|\{\text{rainbow cycles}\}| / |\{\text{total cycles}\}|$.
  Probability a fixed $k$-cycle is rainbow $= k!/k^k$.
\item Vitter reservoir sampling for unbiased cycle subsets.
\item Canonical form: smallest-vertex rotation $+$ lex orientation
  tiebreak.
\item \textbf{Walk-\HSiKAN{} variant.}\;Same DFS, no closure check;
  open length-$L$ simple walks emitted with
  $\mathrm{path}[0] \le \mathrm{path}[L]$ canonicalisation.
\item \textbf{Performance.}\;Slashdot $k{=}4$ in $4.6$\,s for
  $100$\,k cycles via Rust + rayon + atomic early-stop, vs
  $\sim 227$\,s for the equivalent serial Python.
\end{itemize}

\section*{8.\;Information geometry and regularisation}

\begin{itemize}
\item \textbf{Spectral entropy} as a KA-bound:
  \[
    H(W) \;=\; -\sum_i \tilde\lambda_i \log \tilde\lambda_i \big/
                 \log \operatorname{rank}(W).
  \]
\item \textbf{Sanchez--Giraldo matrix-based R\'enyi-$\alpha$
  entropy} estimator (kernel $\alpha$-norm).
\item \textbf{Multi-term Lyapunov wrapping}:
  \[
    \lambda_{\text{eff}}
    \;=\; \lambda_0 \exp\!\big(
      -\eta\, D_{\mathrm{KL}}(\hat\lambda^{(t-1)} \,\|\, \hat\lambda^{(t)})
    \big).
  \]
\item \textbf{Path~I calibration law:}
  $\lambda_{\text{multi}} \sim \lambda_{\text{scalar}} / L$ ---
  depth-dependent rescaling.  Empirically validated by ph18c:
  highway-10 confirmed at $\lambda{=}1.0$ ($+0.092$\,pp,
  $p<0.05$); resmlp-40 ($L{=}40$) recovers at $\lambda{=}0.5$
  ($+0.213$\,pp, $p<0.05$), the predicted ratio
  $\approx 1/4$ of the $L{=}10$ value.
\end{itemize}

\section*{What today's chicken-aggression work added}

\begin{itemize}
\item \textbf{Layer~1 reproduces itself in a flock simulator.}\;
  The Cartwright--Harary structural-balance theorem implies
  aggressors generate \emph{balanced} triads: the
  $(\mathrm{aggressor}, v_1, v_2)$ pattern has signs $(-, -, +)$
  because two victims of the same aggressor are typically
  peaceful with each other.  Sign product $= +1$, so the cycle
  is balanced.  Our unsupervised cycle-balance scorer reads off
  per-vertex aggressor probability directly from this fraction
  --- \emph{graph theory from 1956 identifying which chicken is
  the aggressor with zero labels}.
\item \textbf{Layer~2 generalises to anatomy.}\;The chicken body
  itself is a hypergraph: 12 keypoints, rigid bones (sign $+1$),
  flexible joints (sign $-1$), kinematic chains as multi-vertex
  hyperedges.  Star-expandable through the same machinery.
  Per-bird anatomy nests inside per-flock interaction --- a
  multi-scale signed-hypergraph description of an entire flock.
\end{itemize}

\section*{The aesthetic point}

The stack composes from
\textbf{1946 (Heider)} $\to$
\textbf{1956 (Cartwright--Harary)} $\to$
\textbf{1957 (Kolmogorov)} $\to$
\textbf{1973 (Berge)} $\to$
\textbf{1974 (Catmull--Rom)} $\to$
\textbf{1995 (Alon--Yuster--Zwick)} $\to$
\textbf{2024 (Liu \KAN)} $\to$
\textbf{2026 (\HSiKAN)}.

\medskip
\noindent
Eighty years of pure mathematics, quietly waiting to read off
which chicken is the aggressor.

\end{document}
