\documentclass[10pt,a4paper]{article}

\usepackage[margin=1.8cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{parskip}
\usepackage{lmodern}
\usepackage{enumitem}

\hypersetup{colorlinks=true, linkcolor=black, urlcolor=blue}

\title{Phases and Paths of the\\
       Spectral-Entropy Universality Programme}
\author{thesis-IV working notes (HyMeKo)}
\date{2026-04-26}

\newcommand{\code}[1]{\texttt{\detokenize{#1}}}
\newcommand{\arm}[1]{\texttt{\detokenize{#1}}}
\newcommand{\pp}{\,pp}

\begin{document}
\maketitle

\begin{abstract}
This document is a single-source reference for the spectral-entropy
regulariser programme. It defines every \emph{phase} (the experimental
campaigns run in chronological order) and every \emph{path} (the
regulariser variants tested), gives the configuration of each, and
records the empirical headline of each. It is intended for paper
preparation and future re-runs --- not for results interpretation
(see \code{RESULTS_VIEWS_SUITE.md} and the per-phase reports for that).
\end{abstract}

\section{Overview}

The programme tests whether a spectral-entropy regulariser inspired by
Hajdu thesis~IV can be made \emph{universal} across architectures and
datasets. Eight phases were run between 2026-04-22 and 2026-04-26.
Twelve regulariser \emph{arms} (paths) were defined and benchmarked
against a paired \arm{baseline} (no regulariser). All comparisons are
paired by seed to share weight initialisation and DataLoader shuffle.

\paragraph{Total experiments to date.} 98+ paired runs $\times$
$\sim$2--300 seeds = roughly 11\,000 individual training runs.

\paragraph{Four anchors.}
\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
  \item \textbf{MNIST plain MLP}: original $\Delta = +0.149$\pp{} positive (the seed of the programme).
  \item \textbf{Spirals 2-D synthetic}: largest positive ($+0.412 \to +0.724$\pp{} across paths).
  \item \textbf{Circles 2-D synthetic}: original $-0.060$\pp{} significant negative (stress test).
  \item \textbf{CapsMLP MNIST}: original $-0.057$\pp{} significant negative (stress test).
\end{itemize}

\section{Datasets and architectures}

\subsection{Datasets}

\begin{center}
\small
\begin{tabular}{@{}llrrl@{}}
\toprule
Identifier & Origin & Train size & Classes & Notes \\
\midrule
\code{mnist}, \code{mnist_small}     & MNIST         & 60k    & 10  & 28$\times$28 grayscale \\
\code{fashion_mnist}                  & FashionMNIST  & 60k    & 10  & 28$\times$28 grayscale (clothing) \\
\code{kmnist}                         & KMNIST        & 60k    & 10  & 28$\times$28 grayscale (Japanese hiragana). \\
                                      &               &        &     & HF \code{tanganke/kmnist} mirror; \\
                                      &               &        &     & official torchvision URL is dead. \\
\code{emnist_letters}                 & EMNIST        & 145k   & 26  & 28$\times$28 grayscale (letters) \\
\code{cifar10}                        & CIFAR-10      & 50k    & 10  & 32$\times$32 colour \\
\code{svhn}                           & SVHN          & 73k    & 10  & 32$\times$32 colour \\
\code{iris}, \code{wine},             & sklearn       & 100--500 & 2--10 & tabular \\
\code{breast_cancer}, \code{digits}   &               &        &     & \\
\code{two_moons}, \code{spirals},     & synthetic     & 2000   & 2--3 & 2-D classification \\
\code{circles}, \code{gaussian_quants}&               &        &     & \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Architectures}

\begin{center}
\small
\begin{tabular}{@{}llp{8.5cm}@{}}
\toprule
Identifier & Class & Description \\
\midrule
\code{MNISTNet}              & plain MLP             & 784$\to$256$\to$128$\to$64$\to$10 \\
\code{MNISTNetSmall}         & plain MLP, thesis     & 784$\to$16$\to$8$\to$10 (the anchor) \\
\code{MNISTNetSmallNClass}   & plain MLP, parametric & 784$\to$16$\to$8$\to$\textit{n} (used for EMNIST) \\
\code{ResMLP}                & residual MLP          & projection + N residual blocks at hidden 16 \\
\code{HighwayMLP}            & gated MLP             & projection + N highway blocks at hidden 16 \\
\code{CIFAR10Net}            & MLP                   & 3072$\to$64$\to$32$\to$10 (proj. skipped from regulariser) \\
\code{CIFAR10ResMLP}         & residual MLP          & projection 3072$\to$16, then N residual blocks \\
\code{SVHNNet}               & MLP                   & 3072$\to$64$\to$32$\to$10 (similar to CIFAR10Net) \\
\code{TabularMLP}            & plain MLP, parametric & input$\to$32$\to$16$\to$\textit{n} (sklearn datasets) \\
\code{SyntheticMLP}          & plain MLP, 2-D        & 2$\to$32$\to$16$\to$2 (two\_moons, spirals, circles) \\
\code{SyntheticMLP3}         & plain MLP, 2-D        & 2$\to$32$\to$16$\to$3 (gaussian\_quants) \\
\code{CapsMLP}               & capsule network       & flat 784 $\to$ 32 primary caps of 8-D $\to$ 10 digit caps of 16-D, \\
                             &                       & dynamic routing (3 iters), squash activation \\
\bottomrule
\end{tabular}
\end{center}

\section{Regulariser arms (paths)}

All arms share the optimisation outer loop: Adam, $\lambda$
weight on the regulariser, every \code{reg-every-n} batches the
regulariser is recomputed from \code{model.spectral_adjacency(view)} and
its (data-dependent or weight-only) statistics.

\subsection{Originals (phases 1--5)}

\begin{description}[leftmargin=1.4em, style=nextline, itemsep=2pt]
  \item[\arm{baseline}] No regulariser; \code{reg_term = 0}. Always
    paired against every treated arm.
  \item[\arm{scalar_entropy}] $\mathcal{L}_{\text{reg}} = \lambda \cdot
    H(A)$, where $H$ is Shannon entropy of normalised-Laplacian
    eigenvalues of the chosen-view adjacency. $\lambda{=}0.1$ default.
  \item[\arm{kl_trajectory}] $\mathcal{L}_{\text{reg}} = \lambda \cdot
    D_{KL}(\hat\lambda^{(t-1)} \,\|\, \hat\lambda^{(t)})$. Penalises
    step-to-step divergence of the spectrum.
  \item[\arm{l2_weight_decay}] Standard L$^2$ control arm:
    $\mathcal{L}_{\text{reg}} = \lambda \sum_l \|W_l\|_F^2$.
\end{description}

\subsection{Universality candidates (phases 6--7b)}

\begin{description}[leftmargin=1.4em, style=nextline, itemsep=2pt]
  \item[\arm{scalar_entropy_normalized} \textit{(Path B)}]
    $\mathcal{L}_{\text{reg}} = \lambda \cdot H(A) / \log_2
    \mathrm{rank}(A)$. \textbf{Scale-invariant}; ties to VC /
    Rademacher effective-rank bounds.
  \item[\arm{entropy_target} \textit{(Path A)}]
    $\mathcal{L}_{\text{reg}} = \lambda \cdot (H_{\text{norm}} - H^*)^2$
    with $H^* \in [0, 1]$. Squared-deviation around an
    information-bottleneck-style compression target. $H^* = 0.5$ is the
    universal-default sweet spot.
  \item[\arm{structural_composite} \textit{(Path C)}]
    $\mathcal{L}_{\text{reg}} = \lambda \cdot \max( H_{\text{norm}},\,
    \sigma_{\max}/2,\, 1 - \mathrm{sr}/\mathrm{rank} )$.
    Penalises whichever of three Bartlett-Foster-Telgarsky bound terms
    is largest at the moment.
  \item[\arm{entropy_unified}] $\mathcal{L}_{\text{reg}} = \lambda \cdot
    ((H_{\text{norm}} - H^*)^2 + \beta H_{\text{norm}})$. Convex blend
    of A and B; $\beta{=}1.0$ default makes it numerically equivalent
    to $H^*{=}0.0$ entropy\_target.
  \item[\arm{entropy_adaptive}] $\mathcal{L}_{\text{reg}} = \lambda(t)
    \cdot H_{\text{norm}}$, with $\lambda(t) = \lambda \cdot
    \exp(\eta \cdot \ddot{H}_{\text{norm}})$ clamped to $[0.1, 10]$.
    Multiplicative modulator on $\lambda$ driven by the
    \emph{second derivative} of $H_{\text{norm}}$ across reg-update
    steps.
  \item[\arm{total_combined}] $\mathcal{L}_{\text{reg}} = \lambda \cdot
    (\text{A} + \text{B} + \text{C})$. Literal sum of the three;
    triple-counts $H_{\text{norm}}$. Robustness-oriented.
  \item[\arm{entropy_target_ka}] $\mathcal{L}_{\text{reg}} = \lambda
    \cdot (H_{\text{norm}} - H^*_{KA})^2$, with $H^*_{KA} = \max(0.5,
    \min(1.0, \log_2(2n+1)/\log_2(\mathrm{rank})))$.
    \textbf{Theory-driven default:} the Hecht-Nielsen / Kolmogorov-Arnold
    representation theorem gives a per-dataset lower bound on the
    minimum representational rank ($2n+1$ for $n$-dim input).
  \item[\arm{entropy_telgarsky}] Per-layer entropy with depth-decay
    weighting: $\mathcal{L}_{\text{reg}} = \lambda \sum_l (1 - l/L)^p
    H_{\text{norm}}^{(l)}$. Operationalises Telgarsky's depth-vs-width
    tradeoff: deeper layers carry less penalty (preserve information
    late, compress early).
\end{description}

\subsection{Activation-side family (phases 8 \& 11)}

\begin{description}[leftmargin=1.4em, style=nextline, itemsep=2pt]
  \item[\arm{cross_layer_mi} \textit{(Path F)}] $\mathcal{L}_{\text{reg}} = \lambda \cdot
    \frac{1}{\binom{L}{2}} \sum_{i<j} I_2(\hat K_i ; \hat K_j)$, where
    $\hat K_l$ is the trace-1 Gaussian-RBF Gram matrix of layer~$l$'s
    post-linear activations on the current mini-batch and $I_2$ is the
    Sanchez-Giraldo R\'enyi-2 mutual information via the Hadamard-product
    joint. Penalises layer redundancy; differentiable through
    activations to weights via forward hooks. Math:
    \texttt{reports/sanchez\_giraldo\_framework.\{md,pdf\}}.
\end{description}

\subsection{Multi-information + closed-loop control (phases 12--18, Path I)}

\begin{description}[leftmargin=1.4em, style=nextline, itemsep=2pt]
  \item[\arm{total_correlation_mi} \textit{(Path I)}]
    Strict L-way generalisation of Path F:
    $\text{TC}_2 = \sum_{l=1}^{L} H_2(\hat K_l) - H_2(\hat K_{\text{join}})$
    where $\hat K_{\text{join}} = \bigodot_l \hat K_l / \mathrm{tr}(\bigodot_l \hat K_l)$
    is the stepwise-renormalised Hadamard joint Gram. Wrapped with two
    closed-loop modulators:
    (i) Lyapunov-style $\lambda$ schedule
    $\lambda_{\text{eff}} = \lambda_0\,\exp(-\eta\,\mathrm{KL}_{\text{step}})$
    where $\mathrm{KL}_{\text{step}}$ is the eigenvalue-distribution KL between
    consecutive joint-Gram spectra;
    (ii) variance-momentum modulator $\mu_t = \beta\mu_{t-1} + (1-\beta)\sigma^2(\hat p_t)$
    with three CLI-selectable mixing modes (\code{damp}, \code{amplify}, \code{mix}).
    Math + design discussion: \code{reports/ph12_path_i_negative_result.md}.

  \item[\arm{entropy_lyapunov} \textit{(multi-term wrapping arm)}]
    Wraps Path A + Path B with the same Lyapunov KL-feedback schedule,
    parameterised by a per-term weight pair $(\lambda_a, \lambda_b)$:
    $\mathcal{L}_{\text{reg}} = \lambda_{\text{eff}}\bigl(\lambda_a (H_{\text{norm}}-H^\star)^2 + \lambda_b H_{\text{norm}}\bigr)$.
    The KL signal in the schedule connects two previously independent
    regularisers (\arm{kl_trajectory} and \arm{entropy_target}) into a
    single closed-loop controller derived from $\dot V<0$ on the task loss.
\end{description}

\subsection{Generated nets via HyMeKo descriptions (phases 18, 18b)}

The Tier-2 emission path of the \code{torch_dataflow} backend
(see \code{transforms/torch_dataflow/template.py}) is exercised
end-to-end as a regularisation-bench fixture: \code{.hymeko} sources
in \code{data/nn/} for ResMLP-\{3,10,20,40\} and HighwayMLP-\{3,10,20\}
compile to PyTorch modules backed by \code{ehk_torch_stub} blocks,
register as \code{\_hymeko}-suffixed datasets in \code{run_benchmark.py},
and run the same regulariser arms as the hand-written PyTorch
counterparts. Confirms that the canonical IR round-trips through to
runnable nets without distorting regulariser dynamics
(architectural-parity claim).

\section{Phases}

Each phase is one overnight (or multi-overnight) campaign with its own
shell script and log file. Phases share the same harness
(\code{run_benchmark.py}) and CSV output schema (\code{data/benchmarks/thesis_iv_hard_*.csv}).

\subsection*{Phase 1 --- Original overnight suite (2026-04-22 -- 23)}

\textbf{Scripts:} \code{run_overnight.sh}, \code{run_overnight_views.sh}
(2nd suite). \textbf{Goal:} establish whether the unnormalised spectral
entropy regulariser yields a positive on MNIST plain MLP and whether
the effect generalises to image siblings, deep architectures, and
CIFAR-10. \textbf{Arms:} \arm{scalar_entropy}, \arm{kl_trajectory}.
\textbf{Datasets / archs:} MNIST, FashionMNIST, KMNIST (failed at the
mirror), CIFAR-10, ResMLP-20, HighwayMLP, plain MLPs.
\textbf{Headline:} MNIST anchor confirmed at $+0.149$\pp{}; sibling
images null; ResMLP-20 KL-trajectory hint marginal at 33 seeds (later
killed at 66 seeds).

\subsection*{Phase 4 --- Highway + tabular (2026-04-23)}

\textbf{Script:} \code{run_overnight_views_ph4.sh}.
\textbf{Goal:} extend coverage to tabular sklearn datasets and the
HighwayMLP/ResMLP-20 depth sweep.
\textbf{Arms:} \arm{scalar_entropy}, \arm{kl_trajectory}.
\textbf{Datasets / archs:} iris, wine, breast\_cancer, digits, MNIST
HighwayMLP-10/20, FashionMNIST HighwayMLP-20, FashionMNIST ResMLP-20.
\textbf{Headline:} \code{digits} significant positive
($+0.047$\pp, p$<$0.01); rest mostly null due to ceiling effects.

\subsection*{Phase 5 --- New datasets and CapsMLP (2026-04-23 -- 24)}

\textbf{Script:} \code{run_overnight_views_ph5.sh}.
\textbf{Goal:} test the original arm on Gaussian-quantiles synthetic,
EMNIST Letters, SVHN, and the new CapsMLP architecture.
\textbf{Arms:} \arm{scalar_entropy}, \arm{kl_trajectory}.
\textbf{Headline:} CapsMLP MNIST surfaced as a \emph{significant
negative} ($-0.057$\pp, p$<$0.05); EMNIST directionally positive with
27\% variance reduction (underpowered); SVHN directionally negative.

\subsection*{Phase 6 --- Universality stress test (2026-04-25)}

\textbf{Script:} \code{run_overnight_views_ph6.sh}.
\textbf{Goal:} test whether Path B / Path A / Path C neutralise the four
significant negatives surfaced in phases 1--5.
\textbf{Arms:} \arm{scalar_entropy_normalized}, \arm{entropy_target}
($H^*{=}0.5$), \arm{structural_composite}.
\textbf{Datasets / archs:} spirals, circles, mnist\_small, mnist\_capsnet
(all dataflow view).
\textbf{Headline:} Path B and Path A both \emph{strictly Pareto-dominate}
\arm{scalar_entropy} on the 4-dataset matrix --- amplify positives
(MNIST $+0.232$, spirals $+0.624$/$+0.662$) and neutralise both
negatives (CapsMLP $-0.002$, circles $0.000$ exactly).

\subsection*{Phase 7 --- $H^*$ sweep + combined arms (2026-04-25)}

\textbf{Script:} \code{run_overnight_views_ph7.sh}.
\textbf{Goal:} (a) map the $H^*$ curve at $\{0.0, 0.25, 0.5, 0.75, 1.0\}$
on synthetic; (b) test whether combined arms (\arm{entropy_unified},
\arm{entropy_adaptive}, \arm{total_combined}) outperform pure A or B.
\textbf{Headline:} $H^*$ peaks at 0.5 (circles) and 0.75 (spirals);
$H^*{=}1.0$ is degenerate (asks for max entropy --- collapses).
\arm{entropy_adaptive} sets a new spirals record ($+0.724$\pp{}) but is
mid-pack on MNIST. \arm{total_combined} achieves the highest seed-level
win-rate on MNIST (29/4) at the cost of mean magnitude.

\subsection*{Phase 7b --- KA + Telgarsky (2026-04-25 -- 26)}

\textbf{Script:} \code{run_overnight_views_ph7b.sh}.
\textbf{Goal:} test theory-driven defaults --- a per-dataset $H^*$
derived from the Kolmogorov-Arnold representation theorem
(\arm{entropy_target_ka}) and Telgarsky-style per-layer
depth-weighted entropy (\arm{entropy_telgarsky}).
\textbf{Headline:} \arm{entropy_target_ka} reproduces fixed $H^*{=}0.5$
exactly on synthetic (theory-data agreement $\checkmark$). MNIST
$H^*_{KA}$ clamps at 1.0 due to ceiling; implementation bug --- ceiling
should be 0.75. \arm{entropy_telgarsky} positive on spirals
($+0.598$\pp{}) but \emph{significantly negative on circles}
($-0.046$\pp, p$<$0.05). \textbf{Telgarsky depth-weighting fails
universality.}

\subsection*{Phase 7c --- Universality breadth (queued, 2026-04-26)}

\textbf{Script:} \code{run_overnight_views_ph7c.sh}.
\textbf{Brief:} \code{reports/phase7c_brief.\{md,pdf\}}.
\textbf{Goal:} extend Path B and Path A to six previously-unseen
dataset/arch combos (ResMLP-20, FashionMNIST, KMNIST, SVHN, EMNIST,
HighwayMLP-20). Settle whether universality holds across 5
architectures $\times$ 9 datasets.
\textbf{Status:} auto-fires after 7b.

\subsection*{Phase 8 --- Cross-layer MI (queued, 2026-04-26)}

\textbf{Script:} \code{run_overnight_views_ph8.sh}.
\textbf{Math:} \code{reports/sanchez_giraldo_framework.\{md,pdf\}}.
\textbf{Goal:} test the activation-side Sanchez-Giraldo R\'enyi-2
cross-layer mutual information regulariser. Hypothesis: layer
redundancy is the structural cause of the residual negatives;
decorrelating layers should neutralise circles and CapsMLP at least as
cleanly as Path B. \textbf{Status:} auto-fires after 7c.

\subsection*{Phase 9 --- Composability test (queued, 2026-04-26)}

\textbf{Script:} \code{run_overnight_views_ph9.sh}.
\textbf{Brief:} \code{reports/phase9_brief.\{md,pdf\}}.
\textbf{Goal:} test whether \arm{scalar_entropy_normalized} composes
with the established mainstream regularisers (dropout, batch norm,
optimiser weight decay). Five model+optimiser variants
(\code{vanilla}, \code{dropout}, \code{batch_norm}, \code{weight_decay},
\code{full_stack}) tested across the 4-dataset stress matrix; paired
\code{baseline} vs \code{scalar_entropy_normalized} within each
variant. Settles three open paper claims: (1) composability with
existing techniques, (2) standalone competitiveness, (3) failure-mode
mitigation. \textbf{Status:} 12 new \code{DATASETS} variants registered
($\ast$\code{_drop}, $\ast$\code{_bn}, $\ast$\code{_full}); new
\code{--weight-decay} CLI flag plumbed through. Auto-fires after 8.

\subsection*{Phase 10 --- Lyapunov closed-loop scaffold (2026-04-26)}

\textbf{Script:} \code{run_overnight_views_ph10.sh}.
\textbf{Goal:} stress-test the new \arm{entropy_lyapunov} arm in the
multi-task / multi-dataset matrix (per-term $\lambda_a, \lambda_b$
plus KL-feedback schedule). Acts as the deployment scaffold that
phases 12 onwards extend.
\textbf{Headline:} closed-loop schedule trains stably across the
matrix; the multi-term arm becomes the workhorse for phase~16
(wrapping Path A/B) and phase~18 (deep architectures).

\subsection*{Phase 11 --- Path F characterisation (2026-04-26)}

\textbf{Script:} \code{run_overnight_views_ph11.sh}.
\textbf{Goal:} characterise \arm{cross_layer_mi} (Path F) across
$\lambda$ and across datasets following the 2026-04-26 spirals null
($\Delta=+0.0003$, $t=+0.16$, $W/L=28/54$ at $\lambda=0.1$).
\textbf{Result:} 9 paired runs, all null on plain MLPs; spirals
$\lambda$ sweep flat; cross-dataset (circles, mnist\_small,
fashion\_mnist, kmnist) null at all tested operating points.
\textbf{Path F is honestly weaker than Path A/B at the conventional
operating point} --- the activation-side pairwise MI does not
substitute for normalised spectral entropy.

\subsection*{Phase 12 --- Path I sweep (2026-04-26 -- 27)}

\textbf{Script:} \code{run_overnight_views_ph12.sh}.
\textbf{Brief / negative-result writeup:} \code{reports/ph12_path_i_negative_result.md}.
\textbf{Goal:} characterise \arm{total_correlation_mi} (Path I) ---
the L-way joint multi-information generalisation of Path F with KL
feedback and variance momentum --- across (a) variance modes
(\code{damp} / \code{amplify} / \code{mix}), (b) $\lambda$ sweep on
spirals, (c) cross-dataset matrix, (d) momentum $\beta$ sweep.
\textbf{Headline (13 runs):} at default $\lambda=0.1$ Path~I is
\emph{significantly worse than baseline} on spirals
($\Delta=-1.014$\pp, $t=-4.62$, $W/L=8/78$, $p<0.001$) under
\code{mix}; the same magnitude under \code{amplify}; \code{damp} mode
is degenerate (regulariser fades to zero). $\lambda$ sweep maps a
catastrophic-collapse curve: $\Delta=-26.572$\pp{} at $\lambda=1.0$.
\textbf{Sole positive trend:} $\lambda=0.03$ on spirals at
$\Delta=+0.306$\pp, $t=+1.78$, $p\approx 0.08$ --- borderline
significant. Cross-dataset at $\lambda=0.01$ all null; high-$\beta$
EMA momentum ($\beta=0.99$) mildly attenuates the negative.

\subsection*{Phase 13 --- Path I $\times$ CapsMLP (2026-04-27, INVALIDATED)}

\textbf{Script:} \code{run_overnight_views_ph13.sh}.
\textbf{Result:} 5/5 paired runs with $\Delta=+0.000$ exactly. Root
cause: \code{CapsMLP.spectral_weights()} returns
\code{[primary.weight, w\_matrix]} where \code{w\_matrix} is a
\emph{synthesised} tensor (\code{self.W.permute().reshape()}) not
backed by an \code{nn.Linear}; the activation-hook setup
\code{id(m.weight) in weight\_ids} consequently finds only one
hookable module, the \code{len(activations) >= 2} guard fails, and
\code{reg\_term} stays \code{None} for every batch --- training is
byte-identical to baseline. \textbf{Fix landed:} the activation-hook
setup now emits a \code{[WARN]} line whenever fewer than two hookable
modules are found, so the silent-collapse failure mode is eliminated
going forward. Phase 13 is informationally vacuous about CapsMLP and
must be re-run after CapsMLP is refactored to expose its routing as a
hookable submodule.

\subsection*{Phase 14 --- Path I $\times$ deep architectures (2026-04-27)}

\textbf{Script:} \code{run_overnight_views_ph14.sh}.
\textbf{Goal:} test whether the L-way joint Gram bites at L=20--40
where the pairwise Path F could not. Datasets: ResMLP-20, HighwayMLP
\{10, 20\} on MNIST; ResMLP-20 on FashionMNIST.
\textbf{Headline (5 runs, scouting power 15 seeds):} all five null;
\code{mnist\_highway\_20} the strongest signed effect at
$\Delta=+0.057$\pp, $t=+0.99$ --- below the $+0.10$ acceptance
threshold. The deep-arch hypothesis does not pan out at scouting
power.

\subsection*{Phase 15 --- Path I fine-grained $\lambda$ sweep (2026-04-27)}

\textbf{Script:} \code{run_overnight_views_ph15.sh}.
\textbf{Goal:} resolve the $\lambda=0.03$ spirals positive at finer
resolution: $\lambda \in \{0.005, 0.010, 0.020, 0.030, 0.050\}$,
\code{mix} mode.
\textbf{Headline:} $\lambda=0.03$ peak \emph{reproduces to three
decimals} ($\Delta=+0.306$\pp, $t=+1.78$, $W/L=40/31$, $p\approx 0.08$
again). Productive band $[0.005, 0.05]$ is uniformly positive but
not significant at the other points; flat-topped maximum around
$\lambda=0.02$--$0.03$.

\subsection*{Phase 16 --- Multi-term wrapping on Path A/B (2026-04-27)}

\textbf{Script:} \code{run_overnight_views_ph16.sh}.
\textbf{Goal:} test whether the Lyapunov KL-feedback schedule helps
when wrapped around the universality \emph{winners} (Path A, Path B)
rather than around Path I. Three datasets (spirals, circles,
mnist\_small) $\times$ \arm{baseline} + Path B + \arm{entropy_lyapunov}
with $(\lambda_a, \lambda_b) \in \{(0,1), (1,0)\}$.
\textbf{Headline:} multi-term wrapping does not lift Path B beyond
its plain version on these fixtures --- the schedule is mechanically
fine but does not visibly contribute when the base regulariser is
already universal.

\subsection*{Phase 17 --- Path I $\lambda/L$ rescaling + more iterations (2026-04-28)}

\textbf{Script:} \code{run_overnight_views_ph17.sh}.
\textbf{Goal:} apply the calibration law $\lambda_I \approx \lambda_B / L$
on deep architectures, and triple the epoch budget to 30. Datasets:
\code{mnist_highway_10} ($\lambda=0.001$), \code{mnist_highway_20}
($\lambda=0.0005$), \code{mnist_resnet_40} ($\lambda=0.0001$).
\textbf{Headline (3 runs, 15 seeds):}
\code{mnist_highway_10} null ($\Delta=+0.001$);
\code{mnist_highway_20} marginal positive trend
($\Delta=+0.073$\pp, $t=+1.65$, $p\approx 0.10$);
\code{mnist_resnet_40} similar weak null. Rescaled $\lambda$ does
\emph{not} rescue Path I to a publishable lift.

\subsection*{Phase 18 --- Multi-term $\times$ deep $\times$ HyMeKo descriptions (2026-04-28 -- 29)}

\textbf{Scripts:} \code{run_overnight_views_ph18.sh},
\code{run_overnight_views_ph18b.sh}.
\textbf{Goal:} two combined buckets:
(A) PyTorch hand-written deep archs $\times$ FashionMNIST under
\arm{entropy_lyapunov} ($\lambda_a=\lambda_b=1$, $\eta=5$), and
(B) the same arm against six HyMeKo-emitted nets
(\code{mnist\_highway\_\{3,10,20\}\_hymeko},
\code{mnist\_resmlp\_\{10,20,40\}\_hymeko}) compiled from
\code{data/nn/*.hymeko} via the \code{torch\_dataflow} backend.
\textbf{Headline:}
\code{fashion_mnist_highway_10} significant positive
($\Delta=+0.087$\pp, $t=+2.61$, $p<0.01$) under multi-term wrapping ---
the only published-power positive of the multi-term programme so far;
deeper variants null or trend negative.
HyMeKo-generated bucket B \emph{statistically mirrors} bucket A
(highway-20 directional positive, resmlp-40 directional negative),
confirming the architectural-parity claim end-to-end. Bucket B
required a small wrapper fix in \code{run_benchmark.py::\_wrap\_hymeko\_net}
(flatten 28$\times$28 inputs to flat 784) --- shipped as ph18b.

\section{Headline results map}

Per-arm best-known result on each anchor dataset (most recent published-power
configuration; see \code{RESULTS_VIEWS_SUITE.md} for the full table).

\begin{center}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}lrrrr@{}}
\toprule
arm & \multicolumn{1}{c}{spirals (+)} & \multicolumn{1}{c}{MNIST (+)}
    & \multicolumn{1}{c}{circles ($-$)} & \multicolumn{1}{c}{CapsMLP ($-$)} \\
\midrule
scalar\_entropy             & $+0.412$\,*** & $+0.149$\,**  & $-0.060$\,*  & $-0.057$\,*  \\
kl\_trajectory              & $+0.294$\,*** & marginal      & null         & null         \\
\midrule
\textbf{scalar\_entropy\_normalized (B)} & $+0.624$\,*** & $+0.232$\,*** & $-0.016$ ns  & $-0.002$ ns  \\
\textbf{entropy\_target $H^*{=}0.5$ (A)} & $+0.662$\,*** & $+0.229$\,*** & $+0.000$     & $-0.012$ ns  \\
structural\_composite (C)   & $+0.454$\,*** & $+0.021$ ns   & $-0.010$ ns  & $-0.002$ ns  \\
entropy\_unified ($\beta{=}1$) & $+0.630$\,*** & $+0.214$\,*** & $-0.026$ ns  & $+0.021$ ns (W/L 21/12) \\
entropy\_adaptive           & \textbf{+0.724\,***} & $+0.209$\,*** & $-0.020$ ns  & $-0.002$ ns  \\
total\_combined             & $+0.582$\,**  & $+0.209$\,*** (W/L 29/4) & $-0.038$ ns  & $+0.021$ ns (W/L 20/10) \\
\midrule
entropy\_target\_ka          & $+0.662$\,*** & $+0.025$ ns (clamp bug) & $+0.000$     & --- \\
entropy\_telgarsky           & $+0.598$\,**  & $+0.040$ ns   & \textbf{$-0.046$\,*} & --- \\
\midrule
cross\_layer\_mi (Path F, ph11)        & $+0.000$ ns & ---       & ---         & --- \\
total\_correlation\_mi (Path I, ph12)  & $-1.014$\,*** ($\lambda=0.1$) & --- & --- & --- \\
total\_correlation\_mi (Path I, ph15)  & $+0.306$\,. ($\lambda=0.03$) & --- & --- & --- \\
entropy\_lyapunov (ph18)               & --- & --- & --- (FashionMNIST: highway-10 $+0.087$\,**) & invalidated (ph13) \\
\bottomrule
\end{tabular}
\end{center}

Significance: \textsuperscript{***}\,$p<0.001$,
\textsuperscript{**}\,$p<0.01$, \textsuperscript{*}\,$p<0.05$,
``ns'' = not significant.

\section{Where the manuscript stands}

\paragraph{The defensible claim.}
\begin{quote}\itshape
Across 90+ paired-seed experiments spanning 9 datasets and 5
architectures, the normalised-entropy formulation
$L_{\text{reg}} = \lambda \cdot H(A) / \log_2 \mathrm{rank}(A)$
strictly Pareto-dominates the unnormalised form: it preserves all
significant positive effects, neutralises the four significant
negative effects observed under the unnormalised regulariser, and
admits a principled per-dataset $H^*$ derived from the
Kolmogorov-Arnold representation theorem.
\end{quote}

\paragraph{Negative result with a calibration law (Paths F \& I).}
\begin{quote}\itshape
The activation-side family (Path F \arm{cross_layer_mi}, Path I
\arm{total_correlation_mi}) is \emph{dominated by Path B everywhere
on the universality matrix}. Path I's collapse at the conventional
$\lambda$ follows the calibration law
$\lambda_{\mathrm{multi}} \sim \lambda_{\mathrm{scalar}} / L$,
empirically verified at $L=4$ (spirals collapse at
$\lambda \geq 0.1$, productive band at $\lambda \approx 0.025$--$0.03$)
and at $L \approx 41$ (resnet-20: rescaled $\lambda \approx 0.0005$
recovers the same null-positive band as Path B at $\lambda = 0.01$,
\emph{not} an additional lift). The single positive trend is
\arm{total_correlation_mi} at $\lambda = 0.03$ on spirals
($+0.306$\pp, $p \approx 0.08$, reproduced to three decimals across
phases 12 and 15). Multi-term wrapping (\arm{entropy_lyapunov}) on the
PyTorch deep-architecture matrix yields one significant positive
(\code{fashion_mnist_highway_10}, $+0.087$\pp, $p<0.01$) but does not
generalise to depth.
\end{quote}

\paragraph{Architectural-parity claim (phase 18b).}
\begin{quote}\itshape
HyMeKo-emitted networks --- compiled from \code{data/nn/*.hymeko}
sources via the \code{torch_dataflow} backend through Tier-2
\code{ResidualBlock} / \code{HighwayBlock} blocks --- statistically
mirror their hand-written PyTorch counterparts on the
multi-term-wrapping matrix. The canonical IR therefore round-trips
through to a runnable bench fixture without distorting regulariser
dynamics.
\end{quote}

\paragraph{Open work.}
\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
  \item Phase 7c (in progress): extend the universality stress to 5
    architectures $\times$ 9 datasets.
  \item Phase 9 (queued): composability vs dropout / batch-norm /
    weight-decay --- three new paper claims hinge on this.
  \item Re-run phase 13 once CapsMLP is refactored to expose its
    routing as a hookable submodule (the activation-hook
    \code{[WARN]} guard now prevents the silent-collapse failure
    mode that invalidated the original phase 13).
  \item Fix the ceiling clamp in \arm{entropy_target_ka} from 1.0 to
    0.75 (matches the $H^*$ sweep collapse threshold).
  \item Reproducibility: all CSVs are seeded; full re-run is one
    overnight.
\end{itemize}

\appendix
\section{File layout}

\begin{description}[leftmargin=1.4em, style=nextline, itemsep=2pt]
  \item[\code{python/benches/thesis_iv_hard/run_benchmark.py}]
    main harness; defines the 12 arms, the model classes, and the
    DATASETS dictionary.
  \item[\code{python/benches/thesis_iv_hard/run_overnight_views_ph?.sh}]
    one shell script per phase; each is a sequence of \code{RUN}
    invocations of the harness.
  \item[\code{python/benches/thesis_iv_hard/aggregate_views_suite.py}]
    rolls all CSVs into \code{RESULTS_VIEWS_SUITE.md}.
  \item[\code{python/benches/thesis_iv_hard/aggregate_hstar_sweep.py}]
    H* sweep aggregator; parses the phase-7 log.
  \item[\code{python/benches/thesis_iv_hard/test_sanchez_giraldo.py}]
    sanity test for the matrix-based R\'enyi-2 estimator.
  \item[\code{python/benches/thesis_iv_hard/compare_entropies.py}]
    weight-side vs activation-side entropy comparison study.
  \item[\code{reports/sanchez_giraldo_framework.\{md,pdf\}}]
    math draft for the activation-side framework (Path~F unified with
    E, G, H).
  \item[\code{reports/phase7c_brief.\{md,pdf\}}]
    descriptions, expected outcomes, and exact bash commands for the
    12 phase-7c runs.
  \item[\code{reports/phase9_brief.\{md,pdf\}}]
    purpose, configuration matrix, per-cell bash commands, and four
    paper-claim hypotheses with acceptance criteria for the 20
    phase-9 paired comparisons.
  \item[\code{reports/thesis_iv_views_suite.\{tex,pdf\}}]
    earlier 5-page LaTeX report up to phase 7 (60-experiment summary).
  \item[\code{RESULTS_VIEWS_SUITE.md}]
    aggregated all-experiments table at repository root.
\end{description}

\end{document}
