% reports/thesis_iv_executive_summary.tex
%
% One-page executive abstract of Thesis IV (the spectral-entropy
% regularisation programme). Drop-in for a thesis-chapter intro or
% a short stand-alone summary; pairs with the full taxonomy at
% reports/phases_and_paths.tex and the negative-result writeup at
% reports/ph12_path_i_negative_result.md.

\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.2cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{parskip}
\hypersetup{colorlinks=true, linkcolor=black, urlcolor=blue}

\newcommand{\arm}[1]{\texttt{#1}}
\newcommand{\pp}{\,pp}

\title{Thesis IV --- Executive Summary\\
       \large Spectral-Entropy Regularisation on Signed-Incidence Hypergraphs}
\author{C.~Hajdu, \'A.~B.~Csap\'o, K.~Kov\'acs}
\date{2026-04-29}

\begin{document}
\maketitle

\section*{Question}

Can a single regulariser, derived from the spectral entropy of a
signed-incidence hypergraph view of a neural network, deliver
\emph{universal} improvement across architectures and datasets, or do
the empirical positives reported in the literature collapse under
matched cross-fixture testing?

\section*{Programme}

\textbf{18 phases, 2026-04-22 to 2026-04-29.} Twelve regulariser
\emph{paths} (A--I plus combination arms) tested under a single
benchmark harness on nine datasets (MNIST, FashionMNIST, KMNIST,
CIFAR-10, EMNIST, SVHN, two\_moons, spirals, circles) and five
architectures (plain MLP, ResMLP, HighwayMLP, CapsMLP, plus the
HyMeKo-emitted variants of the first three). Total volume:
roughly 90+ paired-seed comparisons, $\sim$3{,}000 individual
training runs, 0 statistical-power failures, all CSVs reproducible
from seeded scripts.

\section*{Settled positive --- the universality claim}

The normalised-entropy formulation
\[
\mathcal{R}_B(A) = H_{\text{norm}}(A) = \frac{H(A)}{\log_2 \mathrm{rank}(A)}
\]
\textbf{strictly Pareto-dominates} the unnormalised form
$\mathcal{R}_0 = \lambda H(A)$ across the universality matrix:
every significant-positive effect of $\mathcal{R}_0$ is preserved or
amplified, and \emph{all four} significant-negative effects of
$\mathcal{R}_0$ (capsule networks on MNIST, two-circle synthetic, deep
ResMLP-20, Highway networks at depth) are neutralised
($p\!\geq\!0.05$). The Kolmogorov--Arnold-derived target
$H^\star_{KA}=\log_2(2n+1)/\log_2 \mathrm{rank}(A)$ supplies a
principled per-dataset $H^\star$ and removes the one remaining free
hyperparameter.

\section*{Settled negative --- the activation-side family with a calibration law}

The activation-side paths --- \arm{cross_layer_mi} (Path F,
pairwise Sanchez-Giraldo Rényi-2 MI) and \arm{total_correlation_mi}
(Path I, L-way joint multi-information with KL-feedback $\lambda$ and
variance-momentum modulator) --- are \textbf{dominated by Path B
everywhere on the matrix}. The collapse is not arbitrary: it follows
the calibration law
\[
\lambda_{\mathrm{multi}} \sim \lambda_{\mathrm{scalar}} / L,
\]
empirically verified at $L\!=\!4$ (spirals collapse at
$\lambda\!\geq\!0.1$, productive band at
$\lambda\!\approx\!0.025$--$0.03$) and at $L\!\approx\!41$
(ResMLP-20: rescaled $\lambda\!\approx\!0.0005$ recovers the same
null-positive band as Path B at $\lambda\!=\!0.01$, \emph{not} an
additional lift). The single positive trend is Path I at
$\lambda\!=\!0.03$ on spirals at $\Delta\!=\!+0.306$\pp{}
($t\!=\!+1.78$, $p\!\approx\!0.08$), reproduced to three decimal
places across two independent phases.

The closed-loop multi-term wrapping
$\lambda_{\text{eff}}=\lambda_0\exp(-\eta\,D_{KL}(\hat\lambda^{(t-1)}\|\hat\lambda^{(t)}))$
(\arm{entropy_lyapunov}) yields two significant published-power
positives at FashionMNIST: \texttt{fashion\_mnist\_highway\_10}
($L\!=\!10$) at $\Delta\!=\!+0.092$\pp, $p\!<\!0.05$ at 10-seed
(ph18c confirmation of the original ph18 result), and
\texttt{mnist\_resmlp\_40\_hymeko} ($L\!=\!40$) at
$\Delta\!=\!+0.213$\pp, $p\!<\!0.05$ once $\lambda_{a,b}$ is
rescaled to $0.5$ per the depth-aware calibration law
$\lambda_{\rm multi}\sim\lambda_{\rm scalar}/L$. Highway-20
remains a genuine depth-fragility null across
$\lambda\!\in\!\{0.5,1.0,2.0,5.0\}$.

\section*{Architectural-parity claim}

Networks emitted from \texttt{.hymeko} sources via the
\texttt{torch\_dataflow} backend (Tier-2 \texttt{ResidualBlock} /
\texttt{HighwayBlock} blocks; \texttt{*\_hymeko} dataset entries in
the bench harness) statistically mirror their hand-written PyTorch
counterparts on the same regulariser arms, confirming that the
canonical IR round-trips through to a runnable bench fixture
without distorting regulariser dynamics.

\section*{Bounds (what is not yet shown)}

\textit{Composability} (Path B alongside dropout / batch-norm /
optimiser weight decay) is queued as Phase~9; the three paper
claims that hinge on it are not yet defended. \textit{CapsMLP under
activation-side regularisers} requires a refactor of CapsMLP's
synthesised routing tensor before Phase~13 can be re-run. The
\texttt{entropy\_target\_ka} ceiling clamp at 1.0 is a known
implementation bug and should be 0.75 (matches the empirical
$H^\star$ collapse threshold).

\section*{Artifacts}

Code: \texttt{python/benches/thesis\_iv\_hard/} (harness, sweep
scripts, aggregator). Generated nets:
\texttt{python/benches/thesis\_iv\_hard/generated\_nets/} +
\texttt{data/nn/*.hymeko}. Reports:
\texttt{reports/phases\_and\_paths.\{tex,pdf\}} (full taxonomy),
\texttt{reports/ph12\_path\_i\_negative\_result.md} (negative-result
writeup with the 13-run paired-$\Delta$ table),
\texttt{reports/sanchez\_giraldo\_framework.\{md,pdf\}} (math for
Paths E--H of the activation-side family),
\texttt{RESULTS\_VIEWS\_SUITE.md} (aggregated all-experiments
table). Every CSV under \texttt{data/benchmarks/} is seeded and
reproducible by re-running the corresponding \texttt{ph*.sh}
script.

\end{document}
