\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=2.3cm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}
\newcommand{\sig}{\ensuremath{\sigma}}
\newcommand{\Gomb}{G\"omb}
\title{Plan: \Gomb-strict Bitcoin architectural levers}
\date{2026-05-14}
\author{}
\begin{document}
\maketitle

\section{Hypothesis}

The current \Gomb-strict numbers on Bitcoin
($0.8972 \pm 0.0079$ Alpha, $0.9145 \pm 0.0068$ OTC) sit
$\sim 0.03$ AUC below published transductive baselines
(SGCN: $0.929 / 0.942$). The configs were inherited from a prior
joint\_mix Optuna sweep that targeted a different protocol; they
are not architecturally optimised for strict-protocol Bitcoin.

We hypothesise that one or more single-knob changes from the
v5\_combined baseline can lift Bitcoin strict above the published
transductive numbers, on the same consumer hardware. No Optuna ---
manual lever picks based on the dataset's small-and-dense regime.

\section{Levers (each from the v5\_combined baseline)}

\begin{center}
\begin{tabular}{lll}
\toprule
Label & Change vs v5\_combined & Why \\
\midrule
L1 c5\_arity   & cycle\_ks = c3+c4+c5+w2+w3        & Bitcoin has many 5-cycles \\
L2 deeper      & n\_tiers $= 3 \to 5$                & deeper CPML routing \\
L3 wider\_outer & M\_outer $= 8 \to 16$               & more Clifford-FIR kernels \\
L4 long\_train & n\_epochs $= 80 \to 200$            & Bitcoin is small; can afford \\
L5 big\_embed  & d\_embed $= 32 \to 64$              & small graph, big embedding fits \\
\bottomrule
\end{tabular}
\end{center}

Each lever is independent and additive --- if multiple lift the
number, a follow-up phase can combine them (no Optuna, just
stacking the levers that pass the test).

\section{Scope and goal}

5 levers $\times$ 2 datasets (Bitcoin Alpha, Bitcoin OTC)
$\times$ 5 seeds = 50 runs. At $\sim$30 s/seed on the RTX 2070
SUPER, total wall time $\sim$25 min plus overhead $\approx$ 1 h.

Goal: identify the strongest single lever (or two) on Bitcoin
strict. Then a follow-up phase can stack them.

\section{Affected files}

\begin{itemize}
\item \texttt{signedkan\_wip/experiments/run\_gomb\_bitcoin\_levers\_2026\_05\_14.sh}
  --- new runner script.
\item No source edits. The runner just varies CLI flags on the
  existing \texttt{run\_gomb\_smoke.py} entry point.
\end{itemize}

\section{CORE.YAML items touched}

None.

\section{Test strategy}

This is an experiment, not a code change. The protocol is the
same strict-protocol 5-seed evaluation used in the
\Gomb-strict main benchmark
(\texttt{run\_gomb\_strict\_benchmark\_2026\_05\_14.sh}).

\section{Performance budget}

\begin{itemize}
\item Total wall: $\leq 90$ minutes.
\item Per-seed wall: $\leq 120$ s (was 30 s for v5\_combined; long\_train lever extends to $\sim$75 s).
\item Peak RSS: $\leq 8$ GB (Bitcoin is small).
\item GPU memory: $\leq 6$ GB on RTX 2070 SUPER.
\end{itemize}

\section{Rollback path}

The runner script is independent of any source. To rollback:
delete the script. No source files touched.

\section{Risk anticipation}

\begin{itemize}
\item \emph{No-lift outcome.} If every lever fails to beat SGCN
  transductive on Bitcoin, the honest framing
  (``strict-protocol penalty is real on Bitcoin'') stands; the
  experiment is documented as a negative result.
\item \emph{Single-lever lift.} If one lever wins, a follow-up
  phase runs a 5-seed paired comparison with the v5\_combined
  baseline on the same seeds.
\item \emph{Multi-lever interaction.} If two or more lift
  independently, a follow-up phase stacks them; no Optuna.
\end{itemize}

\section{Decision tree}

\begin{itemize}
\item Any lever lifts Alpha \emph{and} OTC by $\geq +0.02$ AUC
  $\Rightarrow$ promote to a 5-seed paired test against
  v5\_combined; update the headline tables.
\item All levers $< +0.01$ AUC $\Rightarrow$ document as a
  negative-result anchor; strict penalty is real on this dataset
  regime.
\item Some levers help one dataset but not the other
  $\Rightarrow$ document dataset-specific configs in the paper's
  ``configuration policy'' section.
\end{itemize}

\section{Bottom line}

5 manual lever picks, 50 runs, $\sim$1 hour total. No Optuna. The
hypothesis is that \Gomb's architectural capacity exceeds what the
current Bitcoin config exploits, and that strict-protocol Bitcoin
can match or exceed published transductive numbers without going
down the hyperparameter-tuning rabbit hole.

\end{document}
