\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{enumitem}
\setlist{nosep,leftmargin=*}
\lstset{basicstyle=\ttfamily\small,frame=single,breaklines=true}
\title{Plan: Vertex Pre-filter for the Per-Vertex Top-$K$ Cycle Enumerator}
\author{HyMeKo / HSiKAN}
\date{2026-05-11}
\begin{document}
\maketitle

\section{Scope and goal}
Add a pluggable \emph{vertex pre-filter} step before the per-vertex DFS in
\texttt{hymeko\_graph::topk\_cycles}, so vertices unlikely to contribute
informative cycles are excluded from being DFS roots.  The lift study on
2026-05-10 measured that \textbf{61\% of Epinions vertices sit on no
balanced 4-cycle reachable from the DFS} --- they participate in the
enumeration cost (per-vertex BFS pre-pass + DFS startup) but contribute
zero output.  Skipping them is pure work elimination.

\textbf{Concrete goal.} Cut the Epinions production-config $k{=}5$
enumeration wall and peak RSS by $\geq 30\%$ at iso-AUC, with a
mechanism that supports multiple filter strategies (degree, triangle,
clustering, local sign-entropy, clique size, balanced-reachable) so
that future regimes (Slashdot, Bitcoin Alpha/OTC, future datasets) can
pick the best filter for their structure.

\section{Affected files}
\textbf{All heuristic computation lives in Rust} --- on Epinions scale
(131k vertices, 841k edges) Python triangle/clustering/clique-size
loops would be 10-100$\times$ slower than the savings.  Python only
handles env-var dispatch and passes the filter selection to Rust.
\begin{itemize}
\item \texttt{hymeko\_graph/src/vertex\_filter.rs} --- NEW.  Defines the
  \texttt{VertexFilter} trait + 6 concrete impls
  (\texttt{NoFilter}, \texttt{DegreeFilter}, \texttt{TriangleFilter},
  \texttt{ClusteringFilter}, \texttt{LocalSignEntropyFilter},
  \texttt{CliqueSizeFilter}, \texttt{BalancedReachableFilter}).  Each
  impl returns a \texttt{Vec<u32>} of vertex ids to keep.
\item \texttt{hymeko\_graph/src/topk\_cycles.rs} --- extend
  \texttt{enumerate\_top\_k\_per\_vertex\_cycles\_par(\_adaptive)} to
  accept an optional \texttt{\&dyn VertexFilter} (or pre-computed
  \texttt{\&[u32]} keep-set).  Default behaviour: iterate over
  \texttt{0..n\_nodes} (no filter).  Strictly additive --- existing
  callers see no behaviour change.
\item \texttt{hymeko\_graph/src/lib.rs} --- re-export the trait and
  filter impls.
\item \texttt{hymeko\_py/src/cycles.rs} --- thread a
  \texttt{filter\_kind: \&str} + \texttt{filter\_params: \&PyDict} through
  the existing PyO3 bindings.  Macro-dispatch over the 7 filter kinds at
  the binding boundary, then pass a typed reference to the Rust
  enumerator (no \texttt{Box<dyn>} indirection in the hot path).
\item \texttt{signedkan\_wip/src/n\_tuples.py} --- read
  \texttt{HSIKAN\_VERTEX\_FILTER} env var + threshold env vars; pass
  them through to the Rust binding as a \texttt{(filter\_kind,
  filter\_params)} pair.  No heuristic computation in Python.
\item \texttt{signedkan\_wip/tests/test\_vertex\_filter.py} --- NEW.
  End-to-end tests via the PyO3 binding: each filter strategy on a
  small SBM fixture returns the expected keep-set; AUC parity at
  \texttt{filter\_kind="none"}.
\item \texttt{hymeko\_graph/tests/vertex\_filter.rs} --- NEW.  Rust-side
  unit tests for each \texttt{VertexFilter} impl on hand-coded fixtures.
\item \texttt{hymeko\_graph/tests/vertex\_prefilter\_integration.rs} ---
  NEW.  Filtered enumerator output is a subset of unfiltered;
  bit-identity at \texttt{filter\_kind="none"}; sequential vs parallel
  parity under filter.
\item \texttt{hymeko\_graph/benches/vertex\_filter\_bench.rs} --- NEW.
  Criterion bench: filter computation cost (target $<5$ s on Epinions
  for cheap filters; $<60$ s for \texttt{balanced\_reachable}).
\end{itemize}

\section{CORE.YAML items touched}
\textbf{None.}  \texttt{hymeko\_graph}, \texttt{hymeko\_py},
\texttt{signedkan\_wip} are not under \texttt{CORE.YAML}'s
\texttt{crates}/\texttt{files}/\texttt{globs}; \texttt{tests/**} and
\texttt{examples/**} are in the \texttt{allowlist}.

\section{Interface changes}

\subsection*{Rust trait + impls}
\begin{lstlisting}[language=rust]
// hymeko_graph/src/vertex_filter.rs
pub trait VertexFilter: Send + Sync {
    /// Return the set of vertex ids that should be DFS roots.
    /// `Vec<u32>` ordered ascending.  Empty vec is a valid result.
    fn keep_set(&self, g: &SignedGraph) -> Vec<u32>;
    fn name(&self) -> &'static str;
}

pub struct NoFilter;
pub struct DegreeFilter { pub min_degree: u32 }
pub struct TriangleFilter;
pub struct ClusteringFilter { pub min_cc: f32 }
pub struct LocalSignEntropyFilter { pub min_entropy: f32 }
pub struct CliqueSizeFilter { pub min_clique: u32 }
pub struct BalancedReachableFilter { pub k: usize }

// Composition (AND).
pub struct AndFilter(pub Vec<Box<dyn VertexFilter>>);
\end{lstlisting}

\subsection*{Per-impl complexity (worst-case Epinions)}
\begin{tabular}{lll}
\toprule
Filter & Cost & Wall on Epinions \\
\midrule
\texttt{NoFilter} & $O(1)$ & $<10$ ms \\
\texttt{DegreeFilter} & $O(|E|)$ via bincount & $<100$ ms \\
\texttt{TriangleFilter} & $O(|E| \cdot \overline{d})$ via shared adjacency & $1$--$2$ s \\
\texttt{ClusteringFilter} & $O(|E| \cdot \overline{d})$ same as triangle & $1$--$2$ s \\
\texttt{LocalSignEntropyFilter} & $O(|E|)$ via per-vertex sign histogram & $<200$ ms \\
\texttt{CliqueSizeFilter} ($q$=3) & $O(|E| \cdot \overline{d})$ & $1$--$2$ s \\
\texttt{CliqueSizeFilter} ($q$=4) & $O(|E| \cdot \overline{d}^2)$ & $10$--$30$ s \\
\texttt{BalancedReachableFilter} & one-shot k-cycle enum + mark & $30$--$60$ s \\
\bottomrule
\end{tabular}

\subsection*{Per-vertex enumerator integration}
\begin{lstlisting}[language=rust]
// hymeko_graph/src/topk_cycles.rs
pub fn enumerate_top_k_per_vertex_cycles_par<P, S, F>(
    g: &SignedGraph, k: usize, m_per_vertex: usize,
    pruner: &P, scorer: &S,
    filter: &F,                      // NEW: zero-cost generic
) -> Vec<Cycle>
where
    P: Pruner + Send + Sync,
    S: Scorer + Send + Sync,
    F: VertexFilter + ?Sized;

// Legacy entry-point (calls into _par with NoFilter) preserved
// for backwards compatibility with existing callers.
\end{lstlisting}

\subsection*{PyO3 binding}
\begin{lstlisting}[language=rust]
// hymeko_py/src/cycles.rs
#[pyfunction]
fn enumerate_top_k_per_vertex_cycles_signed_rs(
    eu, ev, es, n_nodes, k, m_per_vertex,
    scorer_kind: &str, pruner_kind: &str,
    // NEW (with default empty for backwards compat):
    filter_kind: &str,                 // "none" | "degree" | ...
    filter_params: &PyDict,            // {"min_degree": 2, ...}
) -> (PyArray2<u32>, PyArray1<f32>);

// Macro-dispatched: each (scorer x pruner x filter) combination
// monomorphises a single concrete enumerator call.
\end{lstlisting}

\subsection*{Python (env-var dispatch only)}
\begin{lstlisting}[language=python]
# signedkan_wip/src/n_tuples.py (no heuristic logic in Python)
filter_kind = os.environ.get("HSIKAN_VERTEX_FILTER", "none")
filter_params = {
    "min_degree": int(os.environ.get(
        "HSIKAN_VERTEX_FILTER_MIN_DEGREE", "2")),
    "min_cc": float(os.environ.get(
        "HSIKAN_VERTEX_FILTER_MIN_CC", "0.05")),
    "min_entropy": float(os.environ.get(
        "HSIKAN_VERTEX_FILTER_MIN_ENTROPY", "0.4")),
    "min_clique": int(os.environ.get(
        "HSIKAN_VERTEX_FILTER_MIN_CLIQUE", "3")),
}
arr, _scores = hymeko.enumerate_top_k_per_vertex_cycles_signed_rs(
    eu, ev, es, n_nodes, k, K, scorer, pruner,
    filter_kind, filter_params,
)
\end{lstlisting}

\subsection*{Env vars}
\begin{lstlisting}
HSIKAN_VERTEX_FILTER = none | degree | triangle | clustering
                     | local_sign_entropy | clique_size
                     | balanced_reachable
                     | "compose:<f1>,<f2>,..."
HSIKAN_VERTEX_FILTER_MIN_DEGREE  = 2     (default)
HSIKAN_VERTEX_FILTER_MIN_CC      = 0.05
HSIKAN_VERTEX_FILTER_MIN_ENTROPY = 0.4
HSIKAN_VERTEX_FILTER_MIN_CLIQUE  = 3
\end{lstlisting}

\section{Test strategy}

\subsection*{Unit (Python, \texttt{tests/test\_vertex\_filter.py})}
\begin{itemize}
\item Each strategy on a 6-vertex hand-coded fixture; assert exact keep-set.
\item Threshold monotonicity: relaxing the threshold returns a superset.
\item Composition: AND of two strategies returns intersection.
\item Empty-graph edge case: returns \texttt{np.empty(0)} cleanly.
\end{itemize}

\subsection*{Unit (Rust, \texttt{tests/vertex\_prefilter.rs})}
\begin{itemize}
\item \texttt{starting\_vertices=Some(\&[v])} returns cycles starting only
  at $v$ (canonical-rotation-aware).
\item \texttt{starting\_vertices=None} matches the existing
  \texttt{enumerate\_top\_k\_per\_vertex\_cycles\_par} bit-for-bit.
\item Sequential vs parallel parity under filtered starting set.
\end{itemize}

\subsection*{Integration}
\begin{itemize}
\item Filtered enumerator output is a \emph{subset} of unfiltered output
  (same cycles, fewer of them).
\item For \texttt{strategy="all"} (no filter), output bit-equals existing
  enumerator.
\item AUC parity check on a small SBM fixture: 1-seed AUC delta
  $\leq 0.005$ between filtered (degree+triangle) and unfiltered.
\end{itemize}

\subsection*{Performance test}
\begin{itemize}
\item Criterion bench: Epinions $k{=}5$, $m{=}128$, balance pruner, with
  vs without vertex pre-filter.  Assert filtered wall $\leq 0.7\times$
  unfiltered wall (i.e., $\geq 30\%$ reduction).
\item Memory: filtered peak RSS $\leq 0.7\times$ unfiltered peak RSS.
\end{itemize}

\section{Performance budget}
\begin{itemize}
\item Filtered wall: $\leq 70\%$ of unfiltered (Epinions $k{=}5$).
\item Filtered RSS: $\leq 70\%$ of unfiltered.
\item Filter-computation overhead: $\leq 5$ seconds for the cheap
  strategies (degree, triangle); $\leq 60$ seconds for
  balanced\_reachable (one-shot enum + mark).
\item Output cycle count: filtered $\leq$ unfiltered (subset property).
\item AUC delta vs unfiltered (1-seed Epinions abbreviated config):
  $\leq 0.005$.  5-seed paired validation gates promotion to default.
\end{itemize}

\section{Rollback path}
\begin{itemize}
\item Default \texttt{HSIKAN\_VERTEX\_FILTER=all} preserves existing
  behaviour bit-for-bit.  Setting the env var is opt-in.
\item The new Rust function is additive; the existing entry point keeps
  its signature. Reverting the patch removes only new code.
\item All five PyO3 bindings remain backwards-compatible: the new
  optional argument defaults to \texttt{None}.
\end{itemize}

\section{Risk anticipation (CLAUDE.md \S2)}

\subsection*{Performance-contract preservation}
The current per-vertex enumerator passes \texttt{m\_per\_vertex} into the
Rust DFS (the per-vertex heap cap).  A separate fix landed today
(\texttt{n\_tuples.py:\_subsample\_arr\_to\_tuples}) that pushes the
\emph{output} cap down to numpy-array level before Python-tuple
materialisation. \textbf{This plan must preserve both contracts.}  The
filter parameter does not interact with either; it bounds the set of
\emph{starting} vertices, not the per-vertex output count.

\subsection*{Aspirational comments to honour}
\texttt{n\_tuples.py:367-369} says ``Push the cap into the enumerator so
we never materialise more than \texttt{max\_cycles} Python tuples in
memory.''  This plan does \emph{not} fix that --- the
\texttt{cap\_total} pushdown is a separate followup
(\texttt{project\_rust\_topk\_memory\_followup\_2026\_05\_10.md} in
auto-memory).  Filter pre-pass reduces the \emph{number of vertices} we
enumerate from, which indirectly reduces total cycles enumerated, but
the per-arity Rust-side memory bound still depends on the
\texttt{cap\_total} fix.

\subsection*{Worst-case input}
Production-scale Epinions, $k{=}5$, $m{=}128$, balance pruner,
all 6 arities (\texttt{c2,c3,c4,c5,w2,w3}).  Filter computed on a
graph with 131{,}828 vertices and 841{,}372 edges.

\subsection*{What could go wrong}
\begin{itemize}
\item \textbf{Filter is more expensive than the savings}: the cheap
  strategies (degree, triangle) are bounded $O(|E|)$, $O(|E|\cdot\overline{d})$
  --- safely under 5s on Epinions.  The expensive ones
  (balanced\_reachable) are scoped behind a separate env var with their
  own perf gate.
\item \textbf{Filter changes AUC}: keep-set monotonicity test +
  fixed-strategy=``all'' bit-parity test guard against accidental
  behaviour change.  AUC parity tested on the smoke and 5-seed gated for
  promotion.
\item \textbf{Rayon thread imbalance}: if the keep-set is small
  ($<100$ vertices) the parallel work stealing might starve.  Mitigation:
  log keep-set size at start; if $<1000$, suggest a smaller filter.
\end{itemize}

\section{Empty-plan-dir hygiene (CLAUDE.md \S2)}
This plan dir starts populated.  If the work is abandoned at any halt
condition (Section 9), the dir gets a \texttt{HALTED.md} explaining the
abort reason rather than being left empty.

\section{Halt conditions}
\begin{itemize}
\item Filter-computation overhead exceeds budget for the chosen strategy.
\item AUC delta on the 1-seed smoke exceeds $0.01$ at any strategy
  (filter is suppressing informative cycles --- pivot to less
  aggressive thresholds).
\item Composition logic forces a CORE.YAML edit.
\end{itemize}

\section{Implementation order (v1 / v2)}

\subsection*{v1 (this plan)}
\begin{enumerate}
\item Rust API extension (\texttt{starting\_vertices: Option<\&[u32]>}).
\item PyO3 binding update.
\item \texttt{vertex\_filter.py} with \texttt{degree}, \texttt{triangle},
  \texttt{compose}, \texttt{all}.
\item \texttt{n\_tuples.py} env-var dispatch.
\item All unit + integration + criterion tests.
\item Smoke on Epinions $k{=}5$ with \texttt{compose:degree>=2+triangle>=1}.
\end{enumerate}

\subsection*{v2 (separate plan, gated on v1 pass)}
\begin{enumerate}
\item \texttt{clustering}, \texttt{local\_sign\_entropy},
  \texttt{clique\_size}, \texttt{balanced\_reachable} strategies.
\item Per-strategy AUC sweep on production config.
\item 5-seed paired validation for promotion of best strategy to
  default.
\item \textbf{Tiered concentric caps (FPN P-levels)}:
  $m_v$ as a step-function of vertex centrality, not linear in
  degree.  P5 (top 0.1\% hubs): $m_v{=}1024$; P4 (1\%): $m_v{=}256$;
  P3 (5\%): $m_v{=}64$; P2 (20\%): $m_v{=}16$; P1 (peripheral):
  skipped.  Total cycles drop $\sim$3$\times$ vs fixed $m{=}128$;
  signal density per row rises sharply.
\end{enumerate}

\subsection*{v3 (Mixture 2: per-tier algorithm selection)}
Algorithmic complexity matches structural importance.  Each FPN tier
gets its own DFS strategy --- the strategy whose heap-fill rate
matches the tier's $m_v$ best.  Composes naturally with v2's tiered
caps and the user's cognitive-propagation framing (refined planning
at the top, bulk reactive enumeration at the bottom).

\begin{center}
\begin{tabular}{llll}
\toprule
Tier & $m_v$ & Algorithm & Justification \\
\midrule
P5 (top 0.1\%) & 1024 & A*-best-first + shared threshold & high $m_v$ rarely fills $\to$ A* finds top cycles fastest \\
P4 (top 1\%)   & 256  & A*-warmup $\to$ DFS-ABB hybrid    & medium heap, both phases help \\
P3 (top 5\%)   & 64   & DFS-ABB                           & heap fills fast, ABB threshold tightens \\
P2 (top 20\%)  & 16   & DFS no-pruning                    & heap so small that ABB overhead $>$ gain \\
P1 (periph.)   & 0    & skip (vertex pre-filter)          & no cycles contributed \\
\bottomrule
\end{tabular}
\end{center}

The Rust trait surface for v3:
\begin{lstlisting}[language=rust]
pub trait PerVertexStrategy: Send + Sync {
    fn enumerate_from(&self, g: &SignedGraph,
                       v: u32, k: usize, m_v: u32,
                       pruner: &dyn Pruner,
                       scorer: &dyn Scorer,
                       shared_threshold: &AtomicF64,
                       ) -> Vec<HeapEntry>;
}

pub struct AStarStrategy { /* priority queue, shared threshold */ }
pub struct AStarToDfsHybridStrategy { warmup_size: usize }
pub struct DfsAbbStrategy;
pub struct DfsNoPruneStrategy;

pub struct TieredDispatcher {
    pub tiers: Vec<(u32 /*m_v*/, Box<dyn PerVertexStrategy>)>,
    pub tier_of_vertex: Vec<usize>,  // n_nodes-long lookup
}
\end{lstlisting}

Note: the v3 trait is per-vertex, not per-graph --- the dispatch
fires once per starting vertex.  Rayon par\_iter over starting
vertices, each grabbing its own tier's strategy.  Shared threshold
is an \texttt{AtomicF64} (f64-as-u64-bits) so all rayon workers
prune against the same top-K-min, regardless of which strategy
produced the threshold.
\end{enumerate}

\subsection*{v3+ (Mixture 3: Pareto-frontier multi-scorer union)}
\textbf{Orthogonal to v3's per-tier algorithm.}  Within each strategy,
run multiple scorers in parallel and union their top-K's --- the
output is Pareto-diverse rather than single-axis-optimal.

\begin{center}
\begin{tabular}{lll}
\toprule
Tier & Strategy (v3) & Scorer set (v3+) \\
\midrule
P5 & A* + shared thresh & \texttt{fraction\_negative} only \\
P4 & A* $\to$ DFS-ABB    & \texttt{fraction\_negative} + \texttt{local\_sign\_entropy} \\
P3 & DFS-ABB             & \texttt{fraction\_negative} + \texttt{low\_root} + \texttt{sign\_product\_abs} \\
P2 & DFS no-prune        & no scorer (all retained) \\
\bottomrule
\end{tabular}
\end{center}

Mechanic per vertex $v$ in tier $T$:
\begin{enumerate}
\item Let scorer set be $S_T = \{s_1, \dots, s_{|S_T|}\}$.
\item For each $s_i$: enumerate top-$\lceil m_v / |S_T| \rceil$ cycles
  ranked by $s_i$ (using $T$'s algorithm strategy).
\item Union the per-scorer outputs, deduplicating by canonical
  cycle key (sorted vertex tuple + edge-sign tuple).
\item Output: at most $m_v$ cycles, Pareto-diverse across scorers.
\end{enumerate}

The Rust trait extension for v3+:
\begin{lstlisting}[language=rust]
pub struct TierConfig {
    pub m_v: u32,
    pub strategy: Box<dyn PerVertexStrategy>,
    pub scorers: Vec<Box<dyn Scorer>>,   // NEW
}

impl TierConfig {
    pub fn enumerate_from(&self, g: &SignedGraph, v: u32,
                           k: usize, ...) -> Vec<Cycle> {
        let per_scorer = (self.m_v as usize)
                          .div_ceil(self.scorers.len());
        let mut all = HashMap::new();
        for scorer in &self.scorers {
            for cycle in self.strategy.enumerate_from(
                    g, v, k, per_scorer as u32, scorer, ...) {
                all.entry(cycle.canonical_key())
                   .or_insert(cycle);
            }
        }
        all.into_values().take(self.m_v as usize).collect()
    }
}
\end{lstlisting}

\textbf{Why this is good (justification):}
\begin{itemize}
\item Each scorer captures a different aspect of cycle quality
  (balance ratio, root stability, sign-product abs).  No single
  scorer dominates the $M_e$ matrix.
\item Pareto-diverse cycle pool reduces overfitting to any one
  balance metric --- protects against the scorer-bias failure mode
  observed in the 2026-05-10 ABB-vs-per-vertex smoke (signal-only
  global K=10\,000 returned 10\,000 score-1.0 cycles; zero variance
  in $M_e$).
\item Maps to multi-objective optimisation literature; the union
  step is an approximation of the non-dominated front.
\item Composes cleanly with v3: each tier's strategy still fires;
  v3+ wraps it in a multi-scorer outer loop.
\end{itemize}

\textbf{Caveat:} per-scorer enumeration cost is multiplicative
($|S_T| \times$ baseline).  Wall budget per tier:
\begin{itemize}
\item P5: $|S_5| = 1$ (no extra cost; A* already memory-bounded)
\item P4: $|S_4| = 2$ ($\sim 2\times$ DFS work, $\sim 2\times$ wall)
\item P3: $|S_3| = 3$ ($\sim 3\times$ DFS work, but DFS-ABB at
  $m{=}64$ is cheap; expected $\le 30\%$ tier-3 wall increase)
\end{itemize}
Total expected wall vs single-scorer v3: $+10$--$20\%$ (mostly P3).

\subsection*{v4 (MSG/SSG axiom layer, separate plan)}
MSG preprocessor restricts $(V, E)$ to the balance-feasible
substructure.  SSG enumerates minimally balanced cycles within MSG.
Both can drop in as a substrate for v3's per-tier strategies ---
i.e., v3 strategies enumerate \emph{within MSG}, not against
$0..n\_nodes$.

\subsection*{Composed picture (v1 + v2 + v3 + v3+ + v4)}
\begin{enumerate}
\item \textbf{MSG} (v4): restrict $(V, E)$ to balance-feasible
  substructure.
\item \textbf{Vertex prefilter} (v1): inside MSG, drop low-degree
  / no-triangle / etc.\ vertices.
\item \textbf{FPN tier assignment} (v2): bin remaining vertices
  P5..P2 by centrality.
\item \textbf{Per-tier algorithm} (v3): each tier uses the DFS
  variant matched to its $m_v$.
\item \textbf{Per-tier multi-scorer Pareto union} (v3+): within
  each strategy, union top-$K$ across multiple scorers.
\item \textbf{SSG minimality filter} (v4, optional): downstream,
  retain only balance-minimal cycles from the v3+ output.
\end{enumerate}

This is the cognitive-propagation architecture in cycle-generator
form: refined planning (A* + multi-scorer) at the top tiers, bulk
reactive enumeration (DFS-no-prune) at the bottom, with structural
axiom layers (MSG/SSG) bracketing the whole.

\section{Rollback path (revisited)}
\begin{enumerate}
\item Single \texttt{git revert} of the implementation commit removes
  the Rust API addition, the PyO3 binding update, and the Python
  module.
\item Existing callers (yesterday's Slashdot/BitcoinOTC scripts) are
  untouched: they don't set \texttt{HSIKAN\_VERTEX\_FILTER}, so the
  default \texttt{all} branch fires --- bit-equivalent to the current
  enumerator.
\end{enumerate}

\end{document}
