\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=2.3cm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}
\title{Plan: HyMeKo-G\"omb under the unrestricted (transductive) protocol on Epinions}
\date{2026-05-14}
\author{}
\begin{document}
\maketitle

\section{Scope and goal}

Bracket the strict-protocol G\"omb-Epinions number ($0.9526 \pm
0.0018$) from above by measuring G\"omb under the canonical
transductive convention (test-edge signs participate in cycle
$\sigma$-products). The strict number stays the headline; the new
number sits in the supplement and quantifies the leakage path's
contribution on Epinions specifically.

\section{Hypothesis}

Predicted lift over strict: $+0.01$ to $+0.05$ AUC. Lower bound from
``Epinions structural prior already saturated'' (small lift);
upper bound from the Bitcoin-Alpha calibration (HSiKAN-Optuna jumps
$\sim +0.05$ from strict-territory to transductive 0.9970).

\section{Affected files}

\begin{itemize}
\item \texttt{signedkan\_wip/src/run\_gomb\_smoke.py} --- add
  \texttt{--unrestricted-cycles} flag, swap \texttt{(e\_tr, s\_tr)}
  $\to$ \texttt{(g.edges, g.signs)} in the cycle-enumeration block
  when the flag is on. Cycle pool is constructed BEFORE the
  train/val split, so the test-edge signs leak into the
  $\sigma$-product features. Training labels and val/test metric
  computation are unchanged.
\item \texttt{signedkan\_wip/tests/test\_gomb\_unrestricted\_flag.py}
  --- new unit test. Verifies (i) the flag is plumbed,
  (ii) the cycle pool with the flag on enumerates strictly more
  cycles than with it off when there are test edges in the graph,
  (iii) the audit shuffle still flips the unrestricted run to a
  measurably different AUC (sanity).
\item \texttt{signedkan\_wip/experiments/\\
  run\_gomb\_epinions\_unrestricted\_2026\_05\_14.sh} --- 5-seed
  runner using the same v5\_combined config as the strict Epinions
  SOTA-break run.
\end{itemize}

\section{CORE.YAML items touched}

None. \texttt{signedkan\_wip} is not listed in CORE.YAML; only the
core crates (\texttt{hymeko\_core, hymeko\_query, hymeko\_client,
hymeko\_daemon, parser}) and the spec / RTL files are protected.

\section{Interface changes}

Public: a single new CLI flag
\texttt{--unrestricted-cycles} (default \texttt{False}). The
existing strict protocol remains the default. No behaviour change
for callers that do not set the flag. No internal API surface
changes.

\section{Test strategy}

\begin{itemize}
\item \emph{Unit} (1 file, $\sim$3 tests):
  cycle-pool-size monotonicity, flag presence in the argparse, JSON
  output records the flag value.
\item \emph{Integration smoke}: 1-seed run on Bitcoin Alpha
  ($\sim$2 min) with the flag on. Verifies the pipeline doesn't
  crash, the cycle count grows vs strict, the AUC is non-degenerate.
\item \emph{Performance / measurement}: 5-seed Epinions run, same
  v5\_combined config as the strict SOTA-break run, wall-time
  recorded per seed. Same RSS budget ($\sim$6\,GB).
\end{itemize}

\section{Performance budget}

\begin{itemize}
\item Peak RSS: $\leq 8$\,GB (well under the 16\,GB cap).
\item Total wall time: $\leq 60$\,min for 5 seeds
  (strict baseline was $\sim$33\,min; unrestricted pool will be
  modestly larger because test-edge signs enter cycle enumeration,
  but cycle topology is identical, so wall time should rise
  $\leq 1.5\times$).
\item GPU memory: $\sim$5.5--6\,GB.
\end{itemize}

\section{Rollback path}

The flag defaults to \texttt{False}; any call site that doesn't pass
\texttt{--unrestricted-cycles} continues to run the strict protocol
bit-for-bit. The unit test pins this default. To revert, delete
the flag-handling block and the unit test --- $\leq 30$\,LOC
diff total.

\section{Risk anticipation}

\begin{itemize}
\item \emph{Performance-contract preservation.} The cycle-pool
  enumeration already enforces \texttt{max\_cycles}, top-$k$ caps,
  and ABB pruning. Swapping the edge set from \texttt{(e\_tr, s\_tr)}
  to \texttt{(g.edges, g.signs)} preserves these caps because they
  are properties of the enumeration call, not of the edge set.
  Verified by reading the enumeration callsites at
  \texttt{run\_gomb\_smoke.py:464--515}.
\item \emph{Worst-case input.} Epinions: 131\,828 vertices,
  841\,372 edges. Test-edge inclusion raises edge count by
  $\sim$25\,\% (5/4 of training). Cycle pool grows roughly
  proportionally; ABB cap and \texttt{topk=64} per-vertex keep
  pool size bounded.
\item \emph{Audit sanity.} The label-shuffle audit MUST still
  flip the unrestricted run to a different AUC (expected: AUC
  drops, but NOT to chance --- the leakage path is now open, so
  shuffled signs still corrupt the prediction, just less
  dramatically than under strict). This is part of the test plan.
\item \emph{Empty-plan-dir hygiene.} If the experiment is
  abandoned, the plan dir is deleted before commit.
\end{itemize}

\section{Decision tree}

\begin{itemize}
\item Lift $\leq +0.01$: strict number is the result; the
  unrestricted number is supplementary noise. ``Architecture
  saturates the structural prior; leakage path adds nothing.''
\item Lift $+0.01$ to $+0.03$: moderate leakage contribution;
  goes in the methodology paper as an Epinions-specific
  quantification of the canonical convention's information leak.
\item Lift $\geq +0.04$: G\"omb becomes the transductive
  leaderboard leader and the methodology paper has a stronger
  empirical contribution; the strict 0.9526 stays the headline
  because it is the structurally clean number.
\end{itemize}

\section{Bottom line}

A small, additive change to a non-core experiment script.
Unit test pins behaviour; rollback is trivial. The result either
strengthens the strict-protocol headline (by showing the leakage
path adds little) or strengthens the methodology contribution (by
quantifying the leakage path's lift on Epinions).

\end{document}
