\documentclass[11pt]{article}
\usepackage[a4paper,margin=2cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}

\title{Plan: \textbf{HymeKo-Gömb} (Hungarian for sphere) ---\\
  A Concentric Cascade of FIR Volume + HSiKAN Shell + CPML Core\\
  for Signed-Hypergraph Learning}
\author{}
\date{2026-05-11}

\begin{document}
\maketitle

\section*{Etymology}

\textbf{HymeKo-Gömb}: \emph{gömb} is Hungarian for ``sphere / ball
/ globe.'' This name deliberately mirrors and complements
\textbf{Nagare} (流, the Japanese word for ``flow''): the two
sibling architectures of this codebase carry one Japanese and one
Hungarian name, mediated by the HymeKo brand. Together they evoke
a complete cosmology:
\begin{itemize}
  \item \textbf{Nagare} (流, Japanese, ``flow''): the dataflow
    \emph{runtime} (\texttt{hymeko\_nagare} crate, no-autograd,
    closed-form Clifford backward)
  \item \textbf{Gömb} (Hungarian, ``sphere''): the concentric
    \emph{architecture} (this plan; three nested computational shells)
  \item \textbf{HSMM} (Hajdu \& Csapó 2026): the abstract
    \emph{machine} that hosts both
\end{itemize}

Public-facing English alias: \textbf{HymeKo-Sphere}.

\section{Scope}

In contrast to the orthogonal-factorial design (where HSiKAN, CPML,
Clifford-FIR are independently toggled axes), HymeKo-Gömb is a
\emph{cascade}: all three are mandatory but each operates at a
distinct radius through the computation. The architecture is a
single forward path with explicit hierarchical roles:

\begin{itemize}
  \item \textbf{Outer shell (volume)}: $M$ parallel Clifford-FIR
    filter banks process the same cycle pool with different
    learnable filter coefficients. Cheap, redundant coverage.
    Bio-analogue: V1 oriented-edge detectors --- many parallel
    receptive-field copies at the cortical periphery.
  \item \textbf{Middle shell (single HSiKAN layer)}: one Catmull-Rom
    signed-spline aggregator refines the outer-shell output.
    Adds nonlinearity. Bio-analogue: V4 curvature / shape area
    --- a single thick cortical layer transforming oriented
    primitives into shape representations.
  \item \textbf{Inner core (single CPML pass)}: $L=3$ tier-stratified
    multi-layer aggregation compresses to the final per-vertex
    embedding. Bio-analogue: IT (inferotemporal cortex) ---
    multi-scale integration producing object identity.
\end{itemize}

The architecture is loosely inspired by Feature Pyramid Networks
(FPN, Lin et al.\ 2017), but rotated 90°: where FPN stacks
resolution levels in DEPTH (lateral connections between levels),
HymeKo-Gömb stacks specialisation levels at a SINGLE depth (the
``radius'' is the role-in-cascade, not the spatial scale).

\section{Mathematical structure}

Let $X \in \mathbb{R}^{|V| \times d_0}$ be the per-vertex learnable
embedding. Let $\mathcal{C}$ be the cycle pool with associated
edge-sign vectors. Let $\mathrm{tier\_of} : V \to \{0, \ldots, L-1\}$
be the CPML tier assignment.

\subsection{Outer shell (FIR volume)}

Define $M$ Clifford-FIR filters
$\{(a^{(m)}, b^{(m)}) : m = 1, \ldots, M\}$, each of length $k$.
Per-cycle outer outputs:
\[
  \mathrm{outer}_m[c, j] \;=\; \sum_{i=0}^{k-1}
      \mathrm{proj}_{\sigma_i^{(c)}}\!\big(a_i^{(m)} + i b_i^{(m)}\big)
      \cdot X[v_i^{(c)}][j].
\]
Per-vertex outer features via scatter-mean over the cycle pool:
\[
  H_{\mathrm{outer}}^{(m)}[v][j] \;=\;
  \mathrm{mean}_{c: v \in c} \mathrm{outer}_m[c][j].
\]
Concat over the $M$ banks:
\[
  H_{\mathrm{outer}}[v][\,\cdot\,] \;=\;
  \mathrm{concat}_{m=1}^M H_{\mathrm{outer}}^{(m)}[v][\,\cdot\,]
  \;\in\; \mathbb{R}^{M \cdot d_0}.
\]
Optional softmax $\alpha_M$-routing instead of concat for the
multi-bank case --- recovers the existing $\alpha_k$ routing
primitive that drove Slashdot SOTA.

\subsection{Middle shell (HSiKAN)}

One Catmull-Rom signed-spline aggregation:
\[
  H_{\mathrm{middle}} \;=\;
  \mathrm{HSiKAN}\big(H_{\mathrm{outer}}, \mathcal{C}, \sigma\big).
\]
The spline aggregator is the existing
\texttt{MixedAritySignedKAN} forward path with $L_{\mathrm{HSiKAN}} = 1$
inner layer; sign branches via \texttt{use\_minus\_branch=True} as in
the Slashdot recipe.

\subsection{Inner core (CPML pass)}

Tier-stratified $L=3$ aggregation:
\[
  H_{\mathrm{core}} \;=\;
  \mathrm{CPML}_{L=3}\big(H_{\mathrm{middle}}, \mathcal{C},
                          \mathrm{tier\_of}\big),
\]
producing the final per-vertex embedding. Each CPML tier uses an
MLP (the feasibility-stub) or one of $\{\mathrm{HSiKAN}, \mathrm{Clifford\text{-}FIR}\}$;
default for HymeKo-Gömb is MLP (since the spheres already host the
specialised aggregators).

\subsection{Edge predictor head}

\[
  s(u, v) \;=\;
  \mathrm{MLP}_{\mathrm{head}}\big(\mathrm{concat}(H_{\mathrm{core}}[u], H_{\mathrm{core}}[v])\big).
\]

\subsection{Parameter count}

\begin{itemize}
  \item Outer shell: $M \cdot (2k + d_0)$ params (Clifford coefs
    + bias per bank). For $M=8, k=3, d_0=32$: $8 \cdot 38 = 304$.
  \item Middle shell: existing HSiKAN at $M \cdot d_0 \to d_0$
    width: typically $\sim 10^4$ for $d_0=32, \mathrm{grid}=5$.
  \item Inner core: existing CPML at $d_0 \to d_0$ over $L=3$ tiers,
    $\sim 10^4$ params for MLP-stub tiers.
  \item Head: $2 d_0 \cdot d_{\mathrm{head}} + d_{\mathrm{head}}$,
    $\sim 10^3$.
  \item Plus per-vertex embedding: $|V| \cdot d_0 \cdot 4$ bytes
    = $4.2$ MB for Epinions at $d_0=32$.
\end{itemize}

Total: $\sim 4.3$ MB on Epinions. Comparable to the existing
edge\_cr 5-seed-SOTA recipe.

\section{Comparison to the factorial}

The factorial says \emph{which subset of} $\{A, C, F\}$ matters.
HymeKo-Gömb says \emph{all three matter in this specific order}.
The two designs are complementary:

\begin{itemize}
  \item The factorial gives \emph{attribution} (which axis carries
    signal).
  \item HymeKo-Gömb gives a \emph{single high-performance architecture}
    that integrates all three.
\end{itemize}

If the factorial reveals that all three single-axis effects
$\alpha, \gamma, \varphi > 0$ with low interaction terms, HymeKo-Gömb
should outperform any single-axis cell --- it gets the sum of the
three main effects. If interactions are strong (e.g., $\alpha\gamma\varphi > 0$),
HymeKo-Gömb gets that bonus too.

\textbf{If the factorial shows null effects on a given dataset},
HymeKo-Gömb is the cheap test of the cascade hypothesis: maybe the
THREE-ARRANGEMENT is necessary even when no SINGLE axis is.

\section{Affected files}

\begin{lstlisting}
signedkan_wip/src/
├── hymeko_kyu.py                       NEW: Sphere architecture
│   ├── KyuConfig                        (M, k, L_cpml, sphere widths)
│   ├── OuterFIRShell                    M parallel Clifford-FIRs
│   ├── MiddleHSiKAN                     wrapped Catmull-Rom
│   ├── InnerCPMLCore                    tier-stratified MLP/CPML
│   └── HymeKoKyu                        the full cascade
├── run_kyu_sweep.py                    NEW: M / L_cpml sweep runner
└── tests/test_kyu.py                   NEW: forward + backward shapes,
                                          no-cycles ablation,
                                          inner-tier reachability test

docs/plans/2026-05-11-hymeko-gomb-sphere/
├── plan.tex                            this doc
├── plan.pdf                            built
├── plan.tikz                           sphere-shell diagram
└── plan.mmd                            cascade flow
\end{lstlisting}

\textbf{CORE.YAML items touched:} none. Sphere is an additive new
model class; existing pipelines unmodified.

\section{Test strategy}

\textbf{Unit tests} in \texttt{test\_kyu.py}:
\begin{itemize}
  \item \texttt{outer\_shell\_M\_banks\_independent}: $M$ banks
    produce $M$-way concat with distinct coefficient initialisations.
  \item \texttt{middle\_layer\_nonlinearity}: passing identity
    features through Outer + Middle produces non-identity output
    (HSiKAN spline activates).
  \item \texttt{inner\_core\_tier\_reachability}: outer features
    reach the innermost tier through the CPML inward funnelling
    (reuses the CPML feasibility test pattern).
  \item \texttt{forward\_backward\_shape}: end-to-end on a 30-vertex
    fixture, no NaN gradients.
  \item \texttt{no\_cycles\_ablation}: cycles=$\emptyset$ ⇒ all three
    shells degenerate to learnable-embedding pass-through, output
    is the embedding-floor predictor.
  \item \texttt{parameter\_count\_within\_budget}: at $M=8, k=3,
    d_0=32, |V|=10^5$, total params ≤ 5 MB.
\end{itemize}

\textbf{Real-data smoke}: train on Bitcoin OTC 50 epochs single
seed, verify val AUC $\ge 0.85$ (better than the factorial's
mean-MLP cell baseline).

\textbf{Validation gate}: HymeKo-Gömb is reported as a clean
architectural contribution only if its 5-seed paired AUC on at
least one dataset is $\ge$ both:
\begin{enumerate}
  \item the existing edge\_cr Slashdot SOTA recipe at iso-parameters,
  \item the best-cell of the orthogonal factorial.
\end{enumerate}
If neither, it's an interesting null result documented as ``cascade
arrangement does not exceed best-axis-only design.''

\section{Performance budget}

\begin{itemize}
  \item Forward latency at Epinions scale ($|V| = 131K$,
    $|\mathcal{C}| = 530K$, $M = 8$, $d_0 = 32$): expected
    $\sim 200$ ms (outer shell is cheap by virtue of FIR primitives;
    middle and inner are existing operators).
  \item Peak RSS: $\le 5$ GB. Outer shell's $M$-way concat raises
    per-vertex feature width by a factor of $M$; mitigated by
    setting $M = 4$ if $5 \text{GB}$ is breached.
  \item Per-epoch training time (Epinions, 50 epochs): expected
    $\le 3$ min, comparable to the kitchen-sink and edge\_cr recipes.
\end{itemize}

\section{Risk anticipation}

\begin{itemize}
  \item \textbf{Outer-shell collapse}: if $M$ FIR banks converge to
    similar coefficients, the outer shell adds no diversity over
    a single Clifford-FIR layer. Mitigation: encourage diversity
    via orthogonality regularisation on the bank coefficients, or
    diverse initialisation (varied seeds per bank).
  \item \textbf{Middle-layer redundancy}: if outer + inner together
    already capture everything, HSiKAN middle adds no signal.
    Detected via the no-HSiKAN-middle ablation cell. Mitigation:
    if HSiKAN-middle is null, drop it from the published architecture
    and report ``2-shell suffices.''
  \item \textbf{Inner-core dominance}: if CPML core does all the
    work (because it has the most parameters), the outer and middle
    shells are spectators. Detected via the corresponding
    ablations. Mitigation: published architecture clearly attributes
    work via the ablation.
  \item \textbf{$M$-way concat memory}: $M$ banks $\times$ $d_0$ per
    vertex $\times |V|$ floats = $M \cdot d_0 \cdot 4 |V|$ bytes.
    At $M=8, d_0=32, |V|=131K$ that's $\sim 130$ MB; fine. At
    $M=32$ it's $\sim 540$ MB; still fine. At $M=128$ it's $2$ GB;
    bound $M \le 64$ as a default policy.
\end{itemize}

\section{Rollback path}

HymeKo-Gömb is a single Python module + a unit-test file. Removing it
removes the architecture; no existing pipeline regresses. Drop the
plan dir + the module + the test file = clean rollback.

\section{Sequencing}

\begin{enumerate}
  \item Implement \texttt{OuterFIRShell} (re-uses
    \texttt{ClifFIRTierAggregator} per bank).
  \item Implement \texttt{MiddleHSiKAN} (wraps existing
    \texttt{MixedAritySignedKAN}).
  \item Implement \texttt{InnerCPMLCore} (re-uses existing
    \texttt{CPML}).
  \item Compose: \texttt{HymeKoKyu(outer, middle, inner, head)}.
  \item Unit tests (all 6 above).
  \item Smoke train on Bitcoin OTC seed 0 ($\sim$5 min).
  \item Ablations: drop one shell at a time, measure per-cell
    paired Δ on 5 seeds. This is the cascade-attribution
    analogue of the factorial's main-effects decomposition.
\end{enumerate}

\section{Venue}

If HymeKo-Gömb beats the factorial's best cell by $\ge 1.5$ pp paired
on at least one dataset:

\textbf{Primary}: ICLR 2027 (single clear architectural contribution
with cascade decomposition and biological motivation;
``concentric-sphere signed hypergraph learning'').

\textbf{Secondary}: NeurIPS GLF workshop (if the lift is borderline).

\textbf{If null}: archived as a documented negative result alongside
the factorial; the factorial's best-cell becomes the recommended
architecture and HymeKo-Gömb is footnoted as ``cascade does not
exceed orthogonal-best.''

\section{Connection to the cosmology}

The three architectural plans of 2026-05-11 form a coherent set:

\begin{itemize}
  \item \texttt{2026-05-11-hsikan-cpml-fir-orthogonal/}: the
    \textbf{factorial} --- a tool for architectural attribution
    via orthogonal decomposition.
  \item \texttt{2026-05-11-hymeko-gomb-sphere/}: this plan --- the
    \textbf{cascade} architecture that consumes the factorial's
    components and arranges them concentrically.
  \item \texttt{2026-05-11-hymeko-nagare-flow/}: \textbf{Nagare} ---
    the runtime / framework that hosts both, with closed-form
    Clifford backward and no autograd graph.
\end{itemize}

Plus the critical-frozen \texttt{CRITICAL-nagare-to-hsmm-compiler/}
that bridges Nagare to the HSMM substrate when the FPGA prerequisite
lands.

\textbf{Together: factorial \emph{measures}, sphere \emph{combines},
flow \emph{executes}, HSMM \emph{hosts}.} Four plan dirs, one
research program.

\end{document}
