\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=2.3cm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}
\title{Plan: GömbSoma-Ricci-Stim Cluttered MNIST falsification battery}
\date{2026-05-15}
\author{}
\begin{document}
\maketitle

\section{Scope}

Train \texttt{RicciStimDetector} on Cluttered MNIST and compare
against the published HyMeYOLO baselines. The target, from the
Ricci-Stim plan §10:

\begin{itemize}
\item HyMeYOLO \texttt{+kcycle} (broken): 0.204 mAP50 --- beat strongly.
\item HyMeYOLO \texttt{+ricci+kcycle}: 0.235 mAP50 --- beat strongly.
\item HyMeYOLO \texttt{+ricci-mod} (positive control): 0.723 mAP50 --- match or beat.
\item Phase 3-V rolling-shutter G\"ombSoma: beat.
\end{itemize}

\section{Ablation matrix}

Phase 10 exposes the full grid:

\begin{center}
\begin{tabular}{ccccp{6cm}}
\toprule
config & $\alpha$ & $\beta$ & SDRF & what it tests \\
\midrule
A & 0   & 0   & off & bare backbone (Phase 7 baseline) \\
B & 0.1 & 0   & off & Hodge smoothing alone \\
C & 0   & 0.1 & off & Ricci correction alone \\
D & 0.1 & 0.1 & off & full Bochner without graph rewiring \\
E & 0.1 & 0.1 & on  & headline config \\
\bottomrule
\end{tabular}
\end{center}

\section{Affected files}

\begin{itemize}
\item \texttt{signedkan\_wip/src/hymeko\_gomb/soma/vision/ricci\_stim\_train.py}
  --- training infrastructure: anchor assignment, loss, train loop, mAP50 evaluator.
\item \texttt{signedkan\_wip/experiments/run\_ricci\_stim\_cluttered\_mnist\_2026\_05\_15.sh}
  --- orchestration script.
\item \texttt{signedkan\_wip/tests/test\_gomb\_soma\_vision\_ricci\_stim\_train.py}
  --- unit tests for the assignment + loss + smoke train.
\end{itemize}

\section{Performance budget}

\begin{itemize}
\item Per-config wall: $\leq 30$ min on the RTX 2070 SUPER (5000
  images $\times$ 20 epochs $\times$ 5 seeds is the HyMeYOLO baseline budget).
\item Total: 5 configs $\times$ 5 seeds $\approx$ 2.5--3 hours.
\item Peak RSS: $\leq 14$ GB cgroups MemoryMax 16 GB.
\end{itemize}

\section{Test strategy}

\begin{itemize}
\item Unit: anchor-target IoU assignment correctness on small synthetic.
\item Unit: loss computation on hand-built positives + negatives.
\item Integration smoke: 50 images $\times$ 2 epochs $\times$ 1 seed
  ($\leq$ 1 min wall) --- pipeline runs, training loss decreases.
\item Full battery: 5 configs $\times$ 5 seeds (queued; multi-hour).
\end{itemize}

\section{CORE.YAML items touched}

None.

\section{Decision tree}

\begin{itemize}
\item Config E reaches mAP50 $\geq 0.72$ $\Rightarrow$ paper-grade
  YOLO-class result; write up as headline contribution.
\item Config E reaches mAP50 $\in [0.5, 0.72)$ $\Rightarrow$
  partial confirmation; ablate to isolate which component contributes.
\item Config E reaches mAP50 $\in [0.235, 0.5)$ $\Rightarrow$ beats
  the broken HyMeYOLO baseline; partial win; document.
\item Config E reaches mAP50 $< 0.235$ $\Rightarrow$ stimulus-driven
  hypothesis falsified; document as negative result.
\end{itemize}

\section{Risk anticipation}

\begin{itemize}
\item \emph{Anchor-target assignment is non-trivial.} Many detection
  papers spend pages on this. Use simple max-IoU assignment with
  positive threshold 0.3, negative threshold 0.1.
\item \emph{NMS implementation.} Standard greedy NMS at IoU 0.5;
  no learned variants.
\item \emph{Multi-scale anchor scales.} Cluttered MNIST has 28-px
  digits on a 64-px canvas. Use patch\_size\_initial = 32,
  max\_depth = 3 to give anchors from 4 to 32 px.
\item \emph{Per-image variable anchor count.} Loss is computed
  per-image then averaged across batch. Standard pattern.
\item \emph{Empty-plan-dir hygiene.} If shelved, delete the plan dir.
\end{itemize}

\section{Bottom line}

A real experiment on a real dataset against published baselines.
Architectural infrastructure is complete (phases 1--10); this is
the empirical validation. Either we beat YOLO at its own game,
or we have a clean negative result with a known architecture and
a published baseline reference.

\end{document}
