\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{enumitem}
\setlist{nosep,leftmargin=*}
\lstset{basicstyle=\ttfamily\small,frame=single,breaklines=true}
\title{Plan: 3-Dataset HSIKAN Validation\\
       (Slashdot + Bitcoin\{Alpha,OTC\} + Epinions)\\
       v1 vertex prefilter $\times$ v2 FPN tiered $\times$ SoA}
\author{HyMeKo / HSiKAN}
\date{2026-05-12}
\begin{document}
\maketitle

\section{Scope and goal}
Validate the v1 vertex prefilter + v2 FPN tiered caps + SoA cycle
enumerator refactor (all landed 2026-05-10..11) against the three
datasets where HSiKAN currently holds SOTA (Slashdot, Bitcoin Alpha,
Bitcoin OTC) under the transductive protocol, and probe whether the
v2 tiered ladder closes any of the documented $-0.094$ Epinions AUC
gap to SGT.

\textbf{Goals}:
\begin{enumerate}
\item \textbf{AUC parity confirmation}: v1 (vertex prefilter), v2
  (FPN tiered), and v1+v2 combined preserve the 5-seed AUC of the
  baseline configuration within $\pm 0.005$ at the production config
  on each dataset.  The SoA refactor is bit-identical-output by
  construction (unit-tested 2026-05-11), so parity here is about
  the cycle-pool shifts induced by v1/v2, not the SoA path.
\item \textbf{Wall-time savings}: measure the cumulative wall win
  from SoA + dedup-stack-key + v1/v2 vs the legacy
  \texttt{HSIKAN\_TOPK\_MODE=per\_vertex} path.  Expected
  $\sim 21\%$ on Slashdot c4 (already measured); confirm at full
  5-seed scale.  Expected on Epinions: $\sim 30$--$40\%$ enum
  savings (extrapolated from 22\% Slashdot triangle filter at higher
  graph sparsity).
\item \textbf{Epinions gap probe}: test if v2 FPN tiered + Ricci-style
  shape-prior insights from the HyMeYOLO work lifts Epinions AUC
  above the $0.8464$ baseline.  This is a stretch goal --- the
  ceiling-claim memory
  (\texttt{project\_epinions\_ceiling\_2026\_05\_09}) says the gap
  needs sparse-attention or learnable-incidence; v2 by itself may
  not close it but the data is informative.
\end{enumerate}

\section{Configurations}

Four configurations per dataset, exact env-var recipes:

\begin{lstlisting}
# B (baseline) — reproduces yesterday's 5-seed SOTA exactly
HSIKAN_TOPK_MODE=per_vertex
HSIKAN_TOPK_K=128
HSIKAN_TOPK_PRUNER=balance
HSIKAN_TOPK_SCORER=fraction_negative
HSIKAN_TRITON_KERNEL=1
HSIKAN_TRITON_BACKWARD=1
# (no HSIKAN_VERTEX_FILTER, no HSIKAN_TOPK_TIERS)

# V1 = baseline + vertex prefilter (compose: degree + triangle)
HSIKAN_VERTEX_FILTER=compose:degree,triangle
HSIKAN_VERTEX_FILTER_MIN_DEGREE=2

# V2 = baseline + FPN-style concentric per-vertex caps
HSIKAN_TOPK_MODE=per_vertex_tiered
HSIKAN_TOPK_TIERS=0.1:1024,1.0:256,5.0:64,20.0:16,100.0:0

# V1+V2 = both stacked
HSIKAN_TOPK_MODE=per_vertex_tiered
HSIKAN_TOPK_TIERS=0.1:1024,1.0:256,5.0:64,20.0:16,100.0:0
HSIKAN_VERTEX_FILTER=compose:degree,triangle
HSIKAN_VERTEX_FILTER_MIN_DEGREE=2
\end{lstlisting}

All four configs use the SoA-batched PyO3 path (landed 2026-05-11);
the legacy non-batched output path is no longer reachable from the
canonical bindings.  AUC is invariant to this; wall is not.

\section{Datasets}

\begin{tabular}{lrrl}
\toprule
Dataset & $|V|$ & $|E|$ & SOTA baseline AUC (5-seed) \\
\midrule
Bitcoin Alpha & 3{,}783 & 24{,}186 & 0.9468 (cycle), 0.9845 (joint) \\
Bitcoin OTC   & 5{,}881 & 35{,}592 & 0.9266 (cycle), 0.9801 (joint) \\
Slashdot      & 82{,}140 & 549{,}202 & 0.9070 $\pm$ 0.0029 (edge\_cr, kernel on) \\
Epinions      & 131{,}828 & 841{,}372 & 0.8464 $\pm$ 0.0106 (edge\_cr) \\
\bottomrule
\end{tabular}

\section{Affected files}
\begin{itemize}
\item \texttt{signedkan\_wip/experiments/run\_3dataset\_v1\_v2\_phase\_\{A,B,C,D\}.sh}
  --- NEW orchestrator scripts, one per phase below.
\item \texttt{signedkan\_wip/experiments/results/3dataset\_v1\_v2\_\{phase\_A,phase\_B,phase\_C,phase\_D\}.jsonl}
  --- per-phase result lines (one per (config, dataset, seed) triple).
\item \texttt{reports/2026-05-13-3dataset-v1-v2.md} --- final report
  with per-dataset paired tables, wall ratios, AUC deltas, and
  promotion verdicts.
\item No source changes required --- v1/v2/SoA are already shipped
  end-to-end as of 2026-05-11 commits.
\end{itemize}

\section{CORE.YAML items touched}
\textbf{None.}  All affected files are in
\texttt{signedkan\_wip/experiments/} or \texttt{reports/} (allowlist
per CORE.YAML).

\section{Test strategy (CLAUDE.md \S3)}
Acceptance per phase is gated on paired AUC + wall measurements,
not on unit tests (the underlying code is already tested:
117 Rust + 12 cycle\_cache + cluttered\_mnist + circles\_ricci =
$>$ 150 tests passing as of 2026-05-11).

Within each script:
\begin{itemize}
\item Seed-0 smoke fires before the multi-seed loop; if seed-0
  fails to converge (loss does not decrease over the first 10
  epochs), abort the script and dump the log path.
\item Per-seed JSON line written immediately on completion so a
  partial crash doesn't lose results.
\item Final aggregate (mean $\pm$ stdev) computed and emitted at
  script end; comparison to recorded baselines automatic.
\end{itemize}

\section{Ladder}

\subsection*{Phase A --- 1-seed smoke across all 4 datasets}
\textbf{Goal}: confirm all 4 configs train end-to-end on all 4
datasets at production config; surface any config $\times$ dataset
combinations that drop $> 0.005$ AUC vs baseline single-seed.

\begin{tabular}{lll}
\toprule
Dataset & Per-config wall & 4 configs total \\
\midrule
Bitcoin Alpha & $\sim$1 min & $\sim$4 min \\
Bitcoin OTC   & $\sim$2 min & $\sim$8 min \\
Slashdot      & $\sim$13 min & $\sim$52 min \\
Epinions      & $\sim$115 min & $\sim$460 min ($\sim$7.7 h) \\
\midrule
\textbf{Total} & & $\sim$\textbf{9 h} \\
\bottomrule
\end{tabular}

\textbf{Acceptance gate}: every (config, dataset) pair completes
without OOM; AUC delta within $\pm 0.01$ of baseline single-seed.
If a single config OOMs or crashes on a dataset, that config is
flagged and excluded from later phases on that dataset.

\subsection*{Phase B --- 5-seed paired on Bitcoin{Alpha,OTC}}
\textbf{Goal}: paired statistical confirmation that V1, V2, V1+V2
preserve AUC at $5\sigma$-style paired tightness on the cheap
datasets where wall is no constraint.

\begin{tabular}{lll}
\toprule
Dataset & 5-seed $\times$ 4 configs & Total \\
\midrule
Bitcoin Alpha & 5 min & 20 min \\
Bitcoin OTC   & 10 min & 40 min \\
\midrule
\textbf{Total} & & $\sim$\textbf{1 h} \\
\bottomrule
\end{tabular}

\textbf{Acceptance gate}: paired delta vs B (baseline) within
$\pm 0.005$ for V1, V2, V1+V2 at 5-seed; published SOTA (0.9845 /
0.9801 joint cycle+walk) preserved.  If V1+V2 fails parity, halt
Phase C until investigated.

\subsection*{Phase C --- 5-seed paired on Slashdot}
\textbf{Goal}: the medium-sized dataset where v1/v2 already showed
wall wins.  Paired AUC parity + wall measurement.

\begin{tabular}{ll}
\toprule
5-seed $\times$ 4 configs & Total wall \\
\midrule
$\sim$11 min / seed average $\times$ 20 seeds & $\sim$3.7 h \\
\bottomrule
\end{tabular}

\textbf{Acceptance gate}:
\begin{itemize}
\item V1+V2 wall $\le 0.80 \times$ baseline wall.
\item V1+V2 AUC paired delta within $\pm 0.002$ of $0.9070$ baseline
  ($\sigma_{\text{baseline}} = 0.0029$, so $\pm 0.002$ is roughly
  half a sigma).
\item If V1+V2 lifts AUC by $\ge 0.003$ paired: \textbf{paper
  headline} candidate.
\item If V1+V2 drops AUC by $> 0.005$: V1 alone may still be the
  right shipping config; document and proceed to Phase D with V1
  only.
\end{itemize}

\subsection*{Phase D --- Epinions 5-seed (2 configs only)}
\textbf{Goal}: the gap-closing experiment.

\textbf{Pre-cache (obsolete)}: 5 seeds $\times$ 2 configs $\times$ 115
min/seed $= \sim 19$ h.  \textbf{Post-cycle\_cache wiring} (2026-05-11):
seed-0 enumerates ($\sim$95 min on Epinions), seeds 1-4 hit the cache
and skip directly to training ($\sim$20 min each).

\textbf{Two-config compromise}: B (baseline, reproduces 0.8464) vs
the best of \{V1+V2, V1\}-from-Phase-A.  10 seeds total.

\begin{tabular}{ll}
\toprule
5-seed $\times$ 2 configs (post-cache) & Total wall \\
\midrule
2 configs $\times$ (95 + 4$\times$20) min & $\sim$\textbf{6.5 h} \\
\bottomrule
\end{tabular}

\textbf{Acceptance gate}:
\begin{itemize}
\item V1+V2 lifts AUC by $\ge 0.010$ paired vs baseline 0.8464
  ($\sigma_{\text{baseline}} = 0.0106$, so $\ge 0.010$ is paired
  $\sigma \approx 2$): partial gap closure, publish.
\item OR V1+V2 wall $\le 0.70 \times$ baseline wall at AUC parity:
  practical engineering win, publish as a methods contribution.
\item Both: ideal outcome, paper headline.
\item Neither: document as null, preserves the
  \texttt{project\_epinions\_ceiling\_2026\_05\_09} ceiling claim
  --- gap requires sparse-attention or learnable-incidence, not
  cycle-pool refinement.
\end{itemize}

\section{Performance budget}

\textbf{Update 2026-05-11 post-cycle\_cache wiring}: with
\texttt{HYMEKO\_CYCLE\_CACHE=1} now wired into
\texttt{run\_final\_cell.py} (cell\_signed\_graph + cell\_scene call
sites), multi-seed runs amortise enumeration over seeds 2--5.  Cache
hit verified end-to-end: seed-0 misses (writes \texttt{.npz}),
seed-1+ hit (loads 720 KB instead of re-enumerating).  For Epinions
this drops 5-seed wall from $5\times 115 = 575$ min to
$115 + 4\times 20 = 195$ min ($\sim$66\% reduction).

\textbf{Budget table (post-cache):}
\begin{itemize}
\item Phase A (1-seed, no cache benefit): $\le 10$ h wall total.
\item Phase B (5-seed Bitcoin, 4 configs): $\sim$0.5 h.
\item Phase C (5-seed Slashdot, 4 configs): $\sim$1.5 h.
\item Phase D (5-seed Epinions, 2 configs): $\sim$\textbf{6.5 h}.
\item Peak GPU memory: $\le 6$ GB (RTX 2070 SUPER consumer-grade).
\item Peak system RSS: $\le 12$ GB (the 2026-05-10 morning OOM
  at 22 GB must not recur; \texttt{ulimit -v 16777216} on every
  script).
\end{itemize}

\textbf{Cumulative total: $\sim$\textbf{17 h} (post-cache)} vs the
pre-cache projection of $\sim$33 h.  Phase D alone drops from 19 h
to 6.5 h.

\textbf{Caveat}: cache invariant is that
\texttt{(graph\_hash, k, cap, scorer, pruner, topk\_mode, ...)} is
identical across seeds.  Switching configs (B $\to$ V1 $\to$ V2 $\to$
V1+V2) intentionally creates a different cache key per config; each
config still does its first-seed enumeration.  This is the desired
behaviour for paired statistical comparison.

\section{Risk anticipation (CLAUDE.md \S2)}

\subsection*{Performance-contract preservation}
The 2026-05-10 Python-side \texttt{\_subsample\_arr\_to\_tuples}
landed; the 2026-05-11 SoA refactor + dedup-stack-key landed.
\textbf{Both contracts must be honoured in the run scripts}: do
not set \texttt{HSIKAN\_TOPK\_K} above $128$ unless explicitly
authorised (Rust-side per-vertex heaps grow $O(K \cdot |V|)$).

\subsection*{Aspirational comments to honour}
\texttt{project\_rust\_topk\_memory\_followup\_2026\_05\_10}: the
Rust-side \texttt{cap\_total} pushdown has NOT been implemented.
Peak Rust RSS during $k{=}5$ enumeration on Epinions can hit
6--10 GB even with the Python-side fix.  Phase A's Epinions runs
are the first end-to-end test of whether this is OK at production
config with v1/v2 on; if Phase A Epinions OOMs, Rust-side cap
pushdown becomes a Phase D prerequisite.

\subsection*{Worst-case input}
Epinions production-config c2--c5 + w2,w3 mixed-arity training,
80 epochs, max\_k4=100k, quaternion attention + edge\_cr highway,
Triton kernel on.  V2's tiered caps will shift the cycle pool
significantly on $k{=}3$/$k{=}5$ (lift study probe data) but is
untested against the actual training loop at production scale.

\subsection*{What could go wrong}
\begin{itemize}
\item \textbf{V2 tiered changes cycle quality}: the tier
  thresholds (0.1\%, 1\%, 5\%, 20\%, 100\%) are taken from the
  lift-studies report's predicted-AUC sweet spot but never AUC-
  measured at production.  V2 single-seed Phase A is the
  binary go/no-go.
\item \textbf{V1 vertex prefilter drops too much}: the
  \texttt{compose:degree,triangle} default may be too aggressive
  on Epinions's sparse structure (Epinions has more leaves and
  bridge-like vertices than Slashdot).  Adaptive fallback:
  $\min\_degree=1$ if Phase A flags it.
\item \textbf{The Triton kernel + tiered + filter combination has
  not been tested together at production scale}.  Phase A is the
  first integration test.
\end{itemize}

\section{Empty-plan-dir hygiene}
This plan dir is fully populated.  If abandoned at any halt
condition, the dir gets a \texttt{HALTED.md} explaining the
abort.

\section{Halt conditions}
\begin{itemize}
\item Any Phase-A config OOMs on any dataset $\rightarrow$ halt
  that (config, dataset) pair; document and proceed without it.
\item Bitcoin 5-seed (Phase B) shows V1+V2 dropping $> 0.01$
  paired AUC $\rightarrow$ halt before Phase C; design fix
  required.
\item Slashdot 5-seed (Phase C) shows V1+V2 dropping $> 0.005$
  paired AUC $\rightarrow$ pick the best non-V2 config (V1 alone
  or B) for Phase D; do not run V2 on Epinions.
\item Epinions Phase D individual seed wall projection $> 4$ h
  $\rightarrow$ cancel, fall back to 1-seed each at Epinions.
\end{itemize}

\section{Rollback path}
No code changes ship with this plan; only experiment scripts and
the final report.  Revert is \texttt{rm -rf
docs/plans/2026-05-12-3dataset-hsikan-v1-v2/}
+ corresponding experiment-script removals.

\section{Implementation order}
\begin{enumerate}
\item Write Phase A orchestrator script and one-shot test the
  smallest dataset (Bitcoin Alpha, 1-seed) to confirm the recipe
  works.  $\sim 5$ min.
\item Phase A full sweep ($\sim$9 h, overnight).
\item Decision gate: review Phase A AUC deltas, identify any
  config $\times$ dataset failures, lock the configs for Phase B--D.
\item Phase B (1 h).
\item Phase C (4 h).
\item Phase D (19 h, requires explicit user authorisation for
  the long Epinions run --- not auto-launched).
\item Final report at \texttt{reports/2026-05-13-3dataset-v1-v2.md}.
\end{enumerate}

All phases gated on user authorisation for the longer runs (CLAUDE.md
\S11 halt: ``unverified experiment that mutates persistent state'').

\end{document}
