\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=2.3cm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{array}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}
\newcommand{\sig}{\ensuremath{\sigma}}
\newcommand{\Gomb}{G\"omb}
\newcommand{\Soma}{G\"ombSoma}
\newcommand{\RicciStim}{\Soma\textsc{-RicciStim}}
\title{Plan: \RicciStim{} \\
       \large{Adaptive-quadtree stimulus-driven hypergraph network\\
              with Hodge--Bochner--Ricci geometric propagation}}
\date{2026-05-14}
\author{}
\begin{document}
\maketitle

\section{Framing}

The earlier \Soma-Stim plan (\texttt{docs/plans/2026-05-14-gomb-soma-stim/})
proposed swapping YOLO's fixed-grid anchor tiling for a content-driven
top-$K$ salience selector. That plan is preserved for design history.
\textbf{This plan supersedes it} by going further on two axes simultaneously:

\begin{enumerate}
\item \emph{Geometric propagation} --- replace ad-hoc MLP-salience and
  flat message passing with a principled differential-geometric stack:
  Forman-Ricci curvature, Hodge Laplacians at each simplex dimension,
  and Bochner--Weitzenböck-coupled message passing. This is the
  mathematical home of "signed-hypergraph vision."
\item \emph{Adaptive multi-scale anchors} --- replace uniform-grid
  patch tiling with \textbf{curvature-driven quadtree subdivision}.
  Patches with high $|\kappa|$ split recursively; uniform-feature
  regions stay coarse. This gives multi-resolution without YOLO's
  FPN-style separate pyramid pathway.
\end{enumerate}

The hypothesis: these two changes together break the rolling-shutter /
uniform-grid bottleneck that YOLO-class detectors hit on cluttered
small-object scenes.

\section{Background: the differential-geometric stack}

\subsection{Forman-Ricci curvature}

For an edge $e = (u, v)$ in a signed hypergraph $H = (V, E, \sig)$,
the Forman-Ricci curvature is
\[
\kappa(e) = 2 - \deg(u) - \deg(v) + 2 |\Delta(u, v)|
\]
where $|\Delta(u, v)|$ is the number of triangles incident on $e$.
Positive $\kappa$ marks dense / clique-like neighbourhoods;
negative $\kappa$ marks bottleneck edges between otherwise-disjoint
regions. Computed once per layer in $O(|E|)$.

\subsection{Hodge Laplacian on signed simplicial complex}

The signed boundary operator $\partial_k$ maps $k$-simplices to
$(k{-}1)$-simplices respecting $\sig$:
\[
\partial_k [v_0, v_1, \ldots, v_k] = \sum_{i = 0}^{k} (-1)^i \,
  \sig(v_0, \ldots, \hat{v_i}, \ldots, v_k) \,
  [v_0, \ldots, \hat{v_i}, \ldots, v_k].
\]
The Hodge Laplacian at dimension $k$ is
\[
\Delta_k = \partial_{k+1}^* \partial_{k+1} + \partial_k \partial_k^*.
\]
$\Delta_0$ is the standard graph Laplacian on vertices. $\Delta_1$
acts on edges (edge-flow). $\Delta_2$ acts on triangles (face-flow).
The kernel of $\Delta_k$ gives the harmonic forms --- the
topologically invariant features at dimension $k$.

\subsection{Bochner--Weitzenböck identity}

Connects the Hodge Laplacian to Ricci curvature:
\[
\Delta_1 \omega = \nabla^* \nabla \omega + \mathrm{Ric}(\omega^\sharp)^\flat.
\]
On a discrete hypergraph, this means message passing on edges decomposes
into a flat-connection diffusion ($\nabla^* \nabla$, ordinary smoothing)
plus a curvature correction. \textbf{We make this architectural}:
each message-passing step has two coupled components, the diffusion
term and the Ricci term.

\subsection{Curvature-driven graph rewiring (SDRF)}

Topping et al.\ (NeurIPS 2022) showed strongly-negative-$\kappa$ edges
cause exponential over-squashing. Their fix --- Stochastic Discrete
Ricci Flow (SDRF) --- iteratively adds shortcut edges around negative-$\kappa$
bottlenecks until $\min \kappa$ is bounded away from $-\infty$.
We adopt SDRF as a one-time preprocessing step inside the encoder.

\section{Adaptive-quadtree anchor selection}

\subsection{Motivation}

YOLO tiles uniform anchors at every position $\times$ scale. Most are
wasted on background. Phase 3-V's PatchGraphBuilder shares this
defect: 49 anchors on a 28$\times$28 MNIST image regardless of
content. \RicciStim{} replaces this with adaptive subdivision:

\subsection{Algorithm}

\begin{enumerate}
\item Initialise: coarse uniform patches at scale 0 (e.g., 4$\times$4
  patches on a 28$\times$28 image $\to$ 49 level-0 anchors).
\item Compute initial vertex features $x_v$ and Forman $\kappa_v$
  on the current anchor set.
\item Subdivide: every anchor $v$ with $|\kappa_v| > \tau$ (or
  feature variance $> \sigma_\text{var}$, or learned subdivision
  score $> \theta$) is split into 4 sub-anchors (quadtree).
\item Recompute features and $\kappa$ on the new anchor set
  (which is the union of stayed and newly-spawned anchors).
\item Repeat steps 2--4 until max depth $D$ is reached OR
  total-anchor budget $K_\text{max}$ is exhausted OR no anchor
  triggers subdivision.
\item Build the multi-scale signed hypergraph on the final anchor set:
  \emph{same-scale edges} (spatial adjacency at each depth) +
  \emph{cross-scale edges} (parent--child quadtree links). Sign
  every edge by feature-inner-product polarity.
\end{enumerate}

The quadtree IS the hypergraph multi-scale structure. Cross-scale
parent-child edges are an extra hyperedge family the Walk / Polygon /
Triangle layers consume natively (they are topology-agnostic).

\subsection{Budget guarantees}

\begin{itemize}
\item Max depth $D = 4$ (28$\times$28 MNIST: from 4-pixel patches
  down to 1-pixel "pixels-as-vertices" extreme).
\item Total anchor budget $K_\text{max} = 256$ for MNIST,
  $K_\text{max} = 1024$ for Cluttered MNIST (5000 imgs at 5-seed).
\item Quadtree depth and total anchor count are deterministic
  given the curvature threshold $\tau$; no Optuna.
\end{itemize}

\section{Full architecture}

\subsection{Pipeline}

\begin{center}
\begin{tabular}{cl}
\toprule
Stage & Module \\
\midrule
1 & \texttt{PatchPatchifier} --- coarse uniform tiles \\
2 & \texttt{InitialEmbed} --- $\mathrm{Linear}(\text{patch\_dim} \to d_\text{hidden})$ \\
3 & \texttt{FormanCurvatureHead} --- per-vertex $\kappa$ \\
4 & \texttt{AdaptiveQuadtree} --- $\kappa$-driven subdivision loop \\
5 & \texttt{StimulusGraphBuilder} --- multi-scale signed hypergraph \\
6 & \texttt{HodgeLaplacian0,1,2} --- per-dimension propagators \\
7 & \texttt{SDRF} --- one-shot curvature-rewiring \\
8 & \texttt{HypergraphConv}-with-Ricci --- Bochner-coupled message passing \\
9 & \texttt{WalkConv} / \texttt{PolygonConv} / \texttt{TriangleConv} --- existing layers, Ricci-weighted \\
10 & Detection head --- per-anchor cls + offset (or per-image cls) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Bochner-coupled message passing}

The base \texttt{HypergraphConv} forward becomes:
\[
\text{msg}(c) = \psi_{\pi(c)} \Bigl(
   \underbrace{\sum_i W_i^{\pi(c)} x_{v_i}}_{\text{flat connection}}
   + \alpha \cdot \underbrace{\bigl(\Delta_k h\bigr)_c}_{\text{Hodge term}}
   + \beta \cdot \underbrace{\sum_i \kappa(e_i) W_i^{\pi(c)} x_{v_i}}_{\text{Ricci correction}}
\Bigr)
\]
where $\alpha, \beta$ are learned mixing coefficients. The three
terms are the Bochner-Weitzenböck decomposition (flat connection +
Hodge Laplacian + Ricci correction). When $\alpha = \beta = 0$ this
reduces to the existing Phase 1-3-G layers.

\section{Affected files}

\begin{itemize}
\item \texttt{src/hymeko\_gomb/soma/vision/forman.py} ---
  \texttt{FormanCurvatureHead} (vertex / edge $\kappa$).
\item \texttt{src/hymeko\_gomb/soma/vision/hodge.py} ---
  signed boundary operators $\partial_k$, Laplacians $\Delta_0, \Delta_1, \Delta_2$.
\item \texttt{src/hymeko\_gomb/soma/vision/quadtree.py} ---
  \texttt{AdaptiveQuadtree} subdivision driver.
\item \texttt{src/hymeko\_gomb/soma/vision/stim\_graph.py} ---
  multi-scale signed hypergraph builder.
\item \texttt{src/hymeko\_gomb/soma/vision/sdrf.py} ---
  one-shot curvature-rewiring.
\item \texttt{src/hymeko\_gomb/soma/hg\_conv\_bochner.py} ---
  optional Bochner-coupled forward as a mixin / wrapper around the
  Phase-1 \texttt{HypergraphConv}.
\item \texttt{src/hymeko\_gomb/soma/vision/ricci\_stim\_classifier.py}
  --- MNIST-shaped end-to-end model.
\item \texttt{src/hymeko\_gomb/soma/vision/ricci\_stim\_detector.py}
  --- Cluttered MNIST per-anchor detector.
\item Five matching test files (one per module).
\item \texttt{experiments/run\_gomb\_soma\_ricci\_stim\_*.sh}
\end{itemize}

\section{CORE.YAML items touched}

None. All modules under \texttt{signedkan\_wip/}.

\section{Phased rollout (one per session, CLAUDE.md \S6.5 \#11)}

\begin{enumerate}
\item \texttt{FormanCurvatureHead} --- $\kappa$ on the patch graph; unit tests including the analytic check $\sum_e \kappa(e)$ matches Gauss--Bonnet on a closed cycle.
\item \texttt{AdaptiveQuadtree} --- $\kappa$-driven recursive subdivision; unit tests including determinism and depth bound.
\item \texttt{HodgeLaplacian0/1/2} --- signed boundary $\partial_k$ and Hodge $\Delta_k$; unit tests including $\partial_{k} \partial_{k+1} = 0$ and the Hodge decomposition theorem on small examples.
\item \texttt{Bochner-coupled \texttt{HypergraphConv}} --- $\alpha, \beta$ mixing of flat / Hodge / Ricci terms; unit tests including: $\alpha = \beta = 0$ reproduces Phase 1-3-G exactly.
\item \texttt{StimulusGraphBuilder} --- multi-scale signed hypergraph from quadtree + cross-scale edges; unit tests.
\item \texttt{SDRF rewiring} --- iterative curvature-flow edge rewiring; unit tests.
\item End-to-end MNIST classifier; smoke train.
\item End-to-end Cluttered MNIST detector + falsification battery (target $\geq 0.5$ mAP50).
\end{enumerate}

Each phase ships a working module under unit-test coverage. No phase combines.

\section{Test strategy}

Standard for each phase plus:

\begin{itemize}
\item \emph{Gauss--Bonnet check} for $\kappa$: on a closed simply-connected region, $\sum_e \kappa(e) = 2\pi \chi$ ($\chi$ = Euler characteristic). Pinned at low tolerance on a small graph.
\item \emph{Hodge decomposition} on a 5-vertex example: any 1-form $\omega$ decomposes uniquely as harmonic + gradient + curl; verified by reconstruction.
\item \emph{Boundary chain rule} $\partial_k \partial_{k+1} = 0$ pinned exactly.
\item \emph{SDRF monotonicity} after rewiring, $\min_e \kappa(e)$ is strictly higher (bottleneck removed).
\item \emph{Phase-1-3-G reproducibility}: with $\alpha = \beta = 0$, the Bochner-coupled \texttt{HypergraphConv} produces bit-identical output to Phase-1's vanilla forward.
\end{itemize}

\section{Falsification battery}

\begin{center}
\begin{tabular}{lll}
\toprule
Reference                                & Score (mAP50)  & \RicciStim{} prediction \\
\midrule
HyMeYOLO \texttt{+kcycle} (broken)       & 0.204          & \textbf{beat strongly} \\
HyMeYOLO \texttt{+ricci+kcycle}          & 0.235          & beat strongly \\
HyMeYOLO \texttt{+ricci-mod} (positive)  & 0.723          & \textbf{match or beat} \\
HyMeYOLO \texttt{boxes+circles}          & 0.715          & match or beat \\
Phase 3-V rolling-shutter \Soma           & TBD            & beat \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Target:} $\geq 0.72$ mAP50 on Cluttered MNIST at comparable parameter count.

\section{Performance budget}

\begin{itemize}
\item Peak RSS $\leq 14$\,GB (cgroups 16\,GB).
\item Per-seed wall on Cluttered MNIST (5000 imgs, 50 epochs):
  $\leq 30$\,min on the RTX 2070 SUPER (HyMeYOLO baseline $\sim$13\,min;
  Ricci-Stim adds Hodge SpMM and quadtree).
\item GPU memory $\leq 7$\,GB.
\item Full falsification battery (5 seeds): $\leq 3$\,h total.
\end{itemize}

\section{Risk anticipation}

\begin{itemize}
\item \emph{Quadtree training instability.} The subdivision decision is
  non-differentiable. Mitigation: use a soft subdivision score (sigmoid
  on a learned function of $\kappa$ and feature norm) with a
  straight-through estimator; hard subdivide at inference.
\item \emph{Hodge Laplacian on large simplicial complexes.}
  $\Delta_2$ on Cluttered MNIST may be expensive. Cap triangle count
  per anchor; rely on sparsity.
\item \emph{Bochner coupling untraced ground.} The three-term forward
  has been studied in the GNN literature (CurvGN, Topping 2022, sheaf
  neural networks) but never with explicit signed simplicial structure.
  Mitigation: ablation in phase 4 ($\alpha = 0$, $\beta = 0$, both,
  neither) to isolate contributions.
\item \emph{HyMeYOLO baseline reproducibility.} Use identical training
  protocol (data split, augmentation, epochs) to the prior HyMeYOLO
  runs --- this is the basis of the paired comparison.
\item \emph{Empty-plan-dir hygiene.} If shelved, plan dir is deleted before commit.
\end{itemize}

\section{Decision tree}

\begin{itemize}
\item $\geq 0.72$ mAP50 $\Rightarrow$ paper-grade YOLO-class result;
  write up as a contribution: \emph{adaptive-quadtree stimulus-driven
  hypergraph network with Hodge-Bochner-Ricci geometric propagation}.
\item $0.5$--$0.72$ $\Rightarrow$ stimulus-driven hypothesis confirmed
  but not at YOLO scale; document and continue.
\item $0.235$--$0.5$ $\Rightarrow$ partial confirmation; ablate (which
  of Hodge / Ricci / quadtree / $\sig$-balance contributes).
\item $\leq 0.235$ $\Rightarrow$ stimulus-driven hypothesis falsified;
  document as negative result.
\end{itemize}

\section{Why this is the right ambition}

\begin{itemize}
\item \emph{Differential geometry is the right framework for vision.}
  CNN convolution is a clumsy discretisation of $\Delta_0$ heat flow
  on a planar grid. Hodge-aware hypergraph networks are the natural
  generalisation: irregular topology, signed structure, arbitrary
  simplex dimensions.

\item \emph{Adaptive multi-scale without FPN.} Quadtree subdivision
  gives multi-resolution detection from a single forward pass over
  one shared hypergraph. No separate pyramid pathway.

\item \emph{Cartwright--Harary balance becomes a region-coherence
  criterion.} The $\sig$-product on triangles tells us
  \emph{which combinations of regions form a coherent object} ---
  exactly what YOLO can't express.

\item \emph{Empirically anchored.} HyMeYOLO already showed Ricci alone
  works on Cluttered MNIST ($0.723$). The hypothesis is that Ricci
  + Hodge + quadtree + $\sig$-balance can match it at fewer parameters
  with better generalisation.
\end{itemize}

\section{Bottom line}

\RicciStim{} combines three substantive architectural moves with
the existing \Soma{} layer stack: (a) adaptive-quadtree anchor
selection driven by Forman-Ricci curvature; (b) full Hodge-Bochner-
Ricci geometric message passing; (c) SDRF curvature-rewiring against
over-squashing. The plan is phased, each phase produces a tested
module, and the falsification battery has concrete prior baselines.
Target: beat or match HyMeYOLO \texttt{+ricci-mod} ($0.723$ mAP50)
on Cluttered MNIST. A confirmed result is a paper-grade YOLO-class
architectural contribution.

\end{document}
