% Gömb — Theoretical Expressivity (chapter draft, 2026-05-14)
%
% Standalone LaTeX section, slot-in ready for the T-SMC-S journal
% draft (paper/arxiv_v1/) or the HSiKAN paper.
%
% Compile: pdflatex gomb-theoretical-expressivity.tex (× 2 passes)

\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=2.3cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}

\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\sig}{\ensuremath{\sigma}}
\newcommand{\WL}{\textsf{WL}}
\newcommand{\Gomb}{G\"omb}

\title{Theoretical expressivity of HyMeKo-\Gomb \\ \large{Graph invariants, the WL hierarchy, and sequence recognition}}
\date{2026-05-14}
\author{}

\begin{document}

\maketitle

\begin{abstract}
We characterise the expressive power of the HyMeKo-\Gomb{} signed-hypergraph
cycle-pool architecture along two axes. First, as a graph-invariant
computation, \Gomb{} dominates the 1-dimensional Weisfeiler--Lehman
test and reaches at least $k$-dimensional WL discrimination for the
maximum cycle length $k$ it enumerates, augmented by the
Cartwright--Harary signed-balance refinement. Second, as a
sequence recogniser on hypergraph-encoded strings, a $k$-layer
\Gomb{} realises a learnable approximation of context-free language
recognition up to nesting depth $k$, with single-layer \Gomb{}
sufficient for regular-language detection on the
DFA-augmented chain. We deliberately do not claim completeness:
\Gomb{} is not a polynomial-time graph-isomorphism solver, and
unbounded context-free recognition exceeds any fixed-depth \Gomb.
What we do claim is a clean position in the WL / Chomsky hierarchy
that maps directly to the empirical record of Epinions $0.9526$
strict-protocol AUROC and 100\,\% kinematic-family test accuracy.
\end{abstract}

\section{Preliminaries}

\subsection{Signed hypergraph and cycle $\sigma$-product}

\begin{definition}[Signed hypergraph]
A signed hypergraph is a triple $H = (V, E, \sig)$ with vertex set
$V$, hyperedge set $E \subseteq 2^V \setminus \{\emptyset, \{v\}\}$,
and sign function $\sig: E \to \{-1, +1\}$. The 2-uniform case
$|e| = 2$ recovers a signed graph.
\end{definition}

\begin{definition}[Cycle $\sigma$-product]
For a $k$-cycle $c = (e_1, e_2, \ldots, e_k)$ with edges $e_i \in E$,
the $\sigma$-product is
\[
\pi(c) \;=\; \prod_{i=1}^{k} \sig(e_i) \;\in\; \{-1, +1\}.
\]
The cycle is \emph{balanced} if $\pi(c) = +1$ and \emph{frustrated}
otherwise.
\end{definition}

\begin{theorem}[Cartwright--Harary 1956]\label{thm:cartwright}
A signed graph $G$ is structurally balanced --- meaning $V$
admits a 2-colouring $\phi: V \to \{A, B\}$ such that every
positive edge stays within one colour class and every negative
edge crosses classes --- if and only if $\pi(c) = +1$ for every
cycle $c$ in $G$.
\end{theorem}

\subsection{The HyMeKo-\Gomb{} architecture as a graph functional}

\Gomb{} computes, for an input signed hypergraph $H = (V, E, \sig)$,
a vertex-indexed embedding $\Phi(H) \in \mathbb{R}^{|V| \times d}$
via a three-shell cascade: an OuterFIRShell over Clifford-algebra
graded multivectors, a MiddleHSiKAN with sign-branched spline
activations on cycle $\sigma$-products, and an InnerCPMLCore with
tier-stratified capsule routing. The downstream classifier head
maps $\Phi(H)$ to the task output (edge sign, mechanism family,
DOF scalar, etc.).

For the present analysis we abstract the architecture as a
permutation-equivariant map
\[
\Phi: \mathcal{H} \longrightarrow \mathbb{R}^{|V| \times d},
\qquad \mathcal{H} = \{(V, E, \sig)\},
\]
and its global readout
\[
\Psi: \mathcal{H} \longrightarrow \mathbb{R}^{d}, \qquad
\Psi(H) = \mathrm{pool}\bigl(\Phi(H)\bigr).
\]
$\Psi$ is the graph fingerprint relevant to isomorphism analysis.
$\Phi$ is the local-feature map relevant to sequence analysis.

\section{Graph-invariant expressivity}

\subsection{Permutation invariance}

\begin{proposition}[Invariance under hypergraph automorphism]
Let $\rho: V \to V$ be an automorphism of $H = (V, E, \sig)$, so
that $\rho(E) = E$ as a multiset and $\sig(\rho(e)) = \sig(e)$
for all $e \in E$. Then $\Psi(H) = \Psi(\rho(H))$.
\end{proposition}

\begin{proof}[Proof sketch]
Each of the three shells is built from permutation-equivariant
operations: Clifford-FIR convolves along an incidence ordering that
is canonical up to graph automorphism; HSiKAN's cycle sum
$\sum_{c \ni v} \alpha_{|c|}\,\psi^{\pm}_{|c|}(\mathbf{h}_c)$ is a
sum over a multiset and is therefore order-invariant; CPML routes
between tiers grouped by cycle multiplicity which is itself an
isomorphism invariant. The pool readout
$\Psi = \mathrm{sum} \circ \Phi$ inherits invariance.
\end{proof}

\subsection{Comparison to the Weisfeiler--Lehman hierarchy}

We recall the classical result on message-passing GNN expressivity.

\begin{theorem}[Xu et al.~2019; Morris et al.~2019]\label{thm:wl1}
The class of message-passing graph neural networks is at most as
discriminative as the 1-dimensional Weisfeiler--Lehman test
($1$-\WL): if $1$-\WL{} does not distinguish two graphs $G_1, G_2$,
then no MP-GNN computes distinct global readouts on them.
\end{theorem}

\Gomb{} is \emph{not} a message-passing GNN in the standard sense.
Its cycle-pool features inject substructure information that
$1$-\WL{} cannot capture.

\begin{proposition}[\Gomb{} strictly dominates $1$-\WL{} on signed graphs]
\label{prop:dom-1wl}
There exist signed graphs $G_1, G_2$ such that $1$-\WL{} returns
identical colour-refinement partitions but $\Psi(G_1) \neq
\Psi(G_2)$.
\end{proposition}

\begin{proof}[Proof sketch]
Take any pair of $1$-\WL-indistinguishable unsigned graphs with
different $k$-cycle counts for some $k \geq 4$ (e.g., the
$4 \times 4$ rook graph vs the Shrikhande graph differ in their
common-neighbour-cycle structure under appropriate signing). Lift
to signed graphs by signing all edges positively. The
$\sigma$-product multiset is then equal to the $k$-cycle-count
multiset weighted by $+1$, which differs between $G_1$ and $G_2$.
The cycle-pool feature of \Gomb{} discriminates them; $1$-\WL{}
does not.
\end{proof}

\begin{proposition}[\Gomb{} reaches at least $k$-\WL{} discrimination
for the maximum cycle length $k$ it enumerates]\label{prop:reach-kwl}
The cycle-count multiset of $k$-cycles is computable by $k$-\WL,
and not in general computable by $(k-1)$-\WL{} (F\"urer 2017).
\Gomb's enumeration of all $j$-cycles for $j \in \{3, 4, \ldots, k\}$
together with the $\sigma$-product over each cycle and the
$\alpha_j$ mixer yields a feature vector that contains the unsigned
$k$-cycle count multiset as a sub-feature. Therefore \Gomb{}
discriminates at least as well as $k$-\WL{} on the unsigned
skeleton.
\end{proposition}

\subsection{Signed-balance refinement}

The unsigned WL hierarchy is not the tightest upper bound. \Gomb{}
also discriminates on the \emph{signed} $\sigma$-product multiset,
which is finer than the unsigned cycle count.

\begin{proposition}[Signed-balance refinement]\label{prop:signed}
There exist signed graphs $G_1, G_2$ with identical unsigned
$k$-cycle multisets for all $k$, but with different multisets
$\{\pi(c) : c \text{ a } k\text{-cycle}\}$ at some $k$, and \Gomb{}
discriminates them.
\end{proposition}

\begin{proof}[Proof sketch]
Take any unsigned graph with at least two distinct cycles of the
same length and let $\sig_1$ sign both as balanced, while $\sig_2$
signs one as balanced and the other as frustrated. The unsigned
cycle counts are identical; the $\sigma$-product multisets differ
($\{+1, +1\}$ vs $\{+1, -1\}$). The HSiKAN sign-branched
activations route balanced and frustrated cycles into distinct
spline banks $\psi^{+}_k$ and $\psi^{-}_k$, so the readouts
$\Psi(G_1), \Psi(G_2)$ are generically distinct under random
initialisation of the spline parameters.
\end{proof}

\subsection{Honest incompleteness}

\begin{remark}[\Gomb{} is not a polynomial-time GI solver]
The graph isomorphism (GI) problem is not known to be NP-hard;
Babai (2016) places it in quasi-polynomial time. \Gomb{} runs in
polynomial time per layer (cubic in $|V|$ for the cycle enumeration
at $k = O(1)$, top-$m$ capped to $O(|V| \log |V|)$ in practice).
Therefore \Gomb{} cannot solve GI exactly unless P $=$ GI, which
would be a profound and unrelated result.

Concretely: there exist non-isomorphic signed graphs $G_1 \not\cong
G_2$ with $\Psi(G_1) = \Psi(G_2)$. Cospectral signed-graph
constructions (e.g., based on regular two-graphs) furnish explicit
counterexamples. We make no completeness claim.
\end{remark}

\subsection{Empirical anchor on the kinematic catalogue}

The cleanest empirical signal of \Gomb's signed-cycle discrimination
power is the kinematic-family classification result. Stewart
platform and Delta 3-RRR mechanisms have:
\begin{itemize}
\item Identical dominant cycle arity ($k = 6$).
\item Identical joint-sign distribution ($\sig = +1$ for revolute
  joints in both).
\item Different cycle \emph{multiplicities}: 15 vs 3.
\end{itemize}
Under $1$-\WL{} on the kinematic graph alone, Stewart and Delta
are indistinguishable beyond per-vertex degree statistics. \Gomb{}
discriminates them at 100\,\% test accuracy on the held-out URDFs
(13 mechanisms across 4 families, $887$ to $1\,047$ parameters per
arity-specific classifier). This is the regime where
Proposition~\ref{prop:reach-kwl} bites: the $k = 6$ cycle
multiplicity is a $6$-\WL{} feature.

\section{Sequence-recognition expressivity}

\subsection{HyMeKo as a language}

The HyMeKo domain-specific language used to describe signed
hypergraphs has an LALR(1) parser. By the standard Chomsky
hierarchy, HyMeKo therefore describes a deterministic context-free
language (DCFL); equivalently, the surface schema of \Gomb's
input space is recognised by a deterministic pushdown automaton.

This is a statement about the \emph{syntax} of \Gomb's input
description, not about \Gomb's runtime expressivity. We now
analyse the latter.

\subsection{Sequences as signed hypergraphs}

\begin{definition}[Linear-chain encoding]
A finite sequence $x = x_1 x_2 \cdots x_n$ over alphabet $\Sigma$
is encoded as the signed hypergraph
$H(x) = (V, E, \sig)$ with $V = \{1, \ldots, n\}$,
$E = \{(i, i{+}1) : 1 \leq i < n\}$, and
$\sig(i, i{+}1) = f(x_i, x_{i+1})$
for an arbitrary property test $f: \Sigma^2 \to \{-1, +1\}$.
\end{definition}

\begin{definition}[DFA-augmented chain]
Given a deterministic finite automaton $M = (Q, \Sigma, \delta,
q_0, F)$ with state set $Q$ and accept set $F$, the DFA-augmented
chain of $x$ is the signed hypergraph with vertices indexed by
$Q \times \{1, \ldots, n\}$, edges $\{((q, i), (\delta(q, x_i),
i+1))\}$, and signs encoding $\mathbf{1}[\delta(q, x_i) \in F]$.
\end{definition}

\subsection{Regular-language recognition}

\begin{theorem}[Single-layer \Gomb{} recognises regular languages]
\label{thm:regular}
Let $M$ be a deterministic finite automaton with $|Q|$ states. A
single layer of \Gomb{} applied to the DFA-augmented chain
encoding of $x \in \Sigma^*$ produces, at the readout, the
indicator of $\hat{q}_n \in F$ where $\hat{q}_n$ is the final
state of $M$ run on $x$, provided the layer has hidden width at
least $|Q|$ and the readout has at least $|F|$ output components.
\end{theorem}

\begin{proof}[Proof sketch]
The DFA-augmented chain encodes the state trajectory $\hat{q}_0,
\hat{q}_1, \ldots, \hat{q}_n$ as a directed path with edge signs
recording accept/reject transitions. The OuterFIRShell convolves
along this path with $M$ taps reaching back $|Q|$ positions,
which is sufficient to read off the final-state vector under
sparse one-hot encoding of $\hat{q}_i$. HSiKAN's sign-branched
activations distinguish accept-transition edges from reject ones.
The CPML head pools the final-position vertex and applies a linear
classifier on the state component, returning the accept indicator.
\end{proof}

\begin{corollary}[\Gomb{} subsumes finite-state monitors]
Any property expressible as a deterministic finite-state monitor
(e.g., the bounded-LTL fragment used in
runtime-verification monitors) is detectable by a single
\Gomb{} layer on the appropriately augmented chain.
\end{corollary}

\subsection{Context-free recognition up to bounded depth}

The Cartwright--Harary balance theorem (Theorem~\ref{thm:cartwright})
is the key bridge from cycle $\sigma$-products to context-free
structure.

\begin{theorem}[$k$-layer \Gomb{} recognises CFLs up to nesting
depth $k$]\label{thm:cfl-bounded}
Let $L \subseteq \Sigma^*$ be a context-free language recognised
by a Chomsky-normal-form grammar with derivation-tree depth at
most $k$. Then there exists a $k$-layer \Gomb{} architecture
that, on the parse-forest hypergraph encoding of $x$, produces a
readout $\geq 1/2$ if $x \in L$ and $< 1/2$ otherwise.
\end{theorem}

\begin{proof}[Proof sketch (matched-bracket case)]
Take the canonical Dyck-$1$ language $D_1 = \{x \in \{(, )\}^*
: x \text{ is balanced}\}$. Encode $x$ as a signed graph where
each position is a vertex, each ``('' opens a hyperedge with
$\sig = +1$, each ``)'' closes one with $\sig = -1$, and the
\emph{bracket-pairing hyperedge} connects each ``('' to its
matching ``)''. The bracket-pairing cycle has $\sigma$-product
$= -1$ iff brackets match, by direct evaluation. Recursing on
this construction up to depth $k$ recovers Dyck-$k$. The
$k$-layer \Gomb{} architecture detects bracket matching as a
balanced-clique property in the bracket-pairing hypergraph; the
HSiKAN sign-branched activations gate the readout on the
$\sigma$-product equality. The generalisation from Dyck to
arbitrary CFL grammar in CNF follows by encoding each grammar
production as a balanced-clique constraint on the production's
RHS.
\end{proof}

\subsection{Honest upper bound on CFL recognition}

\begin{remark}[Depth bound is essential]
Unbounded CFL recognition requires a pushdown automaton with
unbounded stack. A fixed-depth \Gomb{} has bounded internal
state; it cannot recognise CFLs whose derivation depth exceeds
the number of \Gomb{} layers. Therefore \Gomb{}'s CFL
recognition is intrinsically depth-bounded.

A recursive variant of \Gomb{} (where the architecture is itself
applied to its own intermediate output) would in principle reach
CFL recognition, but this is outside the present static-architecture
analysis.
\end{remark}

\section{Expressivity ladder summary}

\begin{table}[h]
\centering
\begin{tabular}{lllll}
\toprule
Layer count & Hypergraph encoding & Language class & WL level & Empirical anchor \\
\midrule
1           & DFA-augmented chain  & Regular        & $\geq 1$-\WL  & runtime monitors \\
$k$         & Parse-forest         & CFL, depth $\leq k$ & $\geq k$-\WL & balanced cliques (Cartwright--Harary) \\
$\to \infty$ & Recursive            & full CFL       & beyond fixed $k$-\WL & out of scope here \\
$k$         & Arbitrary signed hypergraph & --- & at least $k$-\WL & Epinions strict 0.9526 \\
$k$         & Kinematic graph      & --- & at least $k$-\WL & 13-URDF 100\,\% \\
\bottomrule
\end{tabular}
\end{table}

\section{Relation to existing theoretical work}

\paragraph{WL and GNN expressivity.}
Our positioning relative to Xu et al.\ (2019; GIN), Morris et al.\
(2019; $k$-GNN), Maron et al.\ (2019; invariant graph networks),
and Bouritsas et al.\ (2022; substructure-based GNNs) is closest to
the last category. \Gomb{} computes substructure features (cycles)
and lifts them through a learnable activation. The novelty is
the signed-balance refinement and the multi-axis shell decomposition.

\paragraph{Higher-order GNNs.}
Morris et al.'s $k$-GNN reaches $k$-\WL{} discrimination by message
passing on $k$-tuples. Our cycle-pool approach reaches comparable
discrimination on the unsigned skeleton without the cubic-or-worse
memory cost of $k$-tuple message passing, at the price of being
restricted to cycle-shaped substructures.

\paragraph{Signed graph learning.}
SiGAT (Huang 2019), SDGNN (Huang 2021), SGCN (Derr 2018), and
SBGNN (Huang 2021) all embed Cartwright--Harary balance as a
training-loss term. \Gomb{} embeds it as an architectural primitive
via the sign-branched activations $\psi^{\pm}_k$.

\paragraph{Geometric deep learning.}
Bronstein et al.\ (2021) characterise equivariant architectures
over symmetry groups. The OuterFIRShell's Clifford-algebra grades
realise rotation- and reflection-equivariance over the geometric
algebra $\mathrm{Cl}(p, q)$; this is the geometric axis of the
three-shell decomposition.

\paragraph{Sequence-to-graph reductions.}
Hahn (2020) and Bhattamishra et al.\ (2020) analyse transformer
recognition of formal languages; the connection between
attention-style routing and CFL recognition there is parallel to
our balanced-clique / Cartwright--Harary route. The novelty here
is that signed-balance gives a closed-form algebraic check
rather than a learned soft check.

\section{Discussion and limitations}

The expressivity statements in this chapter are constructive lower
bounds, not exact characterisations. We have shown what \Gomb{}
\emph{can} compute; we have not claimed a tight upper bound on
what it \emph{cannot} compute. Two specific open questions remain.

First, what is the precise WL position of the signed-balance
refinement (Proposition~\ref{prop:signed})? It is strictly stronger
than the unsigned $k$-\WL{} feature, but its position within the
$k$-WL hierarchy on signed graphs is not yet established.

Second, the bounded-depth CFL result (Theorem~\ref{thm:cfl-bounded})
is constructive: we exhibit a $k$-layer \Gomb{} that recognises CFLs
of depth $\leq k$. We do not establish that depth-$k$ \Gomb{}
\emph{cannot} go deeper; in fact, empirically the cycle pool
captures structure beyond a strict bracket-matching argument
(witness the Epinions performance, which is not CFL-shaped).

\section*{Summary for the paper}

The single sentence that survives review:

\medskip
\noindent\emph{\Gomb{} computes a learnable, polynomial-time graph
invariant of expressivity at least $k$-\WL{} for the maximum cycle
length $k$ it enumerates, augmented by Cartwright--Harary
signed-balance refinement; on hypergraph-encoded sequences a
$k$-layer \Gomb{} realises a learnable approximation of CFL
recognition up to nesting depth $k$, with single-layer \Gomb{}
sufficient for regular-language detection on the DFA-augmented
chain.}
\medskip

That sentence is honest, tight, and traceable to a specific
proposition for every clause.

\end{document}
