\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[a4paper,margin=2.3cm]{geometry}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\setlength{\parskip}{0.55em}
\setlength{\parindent}{0pt}
\newcommand{\sig}{\ensuremath{\sigma}}
\newcommand{\Gomb}{G\"omb}
\newcommand{\Soma}{G\"ombSoma}
\title{Plan: \Soma{} --- compositional hypergraph hierarchy for the sensorimotor stack}
\date{2026-05-14}
\author{}
\begin{document}
\maketitle

\section{Framing: \Soma{} is additive to \Gomb, not a replacement}

\Gomb{} (Hungarian: \emph{sphere}; three orthogonal shells:
OuterFIRShell, MiddleHSiKAN, InnerCPMLCore) achieves Epinions
$0.9526 \pm 0.0018$ strict protocol and $0.9527 \pm 0.0018$
unrestricted --- the architecture saturates its structural signal
on social-trust graphs. We hypothesise \Gomb{} is the
\emph{cognitive-level} architecture: it operates on abstractions
(vertices are agents, edges are trust relationships) where the
three axes (geometric, learnable nonlinearity, routing) factor
naturally.

\Soma{} (Greek \emph{soma}: body; this plan) is a different
layout of the same primitives, targeted at the \emph{sensorimotor
level}: where the input is raw or near-raw and structure must be
built up compositionally from short open walks $\to$ closed
polygons $\to$ balanced triangles $\to$ derivative-nodelet
abstractions. \Gomb{} stays canonical; \Soma{} ships as a parallel
branch in \texttt{signedkan\_wip/src/hymeko\_gomb/soma/} and is
empirically falsifiable against \Gomb{} on a curated test battery.

The naming reflects a deliberate division of labour: \Gomb{} (the
sphere; abstract / cognitive) sits above \Soma{} (the body;
sensory \emph{and} motor). A mature system would compose the two
--- \Soma{} processes raw sensory input and emits motor commands,
\Gomb{} reasons about agent-level abstractions on top.

\section{Scope and goal}

Build and test the hypothesis that a compositional hierarchical
hypergraph network --- walk-layer $\to$ polygon-layer $\to$
triangle-layer, with Clifford-FIR as the inter-layer transfer
operator --- outperforms the \Gomb{} flat-mixture shell on tasks
where the input has spatial / compositional / sequential
structure not yet abstracted (vision, kinematic graphs, formal
language detection), while matching \Gomb{} on already-abstracted
signed-link benchmarks.

\section{Architecture}

\subsection{Layer types}

\begin{center}
\begin{tabular}{lll}
\toprule
Layer            & Primitive                & Closure / property                          \\
\midrule
Walk-conv        & open walk $w_k$, $k = 2, 3, \ldots$ & none (open)                        \\
Polygon-conv     & closed cycle $c_k$, $k \geq 4$      & one ring-closure                   \\
Triangle-conv    & $c_3$ with Cartwright--Harary gate   & full structural balance check     \\
Abstraction-conv & derivative-nodelet contraction       & balanced-clique super-vertex      \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Inter-layer Clifford-FIR transfer}

A single Clifford-FIR module per inter-layer boundary, parametrised
by the grade transition: walks live in grade-1 (vector-shaped),
polygons in grade-2 (bivector-shaped), triangles in grade-3
(trivector-shaped). The Clifford geometric product implements the
grade-lifting operator that promotes one layer's feature into the
next layer's natural representation.

\subsection{Compositional rule}

Each closure transition is strictly stronger than the previous:
\[
\text{walk} \xrightarrow{\text{+\,ring closure}}
\text{polygon} \xrightarrow{\text{+\,Cartwright--Harary}}
\text{balanced triangle} \xrightarrow{\text{+\,contraction}}
\text{super-vertex}.
\]
Walks subsume nothing closed; polygons subsume open walks
(as cycle without one edge) plus a closure constraint;
balanced triangles subsume polygons plus a balance constraint;
super-vertices subsume balanced cliques plus a contraction.

\section{Affected files}

All under \texttt{signedkan\_wip/}, outside any CORE.YAML lockdown:

\begin{itemize}
\item \texttt{src/hymeko\_gomb/soma/\_\_init\_\_.py} --- package boundary.
\item \texttt{src/hymeko\_gomb/soma/hg\_conv.py} --- \texttt{HypergraphConv} ABC + canonical implementation; the message-passing primitive shared by all layer types.
\item \texttt{src/hymeko\_gomb/soma/walk\_layer.py} --- \texttt{WalkConvLayer} (HG-conv over $\{w_k\}$).
\item \texttt{src/hymeko\_gomb/soma/polygon\_layer.py} --- \texttt{PolygonConvLayer} (HG-conv over $\{c_k\}_{k \geq 4}$).
\item \texttt{src/hymeko\_gomb/soma/triangle\_layer.py} --- \texttt{TriangleConvLayer} (HG-conv over $c_3$ + Cartwright--Harary balance gate).
\item \texttt{src/hymeko\_gomb/soma/inter\_layer\_clifford.py} --- \texttt{InterLayerCliffordFIR} (grade-lifting Clifford-FIR; replaces the \Gomb{} OuterFIRShell role).
\item \texttt{src/hymeko\_gomb/soma/abstractions.py} --- \texttt{DerivativeNodeletContraction} (balanced-clique $\to$ super-vertex contraction).
\item \texttt{src/hymeko\_gomb/soma/hierarchy.py} --- \texttt{GombSomaHierarchy} (top-level cascade).
\item \texttt{src/run\_gomb\_soma\_smoke.py} --- entry point, kwargs-compatible with \texttt{run\_gomb\_smoke.py} so the same orchestrator scripts apply.
\item \texttt{tests/test\_gomb\_soma\_walk\_layer.py}
\item \texttt{tests/test\_gomb\_soma\_polygon\_layer.py}
\item \texttt{tests/test\_gomb\_soma\_triangle\_layer.py}
\item \texttt{tests/test\_gomb\_soma\_hierarchy.py}
\item \texttt{tests/test\_gomb\_soma\_inter\_layer\_clifford.py}
\end{itemize}

No edits to \Gomb{} files. No edits to \texttt{run\_gomb\_smoke.py}.

\section{CORE.YAML items touched}

None. All \Soma{} code lives in \texttt{signedkan\_wip/}.

\section{Interface changes}

Public: a new entry point \texttt{run\_gomb\_soma\_smoke}, sharing all
CLI flag names with \texttt{run\_gomb\_smoke} that make sense
(\texttt{--dataset}, \texttt{--seed}, \texttt{--n-epochs},
\texttt{--device}, \texttt{--edge-split}, \texttt{--lr},
\texttt{--unrestricted-cycles}, \texttt{--shuffle-train-signs}).
Removed: shell-specific flags (\texttt{--M-outer},
\texttt{--d-outer}, \texttt{--n-tiers}, \ldots). Added:
\texttt{--n-walk-layers}, \texttt{--n-polygon-layers},
\texttt{--n-triangle-layers}, \texttt{--max-walk-len},
\texttt{--max-polygon-len}, \texttt{--abstraction-depth}.

No changes to \Gomb's public interface or wire protocol.

\section{Test strategy}

\subsection{Unit (per layer)}

For each of WalkConv / PolygonConv / TriangleConv:
\begin{enumerate}
\item permutation equivariance: \texttt{layer.forward(perm(H)) == perm(layer.forward(H))} up to numerical tolerance.
\item sparse-aware: layer doesn't materialise dense incidence matrices.
\item signed-correct: layer respects $\sig$-product polarity (positive cycles route to one sub-bank, negative to another).
\item parameter count: $< 100$\,k per layer at default hidden width 16.
\item gradient flow: \texttt{loss.backward()} produces non-zero gradients on every parameter (no dead branches).
\end{enumerate}

\subsection{Unit (inter-layer Clifford-FIR)}

\begin{enumerate}
\item grade transition correctness: input grade-1 multivector
$\to$ output grade-2 multivector; verify the Clifford geometric
product structure.
\item rotation equivariance on a synthetic 3D-embedded graph: rotate
the input, check the output rotates accordingly within the grade.
\end{enumerate}

\subsection{Integration (hierarchy)}

Synthetic SBM with known cycle structure: 4-community graph at
$n = 200$ with planted triangles. \Soma{} hierarchy must:
\begin{enumerate}
\item recover community assignment at $\geq 0.85$ ARI in $\leq 60$ epochs.
\item show walk-layer activates on cross-community edges; polygon-layer activates on within-community closures; triangle-layer activates on planted triangles. Verified via per-layer activation logging.
\end{enumerate}

\subsection{Falsification battery}

Four datasets, \Gomb{} vs \Soma{} paired, 5 seeds each:

\begin{center}
\begin{tabular}{lll}
\toprule
Dataset             & \Gomb{} baseline                   & \Soma{} prediction                 \\
\midrule
Bitcoin Alpha       & $0.8972 \pm 0.0079$            & match (cognitive task)        \\
Slashdot            & $0.9017 \pm 0.0008$            & match (cognitive task)        \\
Epinions strict     & $0.9526 \pm 0.0018$            & match (cognitive task)        \\
Kinematic 13-URDF   & $1.000$ family acc, $k=4/k=6$  & match or improve              \\
HymeYOLO +ricci+kcycle & $\sim 0.20 \pm 0.04$ mAP50 & \textbf{improve} (sensory task) \\
DFA-augmented chain (synthetic) & no \Gomb{} baseline    & $> 0.95$ accuracy             \\
\bottomrule
\end{tabular}
\end{center}

\section{Performance budget}

Each \Soma{} layer is a single hypergraph convolution: linear in
$|E| + \sum_k |C_k|$, comparable per-layer cost to one \Gomb{} shell.
With 4 layers (walk, polygon, triangle, abstraction) the total
compute is $\sim 4/3 \times$ \Gomb.
\begin{itemize}
\item Peak RSS: $\leq 12$\,GB (cgroups MemoryMax 16\,GB).
\item Per-seed wall: $\leq 10$\,min on Epinions ($v_1 = 6.5$\,min).
\item GPU memory: $\leq 7$\,GB on the RTX 2070 SUPER.
\item Total falsification battery (4 datasets $\times$ 5 seeds + 1 vision
  $\times$ 5 seeds + 1 synthetic $\times$ 5 seeds): $\leq 5$\,h.
\end{itemize}

\section{Rollback path}

\Soma{} is fully additive. To disable: delete
\texttt{signedkan\_wip/src/hymeko\_gomb/soma/} and
\texttt{run\_gomb\_soma\_smoke.py}. \Gomb{} entry points unchanged.

\section{Risk anticipation}

\begin{itemize}
\item \emph{Performance-contract preservation.} \Gomb{} already enforces \texttt{max\_cycles}, top-$k$ caps, ABB pruning. \Soma{} layers must inherit the same caps; we ship a single \texttt{EnumerationConfig} dataclass shared with \Gomb's enumeration entry points to prevent drift.
\item \emph{Worst-case input.} Epinions (842\,k edges, 132\,k nodes). The triangle layer's Cartwright--Harary balance gate runs an additional $O(|C_3|)$ pass; we cap $|C_3|$ at the same \texttt{topk} per-vertex cap that \Gomb{} uses.
\item \emph{Empirical falsification.} \Soma{} may match \Gomb{} everywhere $\to$ redundant variant. We ship-deprecate \Soma{} if the 4-dataset paired delta is $\leq 0$ everywhere. \Soma{} may underperform \Gomb{} on sensory tasks $\to$ the compositional-hierarchy hypothesis is wrong; the plan dir is preserved as a negative-result anchor.
\item \emph{Empty-plan-dir hygiene.} If we shelve \Soma, the plan dir is deleted before commit; no half-finished implementations left behind.
\end{itemize}

\section{Decision tree}

\begin{itemize}
\item \Soma{} beats \Gomb{} on sensory (vision, kinematic, sequence) but \emph{matches} on signed-link $\Rightarrow$ confirms the sensory / cognitive layered hypothesis. Both stay shipping; documentation positions them as layered.
\item \Soma{} beats \Gomb{} \emph{everywhere} $\Rightarrow$ \Gomb{} is a flat approximation of \Soma; promote \Soma{} to default, \Gomb{} stays as ablation reference.
\item \Soma{} matches \Gomb{} \emph{everywhere} $\Rightarrow$ redundant variant; \Soma{} is documented as a negative result.
\item \Soma{} loses everywhere $\Rightarrow$ compositional-hierarchy hypothesis is wrong; \Soma{} stays as a negative-result archive.
\end{itemize}

\section{Phases (one per session, per CLAUDE.md §6.5 anti-pattern \#11)}

\begin{enumerate}
\item \texttt{HypergraphConv} ABC + unit tests. No layers yet.
\item Walk layer + unit tests + synthetic SBM smoke.
\item Polygon layer + unit tests + synthetic SBM smoke.
\item Triangle layer + Cartwright--Harary gate + unit tests + synthetic SBM smoke.
\item Inter-layer Clifford-FIR transfer + unit tests + grade-transition correctness test.
\item Top-level hierarchy + \texttt{run\_gomb\_soma\_smoke.py} entry point + Bitcoin Alpha 1-seed production-scale smoke.
\item Falsification battery (4 datasets $\times$ 5 seeds + vision + synthetic) + report.
\end{enumerate}

Each phase produces a working module under unit-test coverage. No phase combines.

\section{Bottom line}

\Soma{} is the next architectural experiment, not a replacement. We
keep \Gomb{} as the canonical signed-link benchmark architecture; we
test \Soma{} specifically against the hypothesis that compositional
walk $\to$ polygon $\to$ triangle $\to$ abstraction is the right
sensory-level structure-building stack. The falsification battery
is concrete and the decision tree is pre-registered.

\end{document}
