\documentclass{article}
\usepackage{hyperref}
\begin{document}
\section*{Plan: G\"omb outer-shell throughput (2026-05-12)}

\paragraph{Scope.}
Vectorize \texttt{scatter\_mean} and batch the $M$ parallel Clifford-FIR banks in
\texttt{OuterFIRShell} (\texttt{signedkan\_wip/src/hymeko\_gomb/shells.py}) so forward
matches the prior numerics (bit-exact is not required; floating-point associativity
should match PyTorch defaults for the same operation order where possible).

\paragraph{Affected files.}
\texttt{signedkan\_wip/src/hymeko\_gomb/shells.py},
\texttt{signedkan\_wip/tests/test\_hymeko\_gomb.py}.

\paragraph{CORE.YAML.}
None.

\paragraph{Interface changes.}
None public beyond \texttt{scatter\_mean} accepting batched input
\texttt{(B, M\_c, d)} in addition to \texttt{(M\_c, d)} (documented in rustdoc-style
comment in the module).

\paragraph{Test strategy.}
Unit tests: parity test for \texttt{scatter\_mean} vs a reference loop implementation
(2D and 3D); existing G\"omb tests unchanged expectations.

\paragraph{Performance budget.}
Wall time for \texttt{OuterFIRShell.forward} should decrease on GPU for $M\ge 4$ at
fixed $(N,M\_c,k,d)$; peak RSS unchanged order-of-magnitude (no materialized
$O(M\cdot N\cdot M\_c)$ temporaries beyond the batched path already implied by the
algorithm).

\paragraph{Rollback.}
Revert \texttt{shells.py} to sequential bank loop + corner loop in
\texttt{scatter\_mean}.

\paragraph{Risks.}
Stacking \texttt{nn.Linear} weights each forward adds small CPU overhead; acceptable
versus eliminated Python loops over $M$ and $k$.
\end{document}
