\documentclass[10pt]{article}
\usepackage[a4paper,margin=2cm]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{listings}
\usepackage{xcolor}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  showstringspaces=false,
  columns=fullflexible,
}

\title{Plan: Per-Vertex Top-$m$ ABB Improvements\\
  (Global-Min, Multi-Criteria, Smaller-$m$, Fullness Gating)}
\author{}
\date{2026-05-11}

\begin{document}
\maketitle

\section*{Background}

The 2026-05-10 ``v1'' per-vertex ABB
(\texttt{enumerate\_top\_k\_per\_vertex\_cycles\_signed\_filtered\_bb\_batched\_rs})
uses the \emph{start vertex's} heap min as the upper-bound threshold.
On Bitcoin Alpha and OTC seed-0 benches this gave $4.78\times$ Rust
walltime savings but lost $-1.9$\,pp / $-7.6$\,pp test AUC --- cycles
useful for some other cycle vertex's heap were thrown away when their
score upper bound fell below the start vertex's threshold.

User directive (2026-05-11): pursue four orthogonal improvement axes
labelled 1, 2, 4, 5.

\section*{Scope}

\begin{itemize}
  \item \textbf{Approach 1 -- Global-min ABB.} Threshold is the
    minimum heap-min across all FULL vertex heaps, shared across
    rayon tasks via \texttt{AtomicU64} (encoding \texttt{f64} bits).
    Monotonically non-increasing; AUC-preserving by construction.
  \item \textbf{Approach 2 -- Multi-criteria composite ABB.} Extend
    \texttt{BoundedScorer} to a \texttt{MultiBoundedScorer} that
    combines score upper bound with axiomatic upper bounds (balance,
    Davis-weak, geometric Ricci). Design only this round.
  \item \textbf{Approach 4 -- Smaller $m_v$ sweep.} Config-only:
    $m \in \{32, 64, 128\}$ on Bitcoin to test whether the
    correctness ceiling on small-data is reached at $m{<}128$.
  \item \textbf{Approach 5 -- Adaptive fullness gating.} ABB only
    fires once a configurable fraction of vertex heaps are at
    capacity; defaults to 25\%. Sweep $\in \{0.0, 0.1, 0.25, 0.5\}$.
\end{itemize}

\section*{Affected files}

\begin{itemize}
  \item \texttt{hymeko\_graph/src/topk\_cycles.rs} ---
    \texttt{dfs\_per\_vertex\_bb\_global},
    \texttt{atomic\_min\_f64},
    \texttt{enumerate\_top\_k\_per\_vertex\_cycles\_par\_adaptive\_starting\_bb\_global\_batched},
    plus three unit tests.
  \item \texttt{hymeko\_graph/src/lib.rs} --- export.
  \item \texttt{hymeko\_py/src/cycles.rs} ---
    \texttt{enumerate\_top\_k\_per\_vertex\_cycles\_signed\_filtered\_bb\_global\_batched\_rs}.
  \item \texttt{hymeko\_py/src/lib.rs} --- registration.
  \item \texttt{signedkan\_wip/src/n\_tuples.py} --- dispatch via
    \texttt{HSIKAN\_USE\_PER\_VERTEX\_ABB\_MODE=global} and
    \texttt{HSIKAN\_PER\_VERTEX\_ABB\_FULLNESS\_GATE} env vars.
\end{itemize}

\section*{CORE.YAML items touched}

None. All edits land in the per-vertex enumerator, its PyO3 binding,
and the Python dispatch shim. The signed-incidence math kernels,
HSIKAN module API, and codegen templates remain untouched.

\section*{Interface changes}

New public Rust function:
\begin{lstlisting}
pub fn enumerate_top_k_per_vertex_cycles_par_adaptive_starting_bb_global_batched<P, S>(
    graph: &SignedGraph,
    k_len: usize,
    pruner: &P,
    m_v: &[u32],
    starting_vertices: &[u32],
    scorer: &S,
    fullness_gate: f64,  // [0,1]; 1.0 effectively disables ABB
) -> TopKCyclesBatch
where P: CyclePruner + Sync, S: BoundedScorer + Sync;
\end{lstlisting}

New Python binding wraps it; output schema unchanged (cycles as
\texttt{(N,k)} \texttt{u32}, scores as \texttt{(N,)} \texttt{f64}).

\section*{Test strategy}

Three Rust unit tests, all on toy fixtures with $\le 8$ nodes:
\begin{enumerate}
  \item \texttt{per\_vertex\_bb\_global\_huge\_caps\_matches\_non\_bb}
    --- caps so large that no heap fills; ABB never fires;
    output count equals non-ABB.
  \item \texttt{per\_vertex\_bb\_global\_subset\_of\_non\_bb} ---
    tight caps + dense graph; ABB fires; output is a SUBSET of
    non-ABB (correctness postcondition).
  \item \texttt{per\_vertex\_bb\_global\_with\_full\_gate\_disables\_abb}
    --- \texttt{fullness\_gate=1.0}; ABB never activates; output
    matches non-ABB.
\end{enumerate}

Python smoke test:
parity of \texttt{(cycles, scores)} between the global-min binding
at huge caps and the non-ABB binding.

Integration test on real data:
Bitcoin Alpha + OTC seed-0 single-pass AUC compared against the
ABB-OFF baseline; tolerance $\le |0.005|$ for ``AUC-preserving''
verdict.

\section*{Performance budget}

\begin{itemize}
  \item Memory: bound by per-vertex heaps. With $m_v{=}128$,
    $k_{\max}{=}6$, \texttt{HeapEntry} $\approx 88$\,B,
    $n{=}50{,}000$ vertices: $\le 0.6$\,GB. Within 16\,GB cap.
  \item Wall-time (enumeration only): expect $\ge 1.5\times$ savings
    over non-ABB at gate$=0.25$, $\le 1.5\times$ slowdown vs v1
    start-ABB. Slashdot $k{=}4$ canonical workload measures this.
  \item Inference latency unchanged (no model-architecture edit).
\end{itemize}

\section*{Rollback path}

All env vars default to OFF / start-mode.
\texttt{HSIKAN\_USE\_PER\_VERTEX\_ABB=0} reverts to the non-ABB
filtered enumerator. The new PyO3 binding is additive --- existing
callers reach the same code path as before.

\section*{Protocol note}

Implementation preceded this plan document due to the user's auto
mode and a continuation-from-previous-context. The plan is filed
retroactively to keep the four-artifact record (\texttt{plan.tex},
\texttt{plan.pdf}, \texttt{plan.tikz}, \texttt{plan.mmd}) intact
for this change. The 2026-05-10 sister plan
\texttt{docs/plans/2026-05-10-abb-global-topk/} covers global
(\emph{not} per-vertex) top-$K$ ABB.

\end{document}
